---
title: "Compile supplementary data (Data S1–S5)"
subtitle: "Mapping subnational sanitation access in LMICs"
author: "Project 0002020937"
format:
  html:
    toc: true
    toc-depth: 2
    self-contained: true
execute:
  warning: false
  message: false
---

## Purpose

Assemble the supplementary data files **Data S1–S5** that reproduce the
main-text figures (and the supplementary figures cited in the text) from the
already-produced pipeline outputs. Each file is written as an `.xlsx` workbook
with **two (or more) sheets**:

- one or more **data** sheets, and
- a **`Dictionary`** sheet that describes every variable *and* the modelled
  outcomes.

Nothing here recomputes the model — it only reads existing outputs, selects and
renames the reporting columns, and documents them. Run it from the repository
root (or from this script's folder — the root is detected automatically).

```{r setup}
library(readr)
library(dplyr)
library(tidyr)
library(openxlsx)   # install.packages("openxlsx") if missing

# --- Locate the repository root (folder containing outputs/ and scripts/) -----
find_repo_root <- function() {
  d <- normalizePath(getwd())
  for (i in 1:6) {
    if (dir.exists(file.path(d, "outputs")) && dir.exists(file.path(d, "scripts")))
      return(d)
    d <- dirname(d)
  }
  stop("Could not locate the repo root (a folder containing outputs/ and scripts/).")
}
repo <- find_repo_root()

# --- Input locations ----------------------------------------------------------
PRED_DIR  <- file.path(repo, "outputs/model_performance/v2/predictions")
TRAIN_DIR <- file.path(repo, "data/processed/training_subcomponents")
EST_DIR   <- file.path(repo, "outputs/survey_estimates")
SHAP_DIR  <- file.path(repo, "outputs/descriptive_statistics/v2")

# --- Output location (new folder) ---------------------------------------------
OUT_DIR <- file.path(repo, "outputs/supplementary_data")
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)

cat("Repo root :", repo, "\n")
cat("Writing to:", OUT_DIR, "\n")
```

## Shared dictionary building blocks

```{r dictionary-helpers}
# Outcome definitions (reused in every Dictionary sheet). Wording follows the
# JMP service-ladder definitions used in the manuscript and the project
# deliverables README.
OUTCOME_DEFS <- tribble(
  ~field,             ~description,
  "basic sanitation", paste("Proportion of the population using at least basic",
                            "sanitation services: improved facilities (e.g. flush/pour-flush",
                            "to sewer, septic tank or pit; ventilated improved pit latrine;",
                            "pit latrine with slab; composting toilet) that are NOT shared",
                            "with other households (JMP service-ladder definition)."),
  "open defecation",  paste("Proportion of the population practising open defecation:",
                            "disposal of human faeces in fields, forests, bushes, open",
                            "bodies of water, beaches or other open spaces (JMP",
                            "service-ladder definition).")
)

MODEL_NOTE <- paste(
  "Outcomes are predicted for the year 2024 at the first sub-national",
  "administrative level (admin-1) across up to 130 low- and middle-income",
  "countries using a TabPFN model trained on the most recent DHS and MICS",
  "household surveys (2015–2024). Each prediction is accompanied by a 90%",
  "prediction interval (pi90_lower–pi90_upper)."
)

# Common identifier columns (definitions/sources from the deliverables README).
COMMON_COLS <- tribble(
  ~field,            ~description,                                   ~units,   ~type,      ~source,
  "GID_1",           "GADM admin-1 district code",                   "",       "character","GADM v3.6 (gadm.org)",
  "NAME_0",          "Country name",                                 "",       "character","GADM v3.6 (gadm.org)",
  "NAME_1",          "Admin-1 district name",                        "",       "character","GADM v3.6 (gadm.org)",
  "sdg_region",      "UN SDG regional grouping",                     "",       "character","JMP country classifications (washdata.org)",
  "wb_income_group", "World Bank income group (L/LM/UM/H)",          "",       "character","World Bank country & lending groups",
  "fragile_context", "OECD fragility classification of the context", "",       "character","OECD States of Fragility"
)

# Assemble a tidy Dictionary data frame: dataset note + model note + outcomes +
# per-column definitions. `columns_df` must have columns
# field/description/units/type/source and represents the data-sheet variables.
make_dict <- function(dataset_id, dataset_desc, columns_df, include_outcomes = TRUE) {
  rows <- list(
    tibble(category = "dataset", field = dataset_id, description = dataset_desc,
           units = "", type = "", source = "This study"),
    tibble(category = "note", field = "modelling", description = MODEL_NOTE,
           units = "", type = "", source = "This study")
  )
  if (include_outcomes) {
    rows <- c(rows, list(
      OUTCOME_DEFS |>
        transmute(category = "outcome", field, description,
                  units = "proportion (0–1)", type = "numeric",
                  source = "JMP service ladder (ref 17)")
    ))
  }
  rows <- c(rows, list(mutate(columns_df, category = "column")))
  bind_rows(rows) |>
    select(category, field, description, units, type, source)
}

# Write a workbook: named data sheet(s) + a Dictionary sheet, lightly formatted.
write_supp <- function(file, data_sheets, dict_df) {
  wb <- createWorkbook()
  hdr <- createStyle(textDecoration = "bold", valign = "top")
  wrap <- createStyle(wrapText = TRUE, valign = "top")

  for (nm in names(data_sheets)) {
    d <- data_sheets[[nm]]
    addWorksheet(wb, nm)
    writeData(wb, nm, d, withFilter = TRUE)
    freezePane(wb, nm, firstActiveRow = 2)
    addStyle(wb, nm, hdr, rows = 1, cols = seq_len(ncol(d)), gridExpand = TRUE)
    setColWidths(wb, nm, cols = seq_len(ncol(d)), widths = "auto")
  }

  addWorksheet(wb, "Dictionary")
  writeData(wb, "Dictionary", dict_df, withFilter = TRUE)
  freezePane(wb, "Dictionary", firstActiveRow = 2)
  addStyle(wb, "Dictionary", hdr, rows = 1, cols = seq_len(ncol(dict_df)),
           gridExpand = TRUE)
  addStyle(wb, "Dictionary", wrap, rows = 2:(nrow(dict_df) + 1), cols = 3,
           gridExpand = TRUE)
  setColWidths(wb, "Dictionary",
               cols = 1:6, widths = c(11, 24, 75, 16, 11, 42))

  saveWorkbook(wb, file, overwrite = TRUE)
  cat("Saved:", file, "\n")
}
```

## Data S1 — Subnational survey estimates used for modelling (Fig. 1)

```{r data-s1}
read_train <- function(path, outcome) {
  # Only read the reporting columns — the training files also carry ~250
  # covariate columns that are not needed here.
  keep <- c("country_outcome", "HH7_region_outcome", "analysis_year",
            "outcome_value", "n_households", "n_psu", "n_psu_weight_scaled",
            "worldpop_sum")
  read_csv(path, col_select = all_of(keep), show_col_types = FALSE) |>
    transmute(
      outcome             = outcome,
      country             = country_outcome,
      admin1_region       = HH7_region_outcome,
      survey_year         = analysis_year,
      observed_proportion = outcome_value,
      n_households        = n_households,
      n_psu               = n_psu,
      psu_weight_scaled   = n_psu_weight_scaled,
      population_2024     = worldpop_sum
    )
}

s1 <- bind_rows(
  read_train(file.path(TRAIN_DIR, "basic_sanitation_training_with_covariates_v2.csv"),
             "basic sanitation"),
  read_train(file.path(TRAIN_DIR, "open_defecation_training_with_covariates_v2.csv"),
             "open defecation")
)

s1_dict <- tribble(
  ~field,               ~description,                                                  ~units,           ~type,      ~source,
  "outcome",            "Sanitation outcome the estimate refers to",                   "",               "character","This study",
  "country",            "Country name (harmonised)",                                   "",               "character","DHS/MICS",
  "admin1_region",      "Admin-1 region name as recorded in the survey",               "",               "character","DHS/MICS",
  "survey_year",        "Year assigned to the survey for analysis",                    "year",           "integer",  "DHS/MICS",
  "observed_proportion","Survey-weighted observed proportion for the outcome",         "proportion (0–1)","numeric", "DHS/MICS",
  "n_households",       "Number of households contributing to the estimate",           "count",          "integer",  "DHS/MICS",
  "n_psu",              "Number of primary sampling units (survey clusters)",          "count",          "integer",  "DHS/MICS",
  "psu_weight_scaled",  "Scaled PSU sampling weight used in modelling",                "",               "numeric",  "This study",
  "population_2024",    "Total admin-1 population (2024)",                             "count",          "numeric",  "WorldPop R2025A"
)

write_supp(
  file.path(OUT_DIR, "Data_S1.xlsx"),
  data_sheets = list("Survey_estimates" = s1),
  dict_df = make_dict(
    "Data S1",
    "Subnational (admin-1) survey estimates of sanitation outcomes used to train the model, from DHS and MICS surveys (2015–2024).",
    s1_dict)
)
```

## Data S2 — Subnational 2024 predictions (Figs 2–5)

```{r data-s2}
read_pred <- function(path, outcome) {
  # The prediction files also carry the full covariate matrix (~250 columns)
  # and are large (~600 MB). Read only the reporting columns plus the national
  # JMP predictor used to flag 2024 data availability. data.table::fread reads
  # selected columns much faster than readr on files this size; fall back to
  # readr if data.table is not installed in your renv.
  jmp_col <- if (outcome == "basic sanitation") "sanitation_basic" else "open_defecation"
  keep <- c("GID_1", "NAME_0", "NAME_1", "sdg_region", "wb_income_group",
            "fragile_context", "median", "pi90_lower", "pi90_upper",
            "worldpop_sum", jmp_col)
  if (requireNamespace("data.table", quietly = TRUE)) {
    df <- data.table::fread(path, select = keep, showProgress = FALSE)
    df <- tibble::as_tibble(df)
  } else {
    df <- read_csv(path, col_select = all_of(keep), show_col_types = FALSE)
  }
  df |>
    transmute(
      outcome            = outcome,
      GID_1, NAME_0, NAME_1, sdg_region, wb_income_group, fragile_context,
      median, pi90_lower, pi90_upper,
      pi90_width         = pi90_upper - pi90_lower,
      population_2024    = worldpop_sum,
      jmp_2024_available = !is.na(.data[[jmp_col]])
    )
}

s2 <- bind_rows(
  read_pred(file.path(PRED_DIR, "basic_sanitation_tabpfn_lmic_predictions.csv"),
            "basic sanitation"),
  read_pred(file.path(PRED_DIR, "open_defecation_tabpfn_lmic_predictions.csv"),
            "open defecation")
)

s2_dict <- bind_rows(
  tibble(field = "outcome",
         description = "Sanitation outcome the prediction refers to",
         units = "", type = "character", source = "This study"),
  COMMON_COLS,
  tribble(
    ~field,              ~description,                                          ~units,            ~type,     ~source,
    "median",            "Predicted proportion for the outcome: median (50th percentile, q50) of the predictive distribution","proportion (0–1)","numeric","This study (TabPFN)",
    "pi90_lower",        "Lower bound of the 90% prediction interval",          "proportion (0–1)", "numeric", "This study (TabPFN)",
    "pi90_upper",        "Upper bound of the 90% prediction interval",          "proportion (0–1)", "numeric", "This study (TabPFN)",
    "pi90_width",        "Width of the 90% prediction interval (upper − lower)", "proportion (0–1)", "numeric", "This study (TabPFN)",
    "population_2024",   "Total admin-1 population (2024)",                      "count",            "numeric", "WorldPop R2025A",
    "jmp_2024_available","Whether a 2024 national JMP estimate was available as a predictor for this country","", "boolean (TRUE/FALSE)","JMP (washdata.org)"
  )
)

write_supp(
  file.path(OUT_DIR, "Data_S2.xlsx"),
  data_sheets = list("Predictions" = s2),
  dict_df = make_dict(
    "Data S2",
    "Subnational (admin-1) predictions of basic sanitation and open defecation for 2024, with 90% prediction intervals, across up to 130 LMICs.",
    s2_dict)
)
```

## Data S3 — Country-level summary of predictions (Fig. S1)

```{r data-s3}
s3 <- s2 |>
  group_by(outcome, NAME_0) |>
  summarise(
    n_admin1             = n(),
    median_prediction    = median(median),
    min_prediction       = min(median),
    max_prediction       = max(median),
    within_country_range = max(median) - min(median),
    median_pi90_width    = median(pi90_width),
    sdg_region           = first(sdg_region),
    wb_income_group      = first(wb_income_group),
    fragile_context      = first(fragile_context),
    .groups = "drop"
  ) |>
  arrange(outcome, median_prediction)

s3_dict <- bind_rows(
  tribble(
    ~field,                ~description,                                                  ~units,            ~type,     ~source,
    "outcome",             "Sanitation outcome the summary refers to",                    "",                "character","This study",
    "NAME_0",              "Country name",                                                "",                "character","GADM v3.6",
    "n_admin1",            "Number of admin-1 units in the country",                      "count",           "integer", "This study",
    "median_prediction",   "Median of admin-1 predicted proportions",                     "proportion (0–1)","numeric", "This study",
    "min_prediction",      "Minimum admin-1 predicted proportion",                        "proportion (0–1)","numeric", "This study",
    "max_prediction",      "Maximum admin-1 predicted proportion",                        "proportion (0–1)","numeric", "This study",
    "within_country_range","Max − min of admin-1 predictions (within-country spread)",    "proportion (0–1)","numeric", "This study",
    "median_pi90_width",   "Median width of the admin-1 90% prediction intervals",        "proportion (0–1)","numeric", "This study"
  ),
  COMMON_COLS |> filter(field %in% c("sdg_region", "wb_income_group", "fragile_context"))
)

write_supp(
  file.path(OUT_DIR, "Data_S3.xlsx"),
  data_sheets = list("Country_summary" = s3),
  dict_df = make_dict(
    "Data S3",
    "Country-level summary of subnational predictions: central tendency, within-country range and median prediction-interval width per outcome.",
    s3_dict)
)
```

## Data S4 — Survey vs prediction comparison (Figs S6, S7)

Two related granularities: region-level agreement (Fig. S6) and the
within-country range comparison (Fig. S7). Kept as two data sheets plus the
shared dictionary.

```{r data-s4}
read_svp <- function(path, outcome) {
  read_csv(path, show_col_types = FALSE) |>
    transmute(
      outcome                 = outcome,
      country,
      admin1_region           = join_key,
      survey_year,
      observed_proportion     = estimate,
      predicted_area_weighted = pred_area,
      predicted_pop_weighted  = pred_pop,
      iou,
      n_gadm_units            = n_gadm
    )
}

s4_region <- bind_rows(
  read_svp(file.path(EST_DIR, "survey_vs_prediction_basic_sanitation.csv"),
           "basic sanitation"),
  read_svp(file.path(EST_DIR, "survey_vs_prediction_open_defecation.csv"),
           "open defecation")
)

s4_range <- read_csv(
  file.path(EST_DIR, "within_country_range_observed_vs_predicted.csv"),
  show_col_types = FALSE) |>
  transmute(
    outcome = recode(outcome,
                     basic_sanitation = "basic sanitation",
                     open_defecation  = "open defecation"),
    country,
    n_admin1           = n,
    observed_range_pp  = obs_range,
    predicted_range_pp = pred_range,
    survey_year,
    iou
  )

s4_dict <- tribble(
  ~field,                    ~description,                                                              ~units,            ~type,      ~source,
  "outcome",                 "Sanitation outcome the comparison refers to",                             "",                "character","This study",
  "country",                 "Country name",                                                            "",                "character","DHS/MICS + GADM",
  "admin1_region",           "Survey region key (country||region) [Region_level sheet]",                "",                "character","DHS/MICS",
  "survey_year",             "Year of the survey being compared with the 2024 prediction",              "year",            "integer",  "DHS/MICS",
  "observed_proportion",     "Survey observed proportion [Region_level sheet]",                          "proportion (0–1)","numeric",  "DHS/MICS",
  "predicted_area_weighted", "2024 prediction, area-weighted over the survey region [Region_level]",     "proportion (0–1)","numeric",  "This study",
  "predicted_pop_weighted",  "2024 prediction, population-weighted over the survey region [Region_level]","proportion (0–1)","numeric",  "This study",
  "iou",                     "Intersection-over-union of survey and GADM boundaries (concordance)",       "0–1",             "numeric",  "This study",
  "n_gadm_units",            "Number of GADM units overlapping the survey region [Region_level]",         "count",           "integer",  "This study",
  "n_admin1",                "Number of admin-1 units in the country [Within_country_range sheet]",       "count",           "integer",  "This study",
  "observed_range_pp",       "Observed within-country range (max − min) [Within_country_range sheet]",    "percentage points","numeric", "DHS/MICS",
  "predicted_range_pp",      "Predicted within-country range (max − min) [Within_country_range sheet]",   "percentage points","numeric", "This study"
)

write_supp(
  file.path(OUT_DIR, "Data_S4.xlsx"),
  data_sheets = list("Region_level"          = s4_region,
                     "Within_country_range"  = s4_range),
  dict_df = make_dict(
    "Data S4",
    "Comparison of survey observations with 2024 predictions over the same areas: region-level agreement (Region_level) and within-country ranges (Within_country_range).",
    s4_dict)
)
```

## Data S5 — Predictor importance (SHAP) (Fig. S8)

```{r data-s5}
read_shap <- function(path, outcome) {
  read_csv(path, show_col_types = FALSE) |>
    mutate(outcome = outcome,
           pct_importance = mean_abs_shap / sum(mean_abs_shap) * 100) |>
    transmute(outcome, feature, label, mean_abs_shap, pct_importance) |>
    arrange(desc(mean_abs_shap))
}

s5 <- bind_rows(
  read_shap(file.path(SHAP_DIR, "basic_sanitation_tabpfn_shap_importance_k45_n250.csv"),
            "basic sanitation"),
  read_shap(file.path(SHAP_DIR, "open_defecation_tabpfn_shap_importance_k90_n250.csv"),
            "open defecation")
)

s5_dict <- tribble(
  ~field,          ~description,                                                       ~units,        ~type,      ~source,
  "outcome",       "Sanitation outcome the importance refers to",                      "",            "character","This study",
  "feature",       "Predictor variable name (as used in the model)",                   "",            "character","This study",
  "label",         "Human-readable predictor label",                                   "",            "character","This study",
  "mean_abs_shap", "Mean absolute SHAP value (mean over admin-1 units)",                "",            "numeric",  "This study (TabPFN + SHAP)",
  "pct_importance","Mean absolute SHAP as a percentage of the outcome's total",         "percent",     "numeric",  "This study"
)

write_supp(
  file.path(OUT_DIR, "Data_S5.xlsx"),
  data_sheets = list("Predictor_importance" = s5),
  dict_df = make_dict(
    "Data S5",
    "Predictor importance for each outcome, measured as the mean absolute SHAP value across admin-1 units.",
    s5_dict, include_outcomes = FALSE)
)
```

## Verification against manuscript values

Recomputes the headline numbers reported in the manuscript from the assembled
data frames and reports `PASS` / `FAIL`. This is a correctness check on the
compilation — if the source outputs change, re-run and confirm all checks pass.

```{r verify}
# --- helpers -----------------------------------------------------------------
fmt <- function(x) {
  if (is.numeric(x)) paste(round(x, 3), collapse = ", ") else paste(x, collapse = ", ")
}
num_check <- function(label, actual, expected, tol) {
  status <- if (length(actual) == 0 || is.na(actual)) "NA"
            else if (abs(actual - expected) <= tol) "PASS" else "FAIL"
  tibble(check = label, actual = fmt(actual),
         expected = paste0(expected, " ± ", tol), status = status)
}
set_check <- function(label, actual_vec, expected_vec) {
  tibble(check = label, actual = fmt(actual_vec), expected = fmt(expected_vec),
         status = if (setequal(actual_vec, expected_vec)) "PASS" else "FAIL")
}
bs_med_by <- function(grp_col, grp_val) {
  v <- s2 |> filter(outcome == "basic sanitation", .data[[grp_col]] == grp_val) |> pull(median)
  if (length(v) == 0) NA_real_ else median(v)
}
bs_n_by <- function(grp_col, grp_val) {
  s2 |> filter(outcome == "basic sanitation", .data[[grp_col]] == grp_val) |> nrow()
}
top_n_countries <- function(oc, col, desc = FALSE, n = 5) {
  d <- s3 |> filter(outcome == oc)
  d <- if (desc) arrange(d, desc(.data[[col]])) else arrange(d, .data[[col]])
  head(d$NAME_0, n)
}

# --- reference tables (for eyeballing / debugging labels) --------------------
cat("SDG region labels:\n"); print(sort(unique(s2$sdg_region)))
cat("\nIncome group labels:\n"); print(sort(unique(s2$wb_income_group)))
cat("\nBasic sanitation median by SDG region:\n")
s2 |> filter(outcome == "basic sanitation") |>
  group_by(sdg_region) |>
  summarise(n = n(), median = round(median(median), 3), .groups = "drop") |>
  arrange(median) |> as.data.frame() |> print()

# --- automated checks --------------------------------------------------------
inc_counts_bs <- s2 |> filter(outcome == "basic sanitation") |>
  count(wb_income_group) |> pull(n) |> sort()

checks <- bind_rows(
  num_check("S2: basic sanitation rows = 2310",
            sum(s2$outcome == "basic sanitation"), 2310, 0),
  num_check("S2: open defecation rows = 2310",
            sum(s2$outcome == "open defecation"), 2310, 0),
  num_check("S2: distinct countries = 130",
            n_distinct(s2$NAME_0), 130, 0),
  set_check("S2: income-group admin-1 counts = {406, 812, 1092}",
            inc_counts_bs, c(406, 812, 1092)),
  num_check("S2: sub-Saharan Africa n (BS) = 662",
            bs_n_by("sdg_region", "Sub-Saharan Africa"), 662, 0),
  num_check("S2: sub-Saharan Africa BS median ~ 0.39",
            bs_med_by("sdg_region", "Sub-Saharan Africa"), 0.39, 0.02),
  num_check("S2: Europe & N. America BS median ~ 0.97",
            bs_med_by("sdg_region", "Europe and Northern America"), 0.97, 0.02),
  num_check("S2: Oceania BS median ~ 0.49",
            bs_med_by("sdg_region", "Oceania"), 0.49, 0.03),
  set_check("S3: lowest-5 BS countries",
            top_n_countries("basic sanitation", "median_prediction"),
            c("Central African Republic", "South Sudan", "Chad",
              "Guinea-Bissau", "Madagascar")),
  set_check("S3: top-5 within-country range (BS)",
            top_n_countries("basic sanitation", "within_country_range", desc = TRUE),
            c("Nigeria", "Yemen", "Tanzania", "Mozambique", "Malawi")),
  # NOTE: the manuscript's "median within-country range 32 p.p. (BS) / 13 p.p.
  # (OD)" correspond to the OBSERVED survey ranges (Data S4), not the predicted
  # ranges. The predicted ranges are markedly smaller because the model
  # compresses local extremes (Fig. S7) — so we verify against the observed
  # ranges here and report the (smaller) predicted ranges alongside.
  num_check("S4: median OBSERVED within-country range BS ~ 32 p.p.",
            median(s4_range$observed_range_pp[s4_range$outcome == "basic sanitation"], na.rm = TRUE), 32, 3),
  num_check("S4: median OBSERVED within-country range OD ~ 13 p.p.",
            median(s4_range$observed_range_pp[s4_range$outcome == "open defecation"], na.rm = TRUE), 13, 3),
  num_check("S1: basic sanitation region-estimates ~ 1168",
            sum(s1$outcome == "basic sanitation"), 1168, 5),
  num_check("S1: open defecation region-estimates ~ 1205",
            sum(s1$outcome == "open defecation"), 1205, 5),
  num_check("S4: median IoU ~ 0.96",
            median(s4_region$iou, na.rm = TRUE), 0.96, 0.03),
  num_check("S5: top BS predictor importance ~ 38%",
            s5 |> filter(outcome == "basic sanitation") |>
              slice_max(mean_abs_shap, n = 1) |> pull(pct_importance),
            38, 3)
)

# Within-country range compression (predicted < observed) — Fig. S7
comp <- s4_range |> group_by(outcome) |>
  summarise(obs = median(observed_range_pp, na.rm = TRUE),
            pred = median(predicted_range_pp, na.rm = TRUE), .groups = "drop") |>
  mutate(status = ifelse(pred < obs, "PASS", "FAIL"))
checks <- bind_rows(checks, tibble(
  check = paste0("S4: predicted range < observed range (", comp$outcome, ")"),
  actual = paste0("pred=", round(comp$pred, 1), " obs=", round(comp$obs, 1)),
  expected = "pred < obs", status = comp$status))

cat("\n\n=== Verification results ===\n")
knitr::kable(checks)

# Context: observed vs predicted within-country ranges (the manuscript's 32/13
# p.p. are the OBSERVED ranges; predicted ranges are smaller — Fig. S7).
cat("\nWithin-country range (median, percentage points):\n")
s3 |> group_by(outcome) |>
  summarise(predicted_all_countries_pp = round(median(within_country_range) * 100, 1),
            .groups = "drop") |>
  left_join(
    s4_range |> group_by(outcome) |>
      summarise(observed_surveyed_pp  = round(median(observed_range_pp, na.rm = TRUE), 1),
                predicted_surveyed_pp = round(median(predicted_range_pp, na.rm = TRUE), 1),
                .groups = "drop"),
    by = "outcome") |>
  as.data.frame() |> print()

n_fail <- sum(checks$status == "FAIL")
if (n_fail == 0) {
  cat("\nAll", nrow(checks), "checks passed.\n")
} else {
  cat("\n", n_fail, "check(s) FAILED — inspect the table above.\n")
}
```

## Validation

```{r validate}
summary_tbl <- tibble::tribble(
  ~file,        ~rows,          ~cols,
  "Data_S1",    nrow(s1),        ncol(s1),
  "Data_S2",    nrow(s2),        ncol(s2),
  "Data_S3",    nrow(s3),        ncol(s3),
  "Data_S4 (region)", nrow(s4_region), ncol(s4_region),
  "Data_S4 (range)",  nrow(s4_range),  ncol(s4_range),
  "Data_S5",    nrow(s5),        ncol(s5)
)
knitr::kable(summary_tbl)

sessionInfo()
```
