---
title: "Covariate Cluster Analysis"
subtitle: "Hierarchical clustering of geospatial covariates for feature selection"
---

# 1. Setup

## Libraries

```{r}
library(tidyverse)
library(here)
library(ClustOfVar)
```

## Paths

```{r}
PATH_PRED_COVARIATES   <- here::here("data/prediction/prediction_covariates_2024.csv")
OUTPUT_DIR_SMDW        <- here::here("outputs/03b_cluster_analysis/smdw")
OUTPUT_DIR_SANITATION  <- here::here("outputs/03b_cluster_analysis/sanitation")

dir.create(OUTPUT_DIR_SMDW,       recursive = TRUE, showWarnings = FALSE)
dir.create(OUTPUT_DIR_SANITATION,  recursive = TRUE, showWarnings = FALSE)
```

------------------------------------------------------------------------

# 2. Input data

Prediction covariates prepared in `05_prepare_prediction_data.qmd`.
Each row is one GADM admin-1 region.

```{r}
df_pred_covariates <- readr::read_csv(PATH_PRED_COVARIATES, show_col_types = FALSE)

cat("Regions:", nrow(df_pred_covariates), "\n")
cat("Countries:", dplyr::n_distinct(df_pred_covariates$NAME_0), "\n")
cat("Columns:", ncol(df_pred_covariates), "\n")
```

------------------------------------------------------------------------

# 3. Variable sets

Two analyses are run in parallel:

- **SMDW**: full variable set including JMP drinking water subcomponents
- **Sanitation**: same set minus the four drinking water variables not
  available in the sanitation training data

The shared EO and country-level variables are defined once and reused.

```{r}
# ── Shared EO and country-level variables ────────────────────────────────────
band_names_shared <- c(
  # ── EO features ────────────────────────────────────────────────────────────
  "CGIAR_Aridity_Index",
  "CGIAR_PET",
  "CHELSA_BIO_Annual_Mean_Temperature",
  "CHELSA_BIO_Annual_Precipitation",
  "CHELSA_BIO_Precipitation_Seasonality",
  "CHELSA_BIO_Precipitation_of_Coldest_Quarter",
  "CHELSA_BIO_Precipitation_of_Driest_Month",
  "CHELSA_BIO_Precipitation_of_Driest_Quarter",
  "CHELSA_BIO_Precipitation_of_Warmest_Quarter",
  "CHELSA_BIO_Precipitation_of_Wettest_Month",
  "CHELSA_BIO_Precipitation_of_Wettest_Quarter",
  "CHELSA_BIO_Temperature_Annual_Range",
  "CHELSA_BIO_Temperature_Seasonality",
  "CIFOR_TropicalPeatlandExtent",
  "CSP_Global_Human_Modification",
  "ConsensusLandCoverClass_Barren",
  "ConsensusLandCoverClass_Cultivated_and_Managed_Vegetation",
  "ConsensusLandCoverClass_Deciduous_Broadleaf_Trees",
  "ConsensusLandCoverClass_Evergreen_Broadleaf_Trees",
  "ConsensusLandCoverClass_Evergreen_Deciduous_Needleleaf_Trees",
  "ConsensusLandCoverClass_Herbaceous_Vegetation",
  "ConsensusLandCoverClass_Mixed_Other_Trees",
  "ConsensusLandCoverClass_Open_Water",
  "ConsensusLandCoverClass_Regularly_Flooded_Vegetation",
  "ConsensusLandCoverClass_Shrubs",
  "ConsensusLandCoverClass_Snow_Ice",
  "ConsensusLandCoverClass_Urban_Builtup",
  "ConsensusLandCover_Human_Development_Percentage",
  "CrowtherLab_Tree_Density",
  "EarthEnvCloudCover_CloudForestPrediction",
  "EarthEnvTexture_CoOfVar_EVI",
  "EarthEnvTexture_Contrast_EVI",
  "EarthEnvTexture_Correlation_EVI",
  "EarthEnvTexture_Dissimilarity_EVI",
  "EarthEnvTexture_Entropy_EVI",
  "EarthEnvTexture_Evenness_EVI",
  "EarthEnvTexture_Homogeneity_EVI",
  "EarthEnvTexture_Maximum_EVI",
  "EarthEnvTexture_Range_EVI",
  "EarthEnvTexture_Shannon_Index",
  "EarthEnvTexture_Simpson_Index",
  "EarthEnvTexture_Std_EVI",
  "EarthEnvTexture_Uniformity_EVI",
  "EarthEnvTexture_Variance_EVI",
  "EarthEnvTopoMed_1stOrderPartialDerivEW",
  "EarthEnvTopoMed_1stOrderPartialDerivNS",
  "EarthEnvTopoMed_Eastness",
  "EarthEnvTopoMed_Elevation",
  "EarthEnvTopoMed_Roughness",
  "EarthEnvTopoMed_Slope",
  "EarthEnvTopoMed_TerrainRuggednessIndex",
  "EarthEnvTopoMed_TopoPositionIndex",
  "EsaCci_BurntAreasProbability",
  "FanEtAl_Depth_to_Water_Table_AnnualMean",
  "FanEtAl_Depth_to_Water_Table_AnnualSD",
  "GHS_Population_Density",
  "GLW3_RuminantsDistribution_downsampled10km",
  "GPWv4_Population_Density",
  "GiriEtAl_MangrovesExtent",
  "MODIS_EVI",
  "MODIS_NDVI",
  "MODIS_NPP",
  "PelletierEtAl_SoilAndSedimentaryDepositThicknesses",
  "SG_Absolute_depth_to_bedrock",
  "SG_Bulk_density_015cm",
  "SG_Depth_to_bedrock",
  "SG_H2O_Capacity_015cm",
  "SG_Saturated_H2O_Content_015cm",
  "TootchiEtAl_WetlandsRegularlyFlooded",
  "WCS_Human_Footprint_2009",
  "chirps_annual_precipitation",
  "era5_temperature_2m",
  "ghsl_built_surface",
  "ghsl_population",
  "ghsl_urban_frac",
  "jrc_building_height",
  "map_friction",
  "modis_evi",
  "modis_ndvi",
  "runoff_max_annualmax",
  "runoff_min_annualmin",
  "temperature_2m_max_annualmax",
  "viirs_average",
  "worldpop",
  "worldpop_sum",
  "ghsl_population_sum",
  "area_km2",
  # ── Country-level features ──────────────────────────────────────────────────
  "gdp_per_capita_constant_2015_usd",
  "secondary_education_duration_years",
  "ww_collection_percent",
  "ww_treatment_percent",
  "ww_reuse_percent",
  "control_of_corruption",
  "governance_effectiveness",
  "political_stability",
  "regulatory_quality",
  "rule_of_law",
  "voice_and_accountability",
  "sanitation_basic",
  "open_defecation"
)

# ── Drinking water JMP variables (SMDW only) ──────────────────────────────────
band_names_dw <- c(
  "drinking_water_accessible_on_premises",
  "drinking_water_available_when_needed",
  "drinking_water_piped",
  "drinking_water_non_piped"
)

band_names_smdw        <- c(band_names_shared, band_names_dw)
band_names_sanitation  <- band_names_shared

cat("SMDW variables:       ", length(band_names_smdw), "\n")
cat("Sanitation variables: ", length(band_names_sanitation), "\n")

# Check all variables present in prediction data
check_vars <- function(vars, label) {
  missing <- setdiff(vars, names(df_pred_covariates))
  if (length(missing) > 0) {
    warning(label, ": variables not found in prediction data:\n",
            paste(" -", missing, collapse = "\n"))
  } else {
    message(label, ": all variables present.")
  }
}

check_vars(band_names_smdw,       "SMDW")
check_vars(band_names_sanitation,  "Sanitation")
```

------------------------------------------------------------------------

# 4. SMDW cluster analysis

k values: 5, 10, 15, 20, 30, 45, 90, and the full set
(`r length(band_names_smdw)`).

## Prepare feature matrix

```{r}
# Extract only the SMDW satellite/covariate columns from the prediction dataset
# and convert everything to numeric (in case any columns were read as text)
samples_smdw <- df_pred_covariates %>%
  dplyr::select(dplyr::all_of(band_names_smdw)) %>%
  dplyr::mutate(dplyr::across(everything(), as.numeric))

# Check for zero-variance columns — variables with only one unique value are
# useless for prediction and can cause models to crash or behave unexpectedly
constant_cols_smdw <- names(samples_smdw)[
  sapply(samples_smdw, function(x) dplyr::n_distinct(x, na.rm = TRUE) <= 1)
]
if (length(constant_cols_smdw) > 0) {
  message("Zero-variance columns (SMDW):\n",
          paste(" -", constant_cols_smdw, collapse = "\n"))
} else {
  message("No zero-variance columns (SMDW).")
}

# Report how many rows have complete data (no missing values across all columns)
# — only complete cases can be passed to the model for prediction
cat("Complete cases:", sum(complete.cases(samples_smdw)), "/",
    nrow(samples_smdw), "\n")
```

## Hierarchical clustering

```{r}
# Set a random seed so the clustering result is reproducible across runs
set.seed(123)

# Run hierarchical clustering on the SMDW variables (columns, not rows).
# This groups similar variables together based on how correlated they are —
# variables that carry nearly the same information end up in the same cluster.
hclust_smdw <- hclustvar(X.quanti = as.data.frame(samples_smdw))

# Plot the aggregation index at each step of the clustering.
# The "elbow" in the curve (where the index drops sharply) suggests
# the natural number of clusters — i.e. how many distinct groups of
# variables exist in the SMDW feature set.
plot(hclust_smdw, type = "index", main = "SMDW — aggregation levels")
```

## Cut tree and save

```{r}
# The shortlist sizes to test — from 5 representative variables up to the full set.
# The last value uses all available SMDW bands (no shortlisting).
k_values_smdw <- c(5, 10, 15, 20, 30, 45, 90, length(band_names_smdw))

# Loop over each shortlist size k and, for each one:
# 1. Cut the variable cluster tree into k groups
# 2. Identify one representative variable per group (the most correlated with its cluster)
# 3. Save the representative variables to a CSV
# 4. Save a dendrogram PDF showing the cluster boundaries
cluster_results_smdw <- purrr::map(k_values_smdw, function(k) {
  set.seed(123)
  message("SMDW — k = ", k)

  # Cut the hierarchical cluster tree into exactly k groups
  cuttree <- cutreevar(hclust_smdw, k = k)

  # Build a tidy table showing which cluster each variable was assigned to
  df_clusters <- cuttree$cluster %>%
    as.data.frame() %>%
    dplyr::rename(cluster = 1) %>%
    tibble::rownames_to_column("variable") %>%
    dplyr::arrange(cluster)

  # For each cluster, pick the variable most correlated with the cluster centroid
  # — this is the single variable that best represents that whole group
  reps <- purrr::map_chr(cuttree$var, function(cluster_df) {
    rownames(cluster_df)[which.max(cluster_df[, 1])]
  })

  # Pair each cluster number with its chosen representative variable name
  df_reps <- tibble::tibble(
    cluster                 = seq_along(reps),
    representative_variable = r
```

## Summary

```{r}
purrr::map_dfr(cluster_results_smdw, ~ tibble::tibble(
  k          = .x$k,
  n_clusters = nrow(.x$representatives)
)) %>%
  knitr::kable(caption = "SMDW — clusters per k")
```

------------------------------------------------------------------------

# 5. Sanitation cluster analysis

k values: 5, 10, 15, 20, 30, 45, 90, and the full set
(`r length(band_names_sanitation)`). Excludes the four JMP drinking
water variables not available in the sanitation training data.

## Prepare feature matrix

```{r}
samples_sanitation <- df_pred_covariates %>%
  dplyr::select(dplyr::all_of(band_names_sanitation)) %>%
  dplyr::mutate(dplyr::across(everything(), as.numeric))

constant_cols_san <- names(samples_sanitation)[
  sapply(samples_sanitation, function(x) dplyr::n_distinct(x, na.rm = TRUE) <= 1)
]
if (length(constant_cols_san) > 0) {
  message("Zero-variance columns (Sanitation):\n",
          paste(" -", constant_cols_san, collapse = "\n"))
} else {
  message("No zero-variance columns (Sanitation).")
}
cat("Complete cases:", sum(complete.cases(samples_sanitation)), "/",
    nrow(samples_sanitation), "\n")
```

## Hierarchical clustering

```{r}
set.seed(123)
hclust_sanitation <- hclustvar(X.quanti = as.data.frame(samples_sanitation))
plot(hclust_sanitation, type = "index", main = "Sanitation — aggregation levels")
```

## Cut tree and save

```{r}
k_values_sanitation <- c(5, 10, 15, 20, 30, 45, 90, length(band_names_sanitation))

cluster_results_sanitation <- purrr::map(k_values_sanitation, function(k) {
  set.seed(123)
  message("Sanitation — k = ", k)

  cuttree <- cutreevar(hclust_sanitation, k = k)

  df_clusters <- cuttree$cluster %>%
    as.data.frame() %>%
    dplyr::rename(cluster = 1) %>%
    tibble::rownames_to_column("variable") %>%
    dplyr::arrange(cluster)

  reps <- purrr::map_chr(cuttree$var, function(cluster_df) {
    rownames(cluster_df)[which.max(cluster_df[, 1])]
  })

  df_reps <- tibble::tibble(
    cluster                 = seq_along(reps),
    representative_variable = reps
  )

  readr::write_csv(
    df_reps,
    file.path(OUTPUT_DIR_SANITATION, paste0("representative_variables_k", k, ".csv"))
  )

  pdf_path <- file.path(OUTPUT_DIR_SANITATION, paste0("dendrogram_k", k, ".pdf"))
  pdf(pdf_path, width = 20, height = 12)
  tryCatch({
    plot(hclust_sanitation, main = paste("Sanitation — k =", k), cex = 0.5)
    rect.hclust(hclust_sanitation, k = k, border = "red")
  }, error = function(e) message("Plotting failed for k = ", k, ": ", e$message))
  dev.off()

  list(k = k, clusters = df_clusters, representatives = df_reps)
})

names(cluster_results_sanitation) <- paste0("k", k_values_sanitation)
message("Sanitation outputs written to: ", OUTPUT_DIR_SANITATION)
```

## Summary

```{r}
purrr::map_dfr(cluster_results_sanitation, ~ tibble::tibble(
  k          = .x$k,
  n_clusters = nrow(.x$representatives)
)) %>%
  knitr::kable(caption = "Sanitation — clusters per k")
```
