---
title: "Covariate Cluster Analysis"
subtitle: "Hierarchical clustering of geospatial covariates for feature selection"
author: ""
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

# 1. Setup

## Libraries

```{r}
library(tidyverse)
library(here)
library(ClustOfVar)
```

## Paths

```{r}
PATH_PRED_COVARIATES <- here::here("data/prediction/prediction_covariates_2024.csv")
OUTPUT_DIR           <- here::here("outputs/cluster_analysis")
```

------------------------------------------------------------------------

# 2. Input data

Prediction covariates prepared in `05_prepare_prediction_data.qmd`.
Each row is one GADM admin-1 region.

```{r}
df_pred_covariates <- readr::read_csv(PATH_PRED_COVARIATES, show_col_types = FALSE)

cat("Regions:", nrow(df_pred_covariates), "\n")
cat("Countries:", dplyr::n_distinct(df_pred_covariates$NAME_0), "\n")
cat("Columns:", ncol(df_pred_covariates), "\n")
```

------------------------------------------------------------------------

# 3. Variable selection

117 variables pre-selected based on their relevance for modelling access
to infrastructure and urban services, as well as anthropogenic,
hydrogeologic, climatic, topographic and biogeographic indicators used
to model safely managed drinking water services (Greenwood et al.,
2024).

```{r}
band_names <- c(
  # ── Static environmental ─────────────────────────────────────────────────
  "CGIAR_Aridity_Index",
  "CGIAR_PET",
  "CHELSA_BIO_Annual_Mean_Temperature",
  "CHELSA_BIO_Annual_Precipitation",
  "CHELSA_BIO_Precipitation_Seasonality",
  "CHELSA_BIO_Precipitation_of_Coldest_Quarter",
  "CHELSA_BIO_Precipitation_of_Driest_Month",
  "CHELSA_BIO_Precipitation_of_Driest_Quarter",
  "CHELSA_BIO_Precipitation_of_Warmest_Quarter",
  "CHELSA_BIO_Precipitation_of_Wettest_Month",
  "CHELSA_BIO_Precipitation_of_Wettest_Quarter",
  "CHELSA_BIO_Temperature_Annual_Range",
  "CHELSA_BIO_Temperature_Seasonality",
  "CIFOR_TropicalPeatlandExtent",
  "CSP_Global_Human_Modification",
  "ConsensusLandCoverClass_Barren",
  "ConsensusLandCoverClass_Cultivated_and_Managed_Vegetation",
  "ConsensusLandCoverClass_Deciduous_Broadleaf_Trees",
  "ConsensusLandCoverClass_Evergreen_Broadleaf_Trees",
  "ConsensusLandCoverClass_Evergreen_Deciduous_Needleleaf_Trees",
  "ConsensusLandCoverClass_Herbaceous_Vegetation",
  "ConsensusLandCoverClass_Mixed_Other_Trees",
  "ConsensusLandCoverClass_Open_Water",
  "ConsensusLandCoverClass_Regularly_Flooded_Vegetation",
  "ConsensusLandCoverClass_Shrubs",
  "ConsensusLandCoverClass_Snow_Ice",
  "ConsensusLandCoverClass_Urban_Builtup",
  "ConsensusLandCover_Human_Development_Percentage",
  "CrowtherLab_Tree_Density",
  "EarthEnvCloudCover_CloudForestPrediction",
  "EarthEnvTexture_CoOfVar_EVI",
  "EarthEnvTexture_Contrast_EVI",
  "EarthEnvTexture_Correlation_EVI",
  "EarthEnvTexture_Dissimilarity_EVI",
  "EarthEnvTexture_Entropy_EVI",
  "EarthEnvTexture_Evenness_EVI",
  "EarthEnvTexture_Homogeneity_EVI",
  "EarthEnvTexture_Maximum_EVI",
  "EarthEnvTexture_Range_EVI",
  "EarthEnvTexture_Shannon_Index",
  "EarthEnvTexture_Simpson_Index",
  "EarthEnvTexture_Std_EVI",
  "EarthEnvTexture_Uniformity_EVI",
  "EarthEnvTexture_Variance_EVI",
  "EarthEnvTopoMed_1stOrderPartialDerivEW",
  "EarthEnvTopoMed_1stOrderPartialDerivNS",
  "EarthEnvTopoMed_Eastness",
  "EarthEnvTopoMed_Elevation",
  "EarthEnvTopoMed_Roughness",
  "EarthEnvTopoMed_Slope",
  "EarthEnvTopoMed_TerrainRuggednessIndex",
  "EarthEnvTopoMed_TopoPositionIndex",
  "EsaCci_BurntAreasProbability",
  "FanEtAl_Depth_to_Water_Table_AnnualMean",
  "FanEtAl_Depth_to_Water_Table_AnnualSD",
  "GHS_Population_Density",
  "GPWv4_Population_Density",
  "GiriEtAl_MangrovesExtent",
  "GLW3_RuminantsDistribution_downsampled10km",
  "MODIS_EVI",
  "MODIS_NDVI",
  "MODIS_NPP",
  "PelletierEtAl_SoilAndSedimentaryDepositThicknesses",
  "SG_Absolute_depth_to_bedrock",
  "SG_Bulk_density_015cm",
  "SG_Depth_to_bedrock",
  "SG_H2O_Capacity_015cm",
  "SG_Saturated_H2O_Content_015cm",
  "TootchiEtAl_WetlandsRegularlyFlooded",
  "WCS_Human_Footprint_2009",
  "map_friction",
  # ── Year-specific ────────────────────────────────────────────────────────
  "chirps_annual_precipitation",
  "era5_temperature_2m",
  "ghsl_built_surface",
  "ghsl_population",
  "ghsl_population_sum",
  "ghsl_urban_frac",
  "jrc_building_height",
  "modis_evi",
  "modis_ndvi",
  "runoff_max_annualmax",
  "runoff_min_annualmin",
  "temperature_2m_max_annualmax",
  "viirs_average",
  # ── Governance (WGI) ─────────────────────────────────────────────────────
  "governance_effectiveness",
  "control_of_corruption",
  "political_stability",
  "regulatory_quality",
  "rule_of_law",
  "voice_and_accountability"
)

cat("Variables defined:", length(band_names), "\n")

# Check all variables are present in the prediction data
missing_vars <- setdiff(band_names, names(df_pred_covariates))
if (length(missing_vars) > 0) {
  warning("Variables in band_names not found in prediction data:\n",
          paste(missing_vars, collapse = "\n"))
} else {
  message("All variables present.")
}
```

## Prepare feature matrix

```{r}
samples_df <- df_pred_covariates %>%
  dplyr::select(dplyr::all_of(band_names)) %>%
  dplyr::mutate(dplyr::across(everything(), as.numeric))

# Flag zero-variance columns (constant across all regions)
constant_cols <- names(samples_df)[
  sapply(samples_df, function(x) dplyr::n_distinct(x, na.rm = TRUE) <= 1)
]
if (length(constant_cols) > 0) {
  message("Zero-variance columns (will cause hclustvar to fail):\n",
          paste(constant_cols, collapse = "\n"))
} else {
  message("No zero-variance columns.")
}

cat("Regions with complete cases:", sum(complete.cases(samples_df)), "/", nrow(samples_df), "\n")
```

------------------------------------------------------------------------

# 4. Hierarchical cluster analysis

Variables are clustered using `hclustvar` (ClustOfVar), which groups
variables by their shared information content rather than their
pairwise correlations. One representative variable is later selected
per cluster.

```{r}
set.seed(123)
hclust_result <- hclustvar(X.quanti = as.data.frame(samples_df))
plot(hclust_result, type = "index")
```

------------------------------------------------------------------------

# 5. Cut tree and select representative variables

The dendrogram is cut at several values of k. For each k, the variable
with the highest squared loading on its cluster component is selected
as representative.

```{r}
k_values <- c(5, 7, 12, 17, 25, 45, 60)

dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)

cluster_results <- purrr::map(k_values, function(k) {
  set.seed(123)
  message("Processing k = ", k, " ...")

  cuttree <- cutreevar(hclust_result, k = k)

  # Cluster membership for every variable
  df_clusters <- cuttree$cluster %>%
    as.data.frame() %>%
    dplyr::rename(cluster = 1) %>%
    tibble::rownames_to_column("variable") %>%
    dplyr::arrange(cluster)

  # Representative variable per cluster (highest squared loading)
  reps <- purrr::map_chr(cuttree$var, function(cluster_df) {
    rownames(cluster_df)[which.max(cluster_df[, 1])]
  })

  df_reps <- tibble::tibble(
    cluster                  = seq_along(reps),
    representative_variable  = reps
  )

  # Save representative variables
  readr::write_csv(
    df_reps,
    file.path(OUTPUT_DIR, paste0("representative_variables_k", k, ".csv"))
  )

  # Save dendrogram PDF
  pdf_path <- file.path(OUTPUT_DIR, paste0("dendrogram_k", k, ".pdf"))
  pdf(pdf_path, width = 20, height = 12)
  tryCatch({
    plot(hclust_result, main = paste("k =", k), cex = 0.5)
    rect.hclust(hclust_result, k = k, border = "red")
  }, error = function(e) {
    message("Plotting failed for k = ", k, ": ", e$message)
  })
  dev.off()

  list(k = k, clusters = df_clusters, representatives = df_reps)
})

names(cluster_results) <- paste0("k", k_values)

message("Done. Outputs written to: ", OUTPUT_DIR)
```

------------------------------------------------------------------------

# 6. Summary

```{r}
purrr::map_dfr(cluster_results, function(res) {
  tibble::tibble(
    k                    = res$k,
    n_clusters           = nrow(res$representatives),
    representative_vars  = paste(res$representatives$representative_variable, collapse = ", ")
  )
}) %>%
  dplyr::select(k, n_clusters) %>%
  knitr::kable(caption = "Number of clusters per k")
```
