---
title: "cluster_analysis"
output: html_document
---

```{r}
library(data.table)
library(factoextra)
library(tidyverse)
library(Hmisc)
library(readr)
library(dplyr)
library(tibble)

```
```{r}
features_pred_df <- read_csv("/Users/esthergreenwood/Documents/GitHub/mapping_sanitation_access_LMICs/outputs/03_sampled_features/worldbank_sampled_pred.csv")

print(names(features_pred_df))
```

## Pre-selected environmental varibales

We pre-elected 117 variables based on their relevance for modelling access to infrastructure and urban services as well as anthropogenic, hydrogeologic, climatic, topographic and biogeographic indicators used to model safely managed drinking water services (Greenwood et al., 2024). 

```{r}
bandNames <- c(
  "CGIAR_Aridity_Index",
  "CGIAR_PET",
  "CHELSA_BIO_Annual_Mean_Temperature",
  "CHELSA_BIO_Annual_Precipitation",
  "CHELSA_BIO_Precipitation_Seasonality",
  "CHELSA_BIO_Precipitation_of_Coldest_Quarter",
  "CHELSA_BIO_Precipitation_of_Driest_Month",
  "CHELSA_BIO_Precipitation_of_Driest_Quarter",
  "CHELSA_BIO_Precipitation_of_Warmest_Quarter",
  "CHELSA_BIO_Precipitation_of_Wettest_Month",
  "CHELSA_BIO_Precipitation_of_Wettest_Quarter",
  "CHELSA_BIO_Temperature_Annual_Range",
  "CHELSA_BIO_Temperature_Seasonality",
  "CIFOR_TropicalPeatlandExtent",
  "CSP_Global_Human_Modification",
  "ConsensusLandCoverClass_Barren",
  "ConsensusLandCoverClass_Cultivated_and_Managed_Vegetation",
  "ConsensusLandCoverClass_Deciduous_Broadleaf_Trees",
  "ConsensusLandCoverClass_Evergreen_Broadleaf_Trees",
  "ConsensusLandCoverClass_Evergreen_Deciduous_Needleleaf_Trees",
  "ConsensusLandCoverClass_Herbaceous_Vegetation",
  "ConsensusLandCoverClass_Mixed_Other_Trees",
  "ConsensusLandCoverClass_Open_Water",
  "ConsensusLandCoverClass_Regularly_Flooded_Vegetation",
  "ConsensusLandCoverClass_Shrubs",
  "ConsensusLandCoverClass_Snow_Ice",
  "ConsensusLandCoverClass_Urban_Builtup",
  "ConsensusLandCover_Human_Development_Percentage",
  "CrowtherLab_Tree_Density",
  "EarthEnvCloudCover_CloudForestPrediction",
  "EarthEnvTexture_CoOfVar_EVI",
  "EarthEnvTexture_Contrast_EVI",
  "EarthEnvTexture_Correlation_EVI",
  "EarthEnvTexture_Dissimilarity_EVI",
  "EarthEnvTexture_Entropy_EVI",
  "EarthEnvTexture_Evenness_EVI",
  "EarthEnvTexture_Homogeneity_EVI",
  "EarthEnvTexture_Maximum_EVI",
  "EarthEnvTexture_Range_EVI",
  "EarthEnvTexture_Shannon_Index",
  "EarthEnvTexture_Simpson_Index",
  "EarthEnvTexture_Std_EVI",
  "EarthEnvTexture_Uniformity_EVI",
  "EarthEnvTexture_Variance_EVI",
  "EarthEnvTopoMed_1stOrderPartialDerivEW",
  "EarthEnvTopoMed_1stOrderPartialDerivNS",
  "EarthEnvTopoMed_Eastness",
  "EarthEnvTopoMed_Elevation",
  "EarthEnvTopoMed_Roughness",
  "EarthEnvTopoMed_Slope",
  "EarthEnvTopoMed_TerrainRuggednessIndex",
  "EarthEnvTopoMed_TopoPositionIndex",
  "EsaCci_BurntAreasProbability",
  "FanEtAl_Depth_to_Water_Table_AnnualMean",
  "FanEtAl_Depth_to_Water_Table_AnnualSD",
  "GHS_Population_Density",
  "GPWv4_Population_Density",
  "GiriEtAl_MangrovesExtent",
  "GLW3_RuminantsDistribution_downsampled10km",
  "MODIS_EVI",
  "MODIS_NDVI",
  "MODIS_NPP",
  "PelletierEtAl_SoilAndSedimentaryDepositThicknesses",
  "SG_Absolute_depth_to_bedrock",
  "SG_Bulk_density_015cm",
  "SG_Depth_to_bedrock",
  "SG_H2O_Capacity_015cm",
  "SG_Saturated_H2O_Content_015cm",
  "TootchiEtAl_WetlandsRegularlyFlooded",
  "WCS_Human_Footprint_2009",
  "map_friction",

  # sampled for specific year
  "chirps_annual_precipitation",
  "era5_temperature_2m",
  "ghsl_built_surface",
  "ghsl_population",
  "ghsl_population_sum",
  "ghsl_urban_frac",
  "jrc_building_height",
  "modis_evi",
  "modis_ndvi",
  "runoff_max_annualmax",
  "runoff_min_annualmin",
  "temperature_2m_max_annualmax",
  "viirs_average"
  #"worldpop_sum"
)

length(bandNames)
```
```{r}
samples_lmics_df <- features_pred_df %>% select(all_of(bandNames))
#%>%
  #na.omit()

samples_lmics_df <- samples_lmics_df %>% mutate(across(all_of(bandNames), as.numeric))

# Returns the names of columns with zero variance
constant_cols <- names(samples_lmics_df)[sapply(samples_lmics_df, function(x) length(unique(x)) <= 1)]
print(constant_cols)

```


## Hierarchical cluster analysis. Here we cluster into 24, 39, 51, 73 and 91 groups. Later we select only one feature from each group.
```{r}

library(ClustOfVar)


# Convert to a standard data frame and only include the numeric columns
hclust <- hclustvar(X.quanti = as.data.frame(samples_lmics_df[, bandNames]))
plot(hclust,type="index")

  

```


```{r}
set.seed(123)

k_values <- c(5,9,13,15,27,40, 50, 60)

for (k in k_values) {
  set.seed(123) 
  
  message(paste("Processing k =", k, "..."))
  
  # 1. Perform the cut
  cuttree <- cutreevar(hclust, k = k)
  
  # 2. Open PDF
  pdf_filename <- paste0("dendrogram_k", k, ".pdf")
  pdf(pdf_filename, width = 20, height = 12) 
  
  # Use tryCatch to ensure dev.off() runs even if plotting fails
  tryCatch({
    # 3. Generate the Plot 
    plot(hclust, main = paste("Clusters:", k), cex = 0.5)
    
    # 4. Draw boxes
    # We use 'which' to ensure it draws correctly for every specific k
    rect.hclust(hclust, k = k, border = "red") 
    
  }, error = function(e) {
    message(paste("Plotting failed for k =", k, ":", e$message))
  })
  
  # 5. ALWAYS close the PDF device
  dev.off()
  
  # 6. Save data frame
  df <- cuttree$cluster %>% 
    as.data.frame() %>% 
    rename(cluster = 1) %>% 
    rownames_to_column("variable") %>% 
    arrange(cluster)
  
  assign(paste0("df_clusters_k", k), df)
}

```

```{r}
  set.seed(123) 
# Create a list to store the representative features for each k
selected_features_summary <- list()

for (k in k_values) {
    set.seed(123) 
  
  # 1. Access the loadings for this specific k
  # (Requires the 'cuttree' object from your previous loop)
  cuttree_temp <- cutreevar(hclust, k = k)
  loadings <- cuttree_temp$var
  
  # 2. Find the variable with the highest squared loading in each cluster
  reps <- sapply(loadings, function(cluster_df) {
    rownames(cluster_df)[which.max(cluster_df[, 1])]
  })
  
  # 3. Create a clean data frame of the results
  rep_df <- data.frame(
    cluster = seq_along(reps),
    representative_variable = reps,
    stringsAsFactors = FALSE
  )
  
  # 4. Save the result to your environment
  assign(paste0("reps_k", k), rep_df)
  selected_features_summary[[paste0("k", k)]] <- reps
  
  message(paste("Representative variables selected for k =", k))
}

# Example: To see the features selected for k=13, run:
# print(reps_k13)


```

```{r}
for (k in k_values) {
  
  # Get the representative variables for this k
  # Using the data frame created in the previous step
  rep_data <- get(paste0("reps_k", k))
  
  #  Define the filename
  csv_filename <- paste0("representative_variables_k", k, ".csv")
  
  #  Write to CSV
  # row.names = FALSE ensures you don't get an extra index column
  write.csv(rep_data, file = csv_filename, row.names = FALSE)
  
  message(paste("CSV file created:", csv_filename))
}
```

