---
title: "Prepare Sanitation Model Training Data"
subtitle: "Basic sanitation and open defecation outcomes with geospatial covariates"
author: ""
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

# Setup

## Libraries and sourced functions

```{r}
library(tidyverse)
library(here)
library(fuzzyjoin)
library(stringdist)
library(stringi)

source(here::here("functions/create_indicators.R"))
source(here::here("functions/create_sanitation_indicators.R"))
source(here::here("functions/join_and_structure_dataframes.R"))
source(here::here("functions/diagnostics.R"))           # generalised diagnostic helpers
source(here::here("functions/country_join_helpers.R"))  # unified add_country_join_key_extended()
source(here::here("configuration/covariate_cols.R"))           # loads `covariate_cols` vector
source(here::here("configuration/manual_region_fixes.R"))      # loads `manual_fixes_sanitation` tibble
```

## Paths

All raw data paths are defined here so they are easy to update when data
moves.

```{r}
PATH_TO_POP_SUMS_OTHER <- "~/switchdrive/Eawag/WorldBankProject/geospatial_covariates/pop_sums_training_other"
FILE_NONPOP_OTHER      <- "~/switchdrive/Eawag/WorldBankProject/geospatial_covariates/other_surveys_covariates_nonpop_v2.csv"

PATH_MICS_SANITATION   <- "~/switchdrive/Eawag/WorldBankProject/HH_surveys/HH_survey_data/df_sanitation_MICS_v1.csv"
PATH_DHS_SANITATION    <- "~/switchdrive/Eawag/WorldBankProject/HH_surveys/HH_survey_data/df_sanitation_DHS.csv"

PATH_COUNTRY_KEY       <- here::here("data/country_name_key_WB.csv")
PATH_GDP               <- "~/switchdrive/Eawag/WorldBankProject/geospatial_covariates/gdp_per_capita_2015.csv"
PATH_EDUCATION         <- "~/switchdrive/Eawag/WorldBankProject/geospatial_covariates/secondary_education_years.csv"
PATH_WASTEWATER        <- "~/switchdrive/Eawag/WorldBankProject/geospatial_covariates/Country_WWctr_Percentage.txt"

OUTPUT_DIR             <- here::here("data/training_subcomponents") #writing training data into this file
CROSSWALK_REVIEW_DIR   <- here::here("data/crosswalk_review")
```

------------------------------------------------------------------------

# Input data

## Household survey data

MICS and DHS sanitation survey extracts are loaded and combined. Each
row is one household observation with region and year identifiers.

```{r}
df_mics <- readr::read_csv(PATH_MICS_SANITATION, show_col_types = FALSE) %>%
  dplyr::mutate(household_data_source = "MICS_sanitation")

df_dhs <- readr::read_csv(PATH_DHS_SANITATION, show_col_types = FALSE) %>%
  dplyr::mutate(household_data_source = "DHS_sanitation")


df_hh_sanitation <- dplyr::bind_rows(df_mics, df_dhs)

df_hh_sanitation <- df_hh_sanitation %>%
  dplyr::mutate(
    country = dplyr::case_when(
      country == "DR Congo" ~ "Democratic Republic of the Congo",
      TRUE ~ country
    )
  )
```

## Lookup tables and metadata

```{r}
df_wgi_lmics        <- readWGI_lmics()
country_name_key_WB <- readr::read_csv(PATH_COUNTRY_KEY, show_col_types = FALSE)
```

------------------------------------------------------------------------

# Survey selection

## Filter to study period (2015–2024)

Surveys outside the modelling window are dropped. Surveys near the
boundary are flagged before filtering so the exclusions are visible.

```{r}
df_hh_sanitation <- df_hh_sanitation %>%
  dplyr::mutate(
    HH5Y_original = HH5Y,
    HH5Y          = standardise_survey_year(country, HH5Y)
  )

surveys_outside_range <- df_hh_sanitation %>%
  dplyr::filter(!is.na(HH5Y), HH5Y < 2015 | HH5Y > 2025) %>%
  dplyr::distinct(country, HH5Y_original, HH5Y) %>%
  dplyr::arrange(country, HH5Y)
print(surveys_outside_range)

countries_with_2024 <- df_hh_sanitation %>%
  dplyr::filter(HH5Y == 2024) %>%
  dplyr::distinct(country)

df_hh_sanitation <- df_hh_sanitation %>%
  dplyr::filter(
    is.na(HH5Y) | (HH5Y >= 2015 & HH5Y <= 2024) |
    (HH5Y == 2025 & country %in% countries_with_2024$country)
  )

surveys_removed <- df_hh_sanitation %>%
  dplyr::anti_join(
    df_hh_sanitation %>% dplyr::filter(
      is.na(HH5Y) | (HH5Y >= 2015 & HH5Y <= 2024) |
      (HH5Y == 2025 & country %in% countries_with_2024$country)
    ),
    by = names(df_hh_sanitation)
  )
```

## Select most recent data source per country

Where a country has both MICS and DHS data, we keep the source with the
most recent survey year. Ties are broken by row count, then source name.

```{r}
country_source_summary <- df_hh_sanitation %>%
  dplyr::mutate(HH5Y = as.integer(HH5Y)) %>%
  dplyr::group_by(country, household_data_source) %>%
  dplyr::summarise(
    min_year     = ifelse(all(is.na(HH5Y)), NA_integer_, min(HH5Y, na.rm = TRUE)),
    max_year     = ifelse(all(is.na(HH5Y)), NA_integer_, max(HH5Y, na.rm = TRUE)),
    survey_years = paste(sort(unique(HH5Y[!is.na(HH5Y)])), collapse = ", "),
    n_rows       = dplyr::n(),
    .groups      = "drop"
  )

knitr::kable(
  country_source_summary %>% dplyr::arrange(country, household_data_source),
  caption = "Available surveys by country and data source"
)

source_to_keep <- country_source_summary %>%
  dplyr::group_by(country) %>%
  dplyr::arrange(desc(!is.na(max_year)), desc(max_year), desc(n_rows), household_data_source) %>%
  dplyr::slice(1) %>%
  dplyr::ungroup() %>%
  dplyr::select(country, household_data_source)

source_to_drop <- country_source_summary %>%
  dplyr::anti_join(source_to_keep, by = c("country", "household_data_source")) %>%
  dplyr::arrange(country, household_data_source)

knitr::kable(
  source_to_drop,
  caption = "Sources dropped (superseded by more recent data)"
)

df_hh_sanitation <- df_hh_sanitation %>%
  dplyr::semi_join(source_to_keep, by = c("country", "household_data_source"))

# Drop sub-national Pakistan MICS surveys — use national DHS instead
pakistan_mics_dropped <- df_hh_sanitation %>%
  dplyr::filter(stringr::str_detect(country, "Pakistan") &
                household_data_source == "MICS_sanitation") %>%
  dplyr::distinct(country, HH5Y, household_data_source)

knitr::kable(
  pakistan_mics_dropped,
  caption = "Pakistan sub-national MICS surveys dropped in favour of national DHS"
)

df_hh_sanitation <- df_hh_sanitation %>%
  dplyr::filter(!(stringr::str_detect(country, "Pakistan") &
                  household_data_source == "MICS_sanitation"))
```

------------------------------------------------------------------------

# Geospatial covariates

## Build covariate table

Population-weighted and non-population covariates are merged into a
single region-level covariate table. The `covariate_cols` vector is
defined in `config/covariate_cols.R`.

```{r}
pop_files_other <- list.files(
  path       = path.expand(PATH_TO_POP_SUMS_OTHER),
  pattern    = "\\.csv$",
  full.names = TRUE
)

df_training_covariates_sanitation <- make_training_covariates(
  pop_files      = pop_files_other,
  nonpop_files   = FILE_NONPOP_OTHER,
  covariate_cols = covariate_cols
)
```

## Check for duplicate covariate keys

Duplicate country–region–year keys would cause row inflation on joining
and must be resolved before proceeding.

```{r}
duplicate_covariate_keys <- check_duplicate_covariate_keys(df_training_covariates_sanitation)
duplicate_covariate_keys
```

------------------------------------------------------------------------

## Restrict to most recent survey year per country

A single data source may contain multiple survey rounds. We keep only
the most recent year per country before building outcomes, so that the
year lookup and aggregation use only the retained survey.

```{r}
keep_latest_survey_year_hh <- function(df, max_survey_duration = 2) {
  df %>%
    dplyr::group_by(country) %>%
    dplyr::filter(HH5Y >= max(HH5Y, na.rm = TRUE) - max_survey_duration) %>%
    dplyr::ungroup()
}

df_hh_sanitation_latest <- keep_latest_survey_year_hh(df_hh_sanitation)

```

## Build regional outcomes

Household-level responses are aggregated to region–year proportions for
each outcome. Outcome definitions:

-   **Basic sanitation**: improved facility (`WS11 == 2`) AND not shared
    (`WS15 == 0`)
-   **Open defecation**: no facility / bush / field (`WS11 == 0`)

```{r}
basic_sanitation_outcome <- make_basic_sanitation_regional_outcome(df_hh_sanitation_latest)
open_defecation_outcome  <- make_open_defecation_regional_outcome(df_hh_sanitation_latest)
```

## Add PSU weights

Survey-design weights based on the number of primary sampling units are
attached to each region–year row.

```{r}
sanitation_year_lookup <- make_year_lookup_from_raw_household_data(
  df_hh_sanitation_latest,
  manual_analysis_year_fixes = tibble::tribble(
    ~country,   ~analysis_year_manual,
    "Nigeria",  2024L
  )
)

basic_sanitation_outcome <- add_n_psu_weight(
  hh_df       = df_hh_sanitation,
  outcome_df  = basic_sanitation_outcome,
  year_lookup = sanitation_year_lookup
)
open_defecation_outcome <- add_n_psu_weight(
  hh_df       = df_hh_sanitation,
  outcome_df  = open_defecation_outcome,
  year_lookup = sanitation_year_lookup
)

# Quick sanity check on PSU coverage
dplyr::bind_rows(
  basic_sanitation_outcome %>% dplyr::mutate(outcome = "basic_sanitation"),
  open_defecation_outcome  %>% dplyr::mutate(outcome = "open_defecation")
) %>%
  dplyr::group_by(outcome) %>%
  dplyr::summarise(
    n_rows        = dplyr::n(),
    missing_n_psu = sum(is.na(n_psu)),
    min_n_psu     = min(n_psu, na.rm = TRUE),
    median_n_psu  = median(n_psu, na.rm = TRUE),
    max_n_psu     = max(n_psu, na.rm = TRUE),
    .groups = "drop"
  )
```

------------------------------------------------------------------------

# Covariate–outcome matching

## Filter covariates to outcome country–years

The full covariate table is subset to only the country–year combinations
present in each outcome, using a country name crosswalk to harmonise
naming.

```{r}
df_covariates_basic_sanitation <- filter_covariates_to_outcome_years(
  outcome_df             = basic_sanitation_outcome,
  df_training_covariates = df_training_covariates_sanitation,
  country_name_key_WB    = country_name_key_WB
)
df_covariates_open_defecation <- filter_covariates_to_outcome_years(
  outcome_df             = open_defecation_outcome,
  df_training_covariates = df_training_covariates_sanitation,
  country_name_key_WB    = country_name_key_WB
)

check_duplicate_covariate_join_keys(df_covariates_basic_sanitation)
check_duplicate_covariate_join_keys(df_covariates_open_defecation)
```

## Diagnose unmatched country–years

Countries present in the outcome but absent from filtered covariates are
flagged here. Common causes: country name mismatch or missing covariate
data.

```{r}
country_year_filtering_issues <- dplyr::bind_rows(
  diagnose_country_year_filtering(
    outcome_name        = "basic_sanitation",
    outcome_df          = basic_sanitation_outcome,
    raw_covariates      = df_training_covariates_sanitation,
    filtered_covariates = df_covariates_basic_sanitation,
    country_name_key_WB = country_name_key_WB
  ),
  diagnose_country_year_filtering(
    outcome_name        = "open_defecation",
    outcome_df          = open_defecation_outcome,
    raw_covariates      = df_training_covariates_sanitation,
    filtered_covariates = df_covariates_open_defecation,
    country_name_key_WB = country_name_key_WB
  )
)
print(country_year_filtering_issues)
```

## Build region crosswalks

Survey region names are matched to covariate region names using fuzzy
string matching. Rows that cannot be matched automatically are flagged
for manual review.

```{r}
crosswalk_basic_sanitation <- build_crosswalk_for_one_outcome(
  outcome_name           = "basic_sanitation",
  outcome_df             = basic_sanitation_outcome,
  df_training_covariates = df_covariates_basic_sanitation,
  country_name_key_WB    = country_name_key_WB
)
crosswalk_open_defecation <- build_crosswalk_for_one_outcome(
  outcome_name           = "open_defecation",
  outcome_df             = open_defecation_outcome,
  df_training_covariates = df_covariates_open_defecation,
  country_name_key_WB    = country_name_key_WB
)

# Inspect rows that need manual attention
crosswalk_basic_sanitation$regions_to_review
crosswalk_open_defecation$regions_to_review
```

## Apply manual region name fixes

Known mismatches are resolved via the lookup table in
`config/manual_region_fixes.R`. Add new entries there (not here) when
new surveys are incorporated.

```{r}
basic_sanitation_crosswalk_final <- apply_manual_fixes_for_one_outcome(
  outcome_name        = "basic_sanitation",
  crosswalk_results   = crosswalk_basic_sanitation,
  manual_region_fixes = manual_fixes_sanitation
)
open_defecation_crosswalk_final <- apply_manual_fixes_for_one_outcome(
  outcome_name        = "open_defecation",
  crosswalk_results   = crosswalk_open_defecation,
  manual_region_fixes = manual_fixes_sanitation
)
```

## Join outcomes to covariates

```{r}
basic_sanitation_training_output <- join_and_save_one_training_dataset(
  outcome_name           = "basic_sanitation",
  outcome_df             = basic_sanitation_outcome,
  df_training_covariates = df_covariates_basic_sanitation,
  region_crosswalk_final = basic_sanitation_crosswalk_final
)
open_defecation_training_output <- join_and_save_one_training_dataset(
  outcome_name           = "open_defecation",
  outcome_df             = open_defecation_outcome,
  df_training_covariates = df_covariates_open_defecation,
  region_crosswalk_final = open_defecation_crosswalk_final
)

basic_sanitation_training <- basic_sanitation_training_output$training_with_covariates
open_defecation_training  <- open_defecation_training_output$training_with_covariates

basic_sanitation_training_output$missing_covariates
open_defecation_training_output$missing_covariates
```

------------------------------------------------------------------------

# Country-level covariates

## GDP per capita and secondary education

Country-year level economic and education indicators from the World Bank
are joined by standardised country name.

```{r}
gdp_clean <- readr::read_csv(PATH_GDP, show_col_types = FALSE) %>%
  dplyr::rename(
    country_name                     = Country.Name,
    analysis_year                    = Year,
    gdp_per_capita_constant_2015_usd = `GDP per capita (constant 2015 US$)`
  ) %>%
  dplyr::mutate(
    analysis_year                    = as.integer(analysis_year),
    gdp_per_capita_constant_2015_usd = as.numeric(gdp_per_capita_constant_2015_usd)
  ) %>%
  add_country_join_key_extended(
    country_col = country_name, country_name_key_WB = country_name_key_WB
  ) %>%
  dplyr::select(country_join, analysis_year, gdp_per_capita_constant_2015_usd) %>%
  dplyr::distinct()

education_clean <- readr::read_csv(PATH_EDUCATION, show_col_types = FALSE) %>%
  dplyr::rename(
    country_name                       = Country.Name,
    analysis_year                      = Year,
    secondary_education_duration_years = `Secondary education, duration (years)`
  ) %>%
  dplyr::mutate(
    analysis_year                      = as.integer(analysis_year),
    secondary_education_duration_years = as.numeric(secondary_education_duration_years)
  ) %>%
  add_country_join_key_extended(
    country_col = country_name, country_name_key_WB = country_name_key_WB
  ) %>%
  dplyr::select(country_join, analysis_year, secondary_education_duration_years) %>%
  dplyr::distinct()

country_year_covariates <- dplyr::full_join(
  gdp_clean, education_clean, by = c("country_join", "analysis_year")
)

add_country_year_covariates <- function(training_df) {
  training_df %>%
    dplyr::mutate(analysis_year = as.integer(analysis_year)) %>%
    add_country_join_key_extended(
      country_col = country_outcome, country_name_key_WB = country_name_key_WB
    ) %>%
    dplyr::left_join(country_year_covariates, by = c("country_join", "analysis_year")) %>%
    dplyr::select(-country_join)
}

basic_sanitation_training <- add_country_year_covariates(basic_sanitation_training)
open_defecation_training  <- add_country_year_covariates(open_defecation_training)

check_missing_covariates(basic_sanitation_training, "basic_sanitation",
  c("gdp_per_capita_constant_2015_usd", "secondary_education_duration_years"))
check_missing_covariates(open_defecation_training, "open_defecation",
  c("gdp_per_capita_constant_2015_usd", "secondary_education_duration_years"))
```

## Wastewater treatment

Country-level wastewater collection, treatment, and reuse percentages.
These are static (no year dimension) and joined by country name only.

```{r}
ww_country_clean <- readr::read_tsv(PATH_WASTEWATER, show_col_types = FALSE) %>%
  dplyr::rename(
    ww_country_name            = Country,
    ww_region                  = Region,
    ww_economic_classification = Economic_Classification,
    ww_collection_percent      = WWc_Percent,
    ww_treatment_percent       = WWt_Percent,
    ww_reuse_percent           = WWr_Percent
  ) %>%
  dplyr::mutate(across(ends_with("_percent"), as.numeric)) %>%
  add_country_join_key_extended(
    country_col = ww_country_name, country_name_key_WB = country_name_key_WB
  ) %>%
  dplyr::distinct(country_join, .keep_all = TRUE)

add_ww_covariates <- function(training_df) {
  training_df %>%
    dplyr::select(-dplyr::any_of(c(
      "country_join", "ww_country_name", "ww_region",
      "ww_economic_classification",
      "ww_collection_percent", "ww_treatment_percent", "ww_reuse_percent"
    ))) %>%
    add_country_join_key_extended(
      country_col = country_outcome, country_name_key_WB = country_name_key_WB
    ) %>%
    dplyr::left_join(
      ww_country_clean %>% dplyr::select(
        country_join, ww_region, ww_economic_classification,
        ww_collection_percent, ww_treatment_percent, ww_reuse_percent
      ),
      by = "country_join"
    ) %>%
    dplyr::select(-country_join)
}

basic_sanitation_training <- add_ww_covariates(basic_sanitation_training)
open_defecation_training  <- add_ww_covariates(open_defecation_training)

# Kosovo has no wastewater record; assign region manually
basic_sanitation_training <- basic_sanitation_training %>%
  dplyr::mutate(ww_region = dplyr::if_else(
    country_outcome == "Kosovo" & is.na(ww_region),
    "Eastern Europe & Central Asia", ww_region
  ))
open_defecation_training <- open_defecation_training %>%
  dplyr::mutate(ww_region = dplyr::if_else(
    country_outcome == "Kosovo" & is.na(ww_region),
    "Eastern Europe & Central Asia", ww_region
  ))

check_missing_covariates(basic_sanitation_training, "basic_sanitation",
  c("ww_collection_percent", "ww_treatment_percent", "ww_reuse_percent"))
check_missing_covariates(open_defecation_training, "open_defecation",
  c("ww_collection_percent", "ww_treatment_percent", "ww_reuse_percent"))
```

## WGI governance indicators

Six World Governance Indicators are joined by country and year.

```{r}
indicator_map <- c(
  "Government Effectiveness - Governance score (0-100)" = "governance_effectiveness",
  "Control of Corruption - Governance score (0-100)"    = "control_of_corruption",
  "Political Stability - Governance score (0-100)"      = "political_stability",
  "Regulatory Quality - Governance score (0-100)"       = "regulatory_quality",
  "Rule of Law - Governance score (0-100)"              = "rule_of_law",
  "Voice and Accountability - Governance score (0-100)" = "voice_and_accountability"
)

wgi_clean <- df_wgi_lmics %>%
  dplyr::filter(`Series.Name` %in% names(indicator_map)) %>%
  dplyr::mutate(indicator = unname(indicator_map[`Series.Name`])) %>%
  tidyr::pivot_longer(
    cols      = dplyr::matches("^X20[0-9]{2}$"),
    names_to  = "analysis_year",
    values_to = "value"
  ) %>%
  dplyr::mutate(
    analysis_year = as.integer(stringr::str_remove(analysis_year, "^X")),
    value         = as.numeric(value)
  ) %>%
  add_country_join_key_extended(
    country_col = `Country.Name`, country_name_key_WB = country_name_key_WB
  ) %>%
  dplyr::select(country_join, analysis_year, indicator, value) %>%
  tidyr::pivot_wider(names_from = indicator, values_from = value) %>%
  dplyr::distinct()

add_wgi_covariates <- function(training_df) {
  training_df %>%
    dplyr::select(-dplyr::any_of(c("country_join", unname(indicator_map)))) %>%
    dplyr::mutate(analysis_year = as.integer(analysis_year)) %>%
    add_country_join_key_extended(
      country_col = country_outcome, country_name_key_WB = country_name_key_WB
    ) %>%
    dplyr::left_join(wgi_clean, by = c("country_join", "analysis_year")) %>%
    dplyr::select(-country_join)
}

basic_sanitation_training <- add_wgi_covariates(basic_sanitation_training)
open_defecation_training  <- add_wgi_covariates(open_defecation_training)

check_missing_covariates(basic_sanitation_training, "basic_sanitation", unname(indicator_map))
check_missing_covariates(open_defecation_training,  "open_defecation",  unname(indicator_map))
```

------------------------------------------------------------------------

# Output

## Write training CSVs

Final training datasets are written to `data/training_subcomponents/`.

```{r}
dir.create(OUTPUT_DIR,           recursive = TRUE, showWarnings = FALSE)
dir.create(CROSSWALK_REVIEW_DIR, recursive = TRUE, showWarnings = FALSE)

readr::write_csv(
  basic_sanitation_training,
  file.path(OUTPUT_DIR, "basic_sanitation_training_with_covariates.csv")
)
readr::write_csv(
  open_defecation_training,
  file.path(OUTPUT_DIR, "open_defecation_training_with_covariates.csv")
)

message("Written: basic_sanitation_training_with_covariates.csv")
message("Written: open_defecation_training_with_covariates.csv")
```

## Summary: wastewater region distribution

A quick check on regional representation in the final training data.

```{r}
dplyr::bind_rows(
  basic_sanitation_training %>% dplyr::mutate(outcome_type = "basic_sanitation"),
  open_defecation_training  %>% dplyr::mutate(outcome_type = "open_defecation")
) %>%
  dplyr::count(outcome_type, ww_region, name = "n_observations") %>%
  dplyr::group_by(outcome_type) %>%
  dplyr::mutate(percent = round(100 * n_observations / sum(n_observations), 2)) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(outcome_type, desc(n_observations))
```
