---
title: "Prepare Sanitation Model Training Data"
subtitle: "Basic sanitation and open defecation outcomes with geospatial covariates"
---

# 1. Setup

## Libraries and sourced functions

```{r}
library(tidyverse)
library(here)
library(fuzzyjoin)
library(stringdist)
library(stringi)

source(here::here("functions/create_SMDW_indicators.R"))
source(here::here("functions/create_sanitation_indicators.R"))
source(here::here("functions/join_and_structure_dataframes.R"))
source(here::here("functions/diagnostics.R"))           # generalised diagnostic helpers
source(here::here("functions/country_join_helpers.R"))  # unified add_country_join_key_extended()
source(here::here("configuration/covariate_cols.R"))           # loads `covariate_cols` vector
source(here::here("configuration/manual_region_fixes.R"))      # loads `manual_fixes_sanitation` tibble
source(here::here("configuration/paths.R"))
```

## Paths

All raw data paths are defined here so they are easy to update when data moves.

```{r}

PATH_MICS_SANITATION   <- here::here("data/processed/household_surveys/df_sanitation_MICS_v2.csv")
PATH_DHS_SANITATION    <- here::here("data/processed/household_surveys/df_sanitation_DHS_v2.csv")

PATH_COUNTRY_KEY       <- here::here("configuration/country_name_key_WB.csv")
PATH_TO_REGION_AREAS <- here::here("data/processed/survey_boundaries/other_surveys/region_areas_km2.csv")

OUTPUT_DIR             <- here::here("data/processed/training_subcomponents") #writing training data into this file
CROSSWALK_REVIEW_DIR   <- here::here("data/processed/crosswalk_review")

```

------------------------------------------------------------------------

# 2. Input data

## Household survey data

MICS and DHS sanitation survey extracts are loaded and combined. Each row is one household observation with region and year identifiers.

```{r}
df_mics <- readr::read_csv(PATH_MICS_SANITATION, show_col_types = FALSE) %>%
  dplyr::mutate(household_data_source = "MICS_sanitation")

df_dhs <- readr::read_csv(PATH_DHS_SANITATION, show_col_types = FALSE) %>%
  dplyr::mutate(household_data_source = "DHS_sanitation")


df_hh_sanitation <- dplyr::bind_rows(df_mics, df_dhs)

df_hh_sanitation <- df_hh_sanitation %>%
  dplyr::mutate(
    country = dplyr::case_when(
      country == "DR Congo" ~ "Democratic Republic of the Congo",
      TRUE ~ country
    )
  )
```

## Lookup tables and metadata

```{r}
df_wgi_lmics        <- readWGI_lmics()
country_name_key_WB <- readr::read_csv(PATH_COUNTRY_KEY, show_col_types = FALSE)

oghist_raw <- suppressMessages(readxl::read_excel(
  PATH_OGHIST,
  sheet     = "Country Analytical History",
  col_names = FALSE
))

jmp_raw <- suppressMessages(readxl::read_excel(
  PATH_JMP,
  sheet     = "Sanitation",
  col_names = FALSE
))
```

------------------------------------------------------------------------

# 3. Survey selection

## Filter to study period (2015–2024)

Surveys outside the modelling window are dropped. Surveys near the boundary are flagged before filtering so the exclusions are visible.

```{r}
df_hh_sanitation <- df_hh_sanitation %>%
  dplyr::mutate(
    HH5Y_original = HH5Y,
    HH5Y          = standardise_survey_year(country, HH5Y)
  )

surveys_outside_range <- df_hh_sanitation %>%
  dplyr::filter(!is.na(HH5Y), HH5Y < 2015 | HH5Y > 2025) %>%
  dplyr::distinct(country, HH5Y_original, HH5Y) %>%
  dplyr::arrange(country, HH5Y)
print(surveys_outside_range)

countries_with_2024 <- df_hh_sanitation %>%
  dplyr::filter(HH5Y == 2024) %>%
  dplyr::distinct(country)

df_hh_sanitation <- df_hh_sanitation %>%
  dplyr::filter(
    is.na(HH5Y) | (HH5Y >= 2015 & HH5Y <= 2024) |
    (HH5Y == 2025 & country %in% countries_with_2024$country)
  )

surveys_removed <- df_hh_sanitation %>%
  dplyr::anti_join(
    df_hh_sanitation %>% dplyr::filter(
      is.na(HH5Y) | (HH5Y >= 2015 & HH5Y <= 2024) |
      (HH5Y == 2025 & country %in% countries_with_2024$country)
    ),
    by = names(df_hh_sanitation)
  )
```

## Select most recent data source per country

Where a country has both MICS and DHS data, we keep the source with the most recent survey year. Ties are broken by row count, then source name.

```{r}
country_source_summary <- df_hh_sanitation %>%
  dplyr::mutate(HH5Y = as.integer(HH5Y)) %>%
  dplyr::group_by(country, household_data_source) %>%
  dplyr::summarise(
    min_year     = ifelse(all(is.na(HH5Y)), NA_integer_, min(HH5Y, na.rm = TRUE)),
    max_year     = ifelse(all(is.na(HH5Y)), NA_integer_, max(HH5Y, na.rm = TRUE)),
    survey_years = paste(sort(unique(HH5Y[!is.na(HH5Y)])), collapse = ", "),
    n_rows       = dplyr::n(),
    .groups      = "drop"
  )

knitr::kable(
  country_source_summary %>% dplyr::arrange(country, household_data_source),
  caption = "Available surveys by country and data source"
)

source_to_keep <- country_source_summary %>%
  dplyr::group_by(country) %>%
  dplyr::arrange(desc(!is.na(max_year)), desc(max_year), desc(n_rows), household_data_source) %>%
  dplyr::slice(1) %>%
  dplyr::ungroup() %>%
  dplyr::select(country, household_data_source)

source_to_drop <- country_source_summary %>%
  dplyr::anti_join(source_to_keep, by = c("country", "household_data_source")) %>%
  dplyr::arrange(country, household_data_source)

knitr::kable(
  source_to_drop,
  caption = "Sources dropped (superseded by more recent data)"
)

df_hh_sanitation <- df_hh_sanitation %>%
  dplyr::semi_join(source_to_keep, by = c("country", "household_data_source"))

# Drop sub-national Pakistan MICS surveys — use national DHS instead
pakistan_mics_dropped <- df_hh_sanitation %>%
  dplyr::filter(stringr::str_detect(country, "Pakistan") &
                household_data_source == "MICS_sanitation") %>%
  dplyr::distinct(country, HH5Y, household_data_source)

knitr::kable(
  pakistan_mics_dropped,
  caption = "Pakistan sub-national MICS surveys dropped in favour of national DHS"
)

df_hh_sanitation <- df_hh_sanitation %>%
  dplyr::filter(!(stringr::str_detect(country, "Pakistan") &
                  household_data_source == "MICS_sanitation"))
```

------------------------------------------------------------------------

## Region level drops and merges

```{r}

# ============================================================
# Region-level drops and merges
# ============================================================

df_hh_sanitation <- df_hh_sanitation %>%
  # Drop uninformative region labels
    dplyr::filter(
    !(country == "Nauru"  & HH7_region == "LOCATION"),
    !(country == "Uganda" & HH7_region == "Refugee settlements")
  ) %>%
  dplyr::mutate(
    HH7_region = dplyr::case_when(

      # Togo: merge Golfe Urbain and Lomé Commune
      country == "Togo" & HH7_region %in% c("Golfe Urbain", "Lomé Commune") ~ "Lomé Commune/Golfe Urbain",

      country == "Sao Tome and Principe" & HH7_region %in% c(
  "DISTRITO DE ÁGUA GRANDE",
  "DISTRITO DE MÉ-ZÓCHI",
  "REGIÃO NORTE OESTE",
  "REGIÃO SUL ESTE"
) ~ "São Tomé",

country == "Sao Tome and Principe" & HH7_region == "REGIÃO AUTÓNOMA DO PRÍNCIPE" ~ "Príncipe",
      # Samoa: merge all Upolu regions
      country == "Samoa" & HH7_region %in% c(
        "Apia Urban Area",
        "North West Upolu",
        "Rest of Upolu"
      ) ~ "Upolu",

      # Samoa: keep Savaii as is
      country == "Samoa" & HH7_region == "Savaii" ~ "Savaii",

      TRUE ~ HH7_region
    )
  )
```

# 4. Geospatial covariates

## Build covariate table

Population sums and non-population covariates are merged into a single region-level covariate table. The `covariate_cols` vector is defined in `configuration/covariate_cols.R`.

```{r}
pop_files_other <- list.files(
  path       = path.expand(PATH_TO_POP_SUMS_OTHER),
  pattern    = "\\.csv$",
  full.names = TRUE
)

df_training_covariates_sanitation <- make_training_covariates(
  pop_files      = pop_files_other,
  nonpop_files   = FILE_NONPOP_OTHER,
  covariate_cols = covariate_cols
)
```

```{r}
# Load region area data and join to covariates
df_region_areas <- readr::read_csv(PATH_TO_REGION_AREAS) %>%
  dplyr::rename(HH7_region = HH7_region) %>%
  dplyr::mutate(HH7_region = stringr::str_trim(HH7_region)) %>%
  dplyr::select(country, analysis_year, HH7_region, area_km2) %>%
  dplyr::group_by(country, analysis_year, HH7_region) %>%
  dplyr::slice(1) %>%
  dplyr::ungroup()

df_training_covariates_sanitation <- df_training_covariates_sanitation %>%
  dplyr::left_join(
    df_region_areas,
    by = c("country", "analysis_year", "HH7_region" = "HH7_region")
  )

# Check how many rows are missing area after join
cat("Regions missing area_km2:", sum(is.na(df_training_covariates_sanitation$area_km2)), "\n")

```

## Check for duplicate covariate keys

Duplicate country–region–year keys would cause row inflation on joining and must be resolved before proceeding.

```{r}
duplicate_covariate_keys <- check_duplicate_covariate_keys(df_training_covariates_sanitation)
duplicate_covariate_keys
```

------------------------------------------------------------------------

## Restrict to most recent survey year per country

A single data source may contain multiple survey rounds. We keep only the most recent year per country before building outcomes, so that the year lookup and aggregation use only the retained survey.

```{r}
keep_latest_survey_year_hh <- function(df, max_survey_duration = 2) {
  df %>%
    dplyr::group_by(country) %>%
    dplyr::filter(HH5Y >= max(HH5Y, na.rm = TRUE) - max_survey_duration) %>%
    dplyr::ungroup()
}

df_hh_sanitation_latest <- keep_latest_survey_year_hh(df_hh_sanitation)

# Write the corrected household-level sanitation analysis file (v2). This
# incorporates the WS11 labelling corrections (via the _v2 MICS/DHS inputs) and
# has the same structure as the earlier df_hh_sanitation_analysis.csv — DHS rows
# retain NA in WS11_raw/WS11_improved/open_defecation/WS11_category, as before.
readr::write_csv(
  df_hh_sanitation_latest,
  PATH_HH_SANITATION_ANALYSIS_V2
)
message(paste("Written:", PATH_HH_SANITATION_ANALYSIS_V2))

# Writing df_hh_sanitation_latest to the switchdrive deliverable
# (df_hh_sanitation_analysis.csv) is commented out deliberately — this is a
# collaborator-facing deliverable (see additional_data_0002020937/HH_surveys/
# README.qmd) and should only be regenerated intentionally, not as a side
# effect of rerunning this script. Uncomment when you actually want to
# publish an updated version. Path defined in configuration/paths.R.
# dir.create(PATH_SWITCHDRIVE_HH_DIR, recursive = TRUE, showWarnings = FALSE)
# readr::write_csv(
#   df_hh_sanitation_latest,
#   PATH_HH_SANITATION_ANALYSIS
# )
# message(paste("Written:", PATH_HH_SANITATION_ANALYSIS))
```

## Build regional outcomes

Household-level responses are aggregated to region–year proportions for each outcome. Outcome definitions:

-   **Basic sanitation**: improved facility (`WS11 == 2`) AND not shared (`WS15 == 0`)
-   **Open defecation**: no facility / bush / field (`WS11 == 0`)

```{r}
basic_sanitation_outcome <- make_basic_sanitation_regional_outcome(df_hh_sanitation_latest)
open_defecation_outcome  <- make_open_defecation_regional_outcome(df_hh_sanitation_latest)
```

## Add PSU weights

Survey-design weights based on the number of primary sampling units are attached to each region–year row.

```{r}
sanitation_year_lookup <- make_year_lookup_from_raw_household_data(
  df_hh_sanitation_latest,
  manual_analysis_year_fixes = tibble::tribble(
    ~country,   ~analysis_year_manual,
    "Nigeria",  2024L
  )
)

basic_sanitation_outcome <- add_n_psu_weight(
  hh_df       = df_hh_sanitation,
  outcome_df  = basic_sanitation_outcome,
  year_lookup = sanitation_year_lookup
)
open_defecation_outcome <- add_n_psu_weight(
  hh_df       = df_hh_sanitation,
  outcome_df  = open_defecation_outcome,
  year_lookup = sanitation_year_lookup
)

# Quick sanity check on PSU coverage
dplyr::bind_rows(
  basic_sanitation_outcome %>% dplyr::mutate(outcome = "basic_sanitation"),
  open_defecation_outcome  %>% dplyr::mutate(outcome = "open_defecation")
) %>%
  dplyr::group_by(outcome) %>%
  dplyr::summarise(
    n_rows        = dplyr::n(),
    missing_n_psu = sum(is.na(n_psu)),
    min_n_psu     = min(n_psu, na.rm = TRUE),
    median_n_psu  = median(n_psu, na.rm = TRUE),
    max_n_psu     = max(n_psu, na.rm = TRUE),
    .groups = "drop"
  )
```

------------------------------------------------------------------------

# 5. Covariate–outcome matching

## Filter covariates to outcome country–years

The full covariate table is subset to only the country–year combinations present in each outcome, using a country name crosswalk to harmonise naming.

```{r}
df_covariates_basic_sanitation <- filter_covariates_to_outcome_years(
  outcome_df             = basic_sanitation_outcome,
  df_training_covariates = df_training_covariates_sanitation,
  country_name_key_WB    = country_name_key_WB
)
df_covariates_open_defecation <- filter_covariates_to_outcome_years(
  outcome_df             = open_defecation_outcome,
  df_training_covariates = df_training_covariates_sanitation,
  country_name_key_WB    = country_name_key_WB
)

check_duplicate_covariate_join_keys(df_covariates_basic_sanitation)
check_duplicate_covariate_join_keys(df_covariates_open_defecation)
```

## Diagnose unmatched country–years

Countries present in the outcome but absent from filtered covariates are flagged here. Common causes: country name mismatch or missing covariate data.

```{r}
country_year_filtering_issues <- dplyr::bind_rows(
  diagnose_country_year_filtering(
    outcome_name        = "basic_sanitation",
    outcome_df          = basic_sanitation_outcome,
    raw_covariates      = df_training_covariates_sanitation,
    filtered_covariates = df_covariates_basic_sanitation,
    country_name_key_WB = country_name_key_WB
  ),
  diagnose_country_year_filtering(
    outcome_name        = "open_defecation",
    outcome_df          = open_defecation_outcome,
    raw_covariates      = df_training_covariates_sanitation,
    filtered_covariates = df_covariates_open_defecation,
    country_name_key_WB = country_name_key_WB
  )
)
print(country_year_filtering_issues)
```

## Build region crosswalks

Survey region names are matched to covariate region names using fuzzy string matching. Rows that cannot be matched automatically are flagged for manual review.

```{r}
crosswalk_basic_sanitation <- build_crosswalk_for_one_outcome(
  outcome_name           = "basic_sanitation",
  outcome_df             = basic_sanitation_outcome,
  df_training_covariates = df_covariates_basic_sanitation,
  country_name_key_WB    = country_name_key_WB,
  output_dir             = CROSSWALK_REVIEW_DIR
)
crosswalk_open_defecation <- build_crosswalk_for_one_outcome(
  outcome_name           = "open_defecation",
  outcome_df             = open_defecation_outcome,
  df_training_covariates = df_covariates_open_defecation,
  country_name_key_WB    = country_name_key_WB,
  output_dir             = CROSSWALK_REVIEW_DIR
)

# Inspect rows that need manual attention
crosswalk_basic_sanitation$regions_to_review
crosswalk_open_defecation$regions_to_review
```

## Apply manual region name fixes

Known mismatches are resolved via the lookup table in `config/manual_region_fixes.R`. Add new entries there (not here) when new surveys are incorporated.

```{r}
basic_sanitation_crosswalk_final <- apply_manual_fixes_for_one_outcome(
  outcome_name        = "basic_sanitation",
  crosswalk_results   = crosswalk_basic_sanitation,
  manual_region_fixes = manual_fixes_sanitation,
  output_dir          = CROSSWALK_REVIEW_DIR
)
open_defecation_crosswalk_final <- apply_manual_fixes_for_one_outcome(
  outcome_name        = "open_defecation",
  crosswalk_results   = crosswalk_open_defecation,
  manual_region_fixes = manual_fixes_sanitation,
  output_dir          = CROSSWALK_REVIEW_DIR
)
```

## Join outcomes to covariates

```{r}
basic_sanitation_training_output <- join_and_save_one_training_dataset(
  outcome_name           = "basic_sanitation",
  outcome_df             = basic_sanitation_outcome,
  df_training_covariates = df_covariates_basic_sanitation,
  region_crosswalk_final = basic_sanitation_crosswalk_final,
  output_dir             = OUTPUT_DIR,
  training_csv_suffix    = "_v2"
)
open_defecation_training_output <- join_and_save_one_training_dataset(
  outcome_name           = "open_defecation",
  outcome_df             = open_defecation_outcome,
  df_training_covariates = df_covariates_open_defecation,
  region_crosswalk_final = open_defecation_crosswalk_final,
  output_dir             = OUTPUT_DIR,
  training_csv_suffix    = "_v2"
)

basic_sanitation_training <- basic_sanitation_training_output$training_with_covariates
open_defecation_training  <- open_defecation_training_output$training_with_covariates

basic_sanitation_training_output$missing_covariates
open_defecation_training_output$missing_covariates
```

```{r}



```

------------------------------------------------------------------------

# 6. Country-level covariates

## GDP per capita and secondary education

Country-year level economic and education indicators from the World Bank are joined by standardised country name. (Downloaded on 27 May 2026 from <https://databank.worldbank.org/source/world-development-indicators/Type/TABLE/preview/on>)

```{r}
gdp_clean <- readr::read_csv(PATH_GDP, show_col_types = FALSE) %>%
  dplyr::rename(
    country_name                     = Country.Name,
    analysis_year                    = Year,
    gdp_per_capita_constant_2015_usd = `GDP per capita (constant 2015 US$)`
  ) %>%
  dplyr::mutate(
    analysis_year                    = as.integer(analysis_year),
    gdp_per_capita_constant_2015_usd = as.numeric(gdp_per_capita_constant_2015_usd)
  ) %>%
  add_country_join_key_extended(
    country_col = country_name, country_name_key_WB = country_name_key_WB
  ) %>%
  dplyr::select(country_join, analysis_year, gdp_per_capita_constant_2015_usd) %>%
  dplyr::distinct()

education_clean <- readr::read_csv(PATH_EDUCATION, show_col_types = FALSE) %>%
  dplyr::rename(
    country_name                       = Country.Name,
    analysis_year                      = Year,
    secondary_education_duration_years = `Secondary education, duration (years)`
  ) %>%
  dplyr::mutate(
    analysis_year                      = as.integer(analysis_year),
    secondary_education_duration_years = as.numeric(secondary_education_duration_years)
  ) %>%
  add_country_join_key_extended(
    country_col = country_name, country_name_key_WB = country_name_key_WB
  ) %>%
  dplyr::select(country_join, analysis_year, secondary_education_duration_years) %>%
  dplyr::distinct()

country_year_covariates <- dplyr::full_join(
  gdp_clean, education_clean, by = c("country_join", "analysis_year")
)

add_country_year_covariates <- function(training_df) {
  training_df %>%
    dplyr::mutate(analysis_year = as.integer(analysis_year)) %>%
    add_country_join_key_extended(
      country_col = country_outcome, country_name_key_WB = country_name_key_WB
    ) %>%
    dplyr::left_join(country_year_covariates, by = c("country_join", "analysis_year")) %>%
    dplyr::select(-country_join)
}

basic_sanitation_training <- add_country_year_covariates(basic_sanitation_training)
open_defecation_training  <- add_country_year_covariates(open_defecation_training)

check_missing_covariates(basic_sanitation_training, "basic_sanitation",
  c("gdp_per_capita_constant_2015_usd", "secondary_education_duration_years"))
check_missing_covariates(open_defecation_training, "open_defecation",
  c("gdp_per_capita_constant_2015_usd", "secondary_education_duration_years"))
```

## Wastewater treatment

Country-level wastewater collection, treatment, and reuse percentages. These are static (no year dimension) and joined by country name only (Downloaded on 21 May 2026 from <https://doi.pangaea.de/10.1594/PANGAEA.918731>)

```{r}
ww_country_clean <- readr::read_tsv(PATH_WASTEWATER, show_col_types = FALSE) %>%
  dplyr::rename(
    ww_country_name            = Country,
    ww_region                  = Region,
    ww_economic_classification = Economic_Classification,
    ww_collection_percent      = WWc_Percent,
    ww_treatment_percent       = WWt_Percent,
    ww_reuse_percent           = WWr_Percent
  ) %>%
  dplyr::mutate(across(ends_with("_percent"), as.numeric)) %>%
  add_country_join_key_extended(
    country_col = ww_country_name, country_name_key_WB = country_name_key_WB
  ) %>%
  dplyr::distinct(country_join, .keep_all = TRUE)

add_ww_covariates <- function(training_df) {
  training_df %>%
    dplyr::select(-dplyr::any_of(c(
      "country_join", "ww_country_name", "ww_region",
      "ww_economic_classification",
      "ww_collection_percent", "ww_treatment_percent", "ww_reuse_percent"
    ))) %>%
    add_country_join_key_extended(
      country_col = country_outcome, country_name_key_WB = country_name_key_WB
    ) %>%
    dplyr::left_join(
      ww_country_clean %>% dplyr::select(
        country_join, ww_region, ww_economic_classification,
        ww_collection_percent, ww_treatment_percent, ww_reuse_percent
      ),
      by = "country_join"
    ) %>%
    dplyr::select(-country_join)
}

basic_sanitation_training <- add_ww_covariates(basic_sanitation_training)
open_defecation_training  <- add_ww_covariates(open_defecation_training)

# Kosovo has no wastewater record; assign region manually
basic_sanitation_training <- basic_sanitation_training %>%
  dplyr::mutate(ww_region = dplyr::if_else(
    country_outcome == "Kosovo" & is.na(ww_region),
    "Eastern Europe & Central Asia", ww_region
  ))
open_defecation_training <- open_defecation_training %>%
  dplyr::mutate(ww_region = dplyr::if_else(
    country_outcome == "Kosovo" & is.na(ww_region),
    "Eastern Europe & Central Asia", ww_region
  ))

check_missing_covariates(basic_sanitation_training, "basic_sanitation",
  c("ww_collection_percent", "ww_treatment_percent", "ww_reuse_percent"))
check_missing_covariates(open_defecation_training, "open_defecation",
  c("ww_collection_percent", "ww_treatment_percent", "ww_reuse_percent"))
```

## WGI governance indicators

Six World Governance Indicators are joined by country and year. (Downloaded on 27 May 2026 from <https://databank.worldbank.org/source/worldwide-governance-indicators/Type/TABLE/preview/on>)

```{r}
indicator_map <- c(
  "Government Effectiveness - Governance score (0-100)" = "governance_effectiveness",
  "Control of Corruption - Governance score (0-100)"    = "control_of_corruption",
  "Political Stability - Governance score (0-100)"      = "political_stability",
  "Regulatory Quality - Governance score (0-100)"       = "regulatory_quality",
  "Rule of Law - Governance score (0-100)"              = "rule_of_law",
  "Voice and Accountability - Governance score (0-100)" = "voice_and_accountability"
)

wgi_clean <- df_wgi_lmics %>%
  dplyr::filter(`Series.Name` %in% names(indicator_map)) %>%
  dplyr::mutate(indicator = unname(indicator_map[`Series.Name`])) %>%
  tidyr::pivot_longer(
    cols      = dplyr::matches("^X20[0-9]{2}$"),
    names_to  = "analysis_year",
    values_to = "value"
  ) %>%
  dplyr::mutate(
    analysis_year = as.integer(stringr::str_remove(analysis_year, "^X")),
    value         = as.numeric(value)
  ) %>%
  add_country_join_key_extended(
    country_col = `Country.Name`, country_name_key_WB = country_name_key_WB
  ) %>%
  dplyr::select(country_join, analysis_year, indicator, value) %>%
  tidyr::pivot_wider(names_from = indicator, values_from = value) %>%
  dplyr::distinct()

add_wgi_covariates <- function(training_df) {
  training_df %>%
    dplyr::select(-dplyr::any_of(c("country_join", unname(indicator_map)))) %>%
    dplyr::mutate(analysis_year = as.integer(analysis_year)) %>%
    add_country_join_key_extended(
      country_col = country_outcome, country_name_key_WB = country_name_key_WB
    ) %>%
    dplyr::left_join(wgi_clean, by = c("country_join", "analysis_year")) %>%
    dplyr::select(-country_join)
}

basic_sanitation_training <- add_wgi_covariates(basic_sanitation_training)
open_defecation_training  <- add_wgi_covariates(open_defecation_training)

check_missing_covariates(basic_sanitation_training, "basic_sanitation", unname(indicator_map))
check_missing_covariates(open_defecation_training,  "open_defecation",  unname(indicator_map))
```

------------------------------------------------------------------------

## World Bank Income Groups

Here we use the file OGHIST_2026_03_10.xlsx which was downloaded from the World Bank website on 13 June 2026 (<https://datahelpdesk.worldbank.org/knowledgebase/articles/906519-world-bank-country-and-lending-groups>)

```{r}
# ============================================================
# World Bank income group classification (OGHIST)
# ============================================================

# Assign column names: cols 1-2 get fixed names, cols 3-40 get the calendar
# years from row 6 (coerced via numeric to handle Excel's numeric cell format).
col_names_vec <- c(
  "country_code", "country_name",
  as.character(suppressWarnings(as.integer(as.numeric(unlist(oghist_raw[6, 3:40])))))
)

# Country data starts at row 12; rows above are header / threshold definitions.
oghist_countries <- oghist_raw[12:nrow(oghist_raw), 1:40]
names(oghist_countries) <- col_names_vec

target_years <- as.character(2015:2024)

income_group_clean <- oghist_countries %>%
  dplyr::select(country_code, country_name, dplyr::all_of(target_years)) %>%
  dplyr::filter(!is.na(country_code), !is.na(country_name)) %>%
  tidyr::pivot_longer(
    cols      = dplyr::all_of(target_years),
    names_to  = "analysis_year",
    values_to = "wb_income_group"
  ) %>%
  dplyr::mutate(
    analysis_year   = as.integer(analysis_year),
    wb_income_group = dplyr::if_else(wb_income_group == "..", NA_character_, wb_income_group)
  )

# Join income group into training datasets via country_join key.
# wb_income_group values: "L", "LM", "UM", "H"
add_income_group_covariates <- function(training_df) {
  training_df %>%
    dplyr::select(-dplyr::any_of(c("country_join", "wb_income_group"))) %>%
    dplyr::mutate(analysis_year = as.integer(analysis_year)) %>%
    add_country_join_key_extended(
      country_col         = country_outcome,
      country_name_key_WB = country_name_key_WB
    ) %>%
    dplyr::left_join(
      income_group_clean %>%
        add_country_join_key_extended(
          country_col         = country_name,
          country_name_key_WB = country_name_key_WB
        ) %>%
        dplyr::select(country_join, analysis_year, wb_income_group),
      by = c("country_join", "analysis_year")
    ) %>%
    dplyr::select(-country_join)
}

basic_sanitation_training <- add_income_group_covariates(basic_sanitation_training)
open_defecation_training  <- add_income_group_covariates(open_defecation_training)

# Diagnostics: rows missing income group
check_missing_income_group <- function(training_df, outcome_name) {
  training_df %>%
    dplyr::filter(is.na(wb_income_group)) %>%
    dplyr::distinct(country_outcome, analysis_year) %>%
    dplyr::mutate(outcome_type = outcome_name) %>%
    dplyr::arrange(country_outcome, analysis_year)
}

missing_income_group <- dplyr::bind_rows(
  check_missing_income_group(basic_sanitation_training, "basic_sanitation"),
  check_missing_income_group(open_defecation_training,  "open_defecation")
)
print(missing_income_group)
```

## JMP Sanitation Indicators

Extracts country-level sanitation estimates (at least basic, open defecation), regional classifications (ISO3, SDG region, fragile context) from the JMP 2025 global dataset (downloaded on 13 June 2026 from <https://washdata.org/data/downloads>) and joins them to the training data by country and year. Time-invariant fields (ISO3, SDG region) are carried forward from any available survey year to fill gaps where a specific year is missing.

```{r}
# ============================================================
# JMP sanitation indicators
# ============================================================

jmp_cols <- c(
  country_name     = 1,
  analysis_year    = 2,
  sanitation_basic = 17,   # "At least basic" (TOTAL)
  open_defecation  = 20,   # "Open defecation" (TOTAL)
  iso3             = 52,   # ISO3
  sdg_region       = 53,   # SDG Region
  fragile_context  = 59    # Fragile contexts
)

jmp_clean <- jmp_raw[4:nrow(jmp_raw), unname(jmp_cols)] %>%
  purrr::set_names(names(jmp_cols)) %>%
  dplyr::filter(!is.na(country_name), !is.na(analysis_year)) %>%
  dplyr::mutate(
    analysis_year = as.integer(analysis_year),
    dplyr::across(
      c(sanitation_basic, open_defecation),
      ~ dplyr::case_when(
          stringr::str_starts(.x, "<") ~ 0,
          stringr::str_starts(.x, ">") ~ 100,
          TRUE                         ~ suppressWarnings(as.numeric(.x))
        )
    )
  )

# Country-level lookup for time-invariant fields: take the first non-NA value
# across all survey years so missing-year joins still get iso3/sdg_region/fragile_context
jmp_country_meta <- jmp_clean %>%
  dplyr::group_by(country_name) %>%
  dplyr::summarise(
    iso3            = dplyr::first(iso3[!is.na(iso3)]),
    sdg_region      = dplyr::first(sdg_region[!is.na(sdg_region)]),
    fragile_context = dplyr::first(fragile_context[!is.na(fragile_context)]),
    .groups = "drop"
  ) %>%
  add_country_join_key_extended(
    country_col         = country_name,
    country_name_key_WB = country_name_key_WB
  ) %>%
  dplyr::select(country_join, iso3, sdg_region, fragile_context)

# Year-varying estimates only
jmp_estimates <- jmp_clean %>%
  dplyr::select(country_name, analysis_year, sanitation_basic, open_defecation) %>%
  add_country_join_key_extended(
    country_col         = country_name,
    country_name_key_WB = country_name_key_WB
  ) %>%
  dplyr::select(country_join, analysis_year, sanitation_basic, open_defecation)

# Join JMP covariates into training datasets
add_jmp_covariates <- function(training_df) {
  training_df %>%
    dplyr::select(-dplyr::any_of(c(
      "country_join", "sanitation_basic", "open_defecation",
      "iso3", "sdg_region", "fragile_context"
    ))) %>%
    dplyr::mutate(analysis_year = as.integer(analysis_year)) %>%
    add_country_join_key_extended(
      country_col         = country_outcome,
      country_name_key_WB = country_name_key_WB
    ) %>%
    dplyr::left_join(jmp_estimates, by = c("country_join", "analysis_year")) %>%
    dplyr::left_join(jmp_country_meta, by = "country_join") %>%
    dplyr::select(-country_join)
}

basic_sanitation_training <- add_jmp_covariates(basic_sanitation_training)
open_defecation_training  <- add_jmp_covariates(open_defecation_training)

# Manual patch for countries whose names don't match JMP — applied directly
# on country_outcome to bypass country_join key mismatches
patch_jmp_meta <- function(training_df) {
  training_df %>%
    dplyr::mutate(
      iso3 = dplyr::case_when(
        country_outcome == "Kosovo"    & is.na(iso3) ~ "XKX",
        country_outcome == "Palestina" & is.na(iso3) ~ "PSE",
        country_outcome == "Tanzania"  & is.na(iso3) ~ "TZA",
        TRUE                                         ~ iso3
      ),
      sdg_region = dplyr::case_when(
        country_outcome == "Kosovo"    & is.na(sdg_region) ~ "Europe and Northern America",
        country_outcome == "Palestina" & is.na(sdg_region) ~ "Northern Africa and Western Asia",
        country_outcome == "Tanzania"  & is.na(sdg_region) ~ "Sub-Saharan Africa",
        TRUE                                               ~ sdg_region
      )
    )
}

basic_sanitation_training <- patch_jmp_meta(basic_sanitation_training)
open_defecation_training  <- patch_jmp_meta(open_defecation_training)

# Diagnostics: rows still missing JMP covariates after meta fallback and patch
jmp_vars <- c("sanitation_basic", "open_defecation", "iso3", "sdg_region")

check_missing_jmp <- function(training_df, outcome_name) {
  training_df %>%
    dplyr::filter(dplyr::if_any(dplyr::all_of(jmp_vars), is.na)) %>%
    dplyr::distinct(country_outcome, analysis_year,
                    dplyr::across(dplyr::all_of(jmp_vars))) %>%
    dplyr::mutate(outcome_type = outcome_name) %>%
    dplyr::arrange(country_outcome, analysis_year)
}

missing_jmp <- dplyr::bind_rows(
  check_missing_jmp(basic_sanitation_training, "basic_sanitation")
)
print(missing_jmp)
```

# 7. Output

## Write training CSVs

`join_and_save_one_training_dataset()` (Section 5) writes these files with the **geospatial** covariates only. The **country-level** covariates (income group, SDG region, GDP, education, wastewater, WGI, JMP) are added afterwards in Section 6, so we re-write the files here with the complete tables the models actually train on. Same `_v2` filenames, so this overwrites the partial versions.

```{r}
dir.create(OUTPUT_DIR,           recursive = TRUE, showWarnings = FALSE)
dir.create(CROSSWALK_REVIEW_DIR, recursive = TRUE, showWarnings = FALSE)

readr::write_csv(
  basic_sanitation_training,
  file.path(OUTPUT_DIR, "basic_sanitation_training_with_covariates_v2.csv")
)
readr::write_csv(
  open_defecation_training,
  file.path(OUTPUT_DIR, "open_defecation_training_with_covariates_v2.csv")
)
message("Written: basic_sanitation_training_with_covariates_v2.csv (with country-level covariates)")
message("Written: open_defecation_training_with_covariates_v2.csv (with country-level covariates)")
```

## Summary: wastewater region distribution

A quick check on regional representation in the final training data.

```{r}
dplyr::bind_rows(
  basic_sanitation_training %>% dplyr::mutate(outcome_type = "basic_sanitation"),
  open_defecation_training  %>% dplyr::mutate(outcome_type = "open_defecation")
) %>%
  dplyr::count(outcome_type, ww_region, name = "n_observations") %>%
  dplyr::group_by(outcome_type) %>%
  dplyr::mutate(percent = round(100 * n_observations / sum(n_observations), 2)) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(outcome_type, desc(n_observations))
```

------------------------------------------------------------------------

# 8. Survey data availability overview

Bar charts showing the number of surveys (distinct country × year combinations) included in the training data, broken down by SDG region and World Bank income group. Three survey groups are compared:

1.  **All MICS surveys** – the full MICS extract before source selection.
2.  **Basic sanitation modelling surveys** – the final `basic_sanitation_training` dataset, excluding Indonesia.

## Inspect available columns

```{r}
cat("── basic_sanitation_training columns ──\n")
print(names(basic_sanitation_training))

cat("\n── df_mics columns ──\n")
print(names(df_mics))
```

## Build survey-count datasets

```{r}
# ── 1. MICS surveys included in basic sanitation training ────────────────────
# source_to_keep (built in section 3) maps country -> household_data_source
mics_countries <- source_to_keep %>%
  dplyr::filter(household_data_source == "MICS_sanitation") %>%
  dplyr::pull(country)

surveys_mics_sdg <- basic_sanitation_training %>%
  dplyr::filter(country_outcome %in% mics_countries) %>%
  dplyr::distinct(country_outcome, HH7_region_outcome, sdg_region) %>%
  dplyr::count(sdg_region, name = "n_surveys") %>%
  dplyr::mutate(group = "MICS surveys in training data")

surveys_mics_income <- basic_sanitation_training %>%
  dplyr::filter(country_outcome %in% mics_countries) %>%
  dplyr::distinct(country_outcome, wb_income_group) %>%
  dplyr::count(wb_income_group, name = "n_surveys") %>%
  dplyr::mutate(group = "MICS surveys in training data")

# ── 2. Basic sanitation modelling surveys (no Indonesia) ─────────────────────
basic_san_filtered <- basic_sanitation_training %>%
  dplyr::filter(country_outcome != "Indonesia")

surveys_basic_san_sdg <- basic_san_filtered %>%
  dplyr::distinct(country_outcome, HH7_region_outcome, sdg_region) %>%
  dplyr::count(sdg_region, name = "n_surveys") %>%
  dplyr::mutate(group = "Basic sanitation (no Indonesia)")

surveys_basic_san_income <- basic_san_filtered %>%
  dplyr::distinct(country_outcome, wb_income_group) %>%
  dplyr::count(wb_income_group, name = "n_surveys") %>%
  dplyr::mutate(group = "Basic sanitation (no Indonesia)")
```

## Bar charts by SDG region

```{r fig.width=10, fig.height=6}
# SDG region order: descending by total surveys across all groups
sdg_order <- dplyr::bind_rows(surveys_mics_sdg, surveys_basic_san_sdg) %>%
  dplyr::group_by(sdg_region) %>%
  dplyr::summarise(total = sum(n_surveys), .groups = "drop") %>%
  dplyr::arrange(desc(total)) %>%
  dplyr::pull(sdg_region)

sdg_plot_data <- dplyr::bind_rows(surveys_mics_sdg, surveys_basic_san_sdg) %>%
  dplyr::filter(!is.na(sdg_region)) %>%
  dplyr::mutate(
    sdg_region = factor(sdg_region, levels = rev(sdg_order)),
    group      = factor(group, levels = c("MICS surveys in training data",
                                          "Basic sanitation (no Indonesia)"))
  )

ggplot2::ggplot(sdg_plot_data,
                ggplot2::aes(x = sdg_region, y = n_surveys, fill = group)) +
  ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.7), width = 0.65) +
  ggplot2::coord_flip() +
  ggplot2::scale_fill_manual(
    values = c("MICS surveys in training data"     = "#4C9BE8",
               "Basic sanitation (no Indonesia)"   = "#E88B4C")
  ) +
  ggplot2::labs(
    title = NULL,
    x     = NULL,
    y     = "Number of regions",
    fill  = NULL
  ) +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(legend.position = "bottom")
```

## Survey coverage summary

```{r}
dplyr::bind_rows(
  basic_sanitation_training %>%
    dplyr::filter(country_outcome %in% mics_countries) %>%
    dplyr::summarise(
      group       = "MICS surveys in training data",
      n_countries = dplyr::n_distinct(country_outcome),
      n_regions   = dplyr::n_distinct(paste(country_outcome, HH7_region_outcome))
    ),
  basic_sanitation_training %>%
    dplyr::filter(country_outcome != "Indonesia") %>%
    dplyr::summarise(
      group       = "Basic sanitation (no Indonesia)",
      n_countries = dplyr::n_distinct(country_outcome),
      n_regions   = dplyr::n_distinct(paste(country_outcome, HH7_region_outcome))
    )
)
```

------------------------------------------------------------------------

# 9. Descriptive statistics table

Country-level summary of survey data and all covariates used in the sanitation model. This is deliberately the last step in this script, so that it reflects the training data actually written out in Section 5-7 (after the outcome/weight/duplicate cleaning in `join_and_save_one_training_dataset()`), not a separately-recomputed household-level view. Written to `outputs/descriptive_statistics/`.

```{r}
# ── Recap of the cleaning applied when the training data was built ──────────
# (join_and_save_one_training_dataset() already printed this when it ran in
# Section 5 — repeated here since this is the step meant to summarise what
# actually went into the training data.)
cat("=== Recap: cleaning summary from training data build ===\n")
print(dplyr::bind_rows(
  basic_sanitation_training_output$cleaning_summary,
  open_defecation_training_output$cleaning_summary
))

# ── Region-level training row counts per country (post-cleaning) ────────────
# Counts rows in the FINAL basic_sanitation_training / open_defecation_training
# objects — i.e. after missing-outcome, missing-weight and duplicate rows have
# already been removed. This is what the models actually train on.
n_regions_basic_sanitation <- basic_sanitation_training %>%
  dplyr::count(country_outcome, name = "n_regions_basic_sanitation_training")

n_regions_open_defecation <- open_defecation_training %>%
  dplyr::count(country_outcome, name = "n_regions_open_defecation_training")

# ── National sanitation estimates from raw survey data ───────────────────────
national_sanitation <- createNationalSanitationEstimates(df_hh_sanitation_latest)

cat("Total households with basic sanitation data: ",
    sum(national_sanitation$n_hh_basic_sanitation), "\n")
cat("Total households with open defecation data:  ",
    sum(national_sanitation$n_hh_open_defecation), "\n")

# ── Survey date range per country ────────────────────────────────────────────
survey_dates <- df_hh_sanitation_latest %>%
  harmonise_household_country_names_for_modelling() %>%
  dplyr::filter(!is.na(HH5Y)) %>%
  dplyr::group_by(country) %>%
  dplyr::summarise(
    start_data_collection = min(
      lubridate::make_date(HH5Y, dplyr::if_else(is.na(HH5M), 1L, as.integer(HH5M)), 1L),
      na.rm = TRUE
    ),
    end_data_collection = max(
      lubridate::make_date(HH5Y, dplyr::if_else(is.na(HH5M), 12L, as.integer(HH5M)), 1L),
      na.rm = TRUE
    ),
    .groups = "drop"
  )

# ── Country-level covariates (one row per country) ───────────────────────────
# Collapse from region-level training data; time-varying covariates are taken
# at the analysis year already stored in the training data.
# Source from BOTH outcomes so that a country dropped from basic sanitation
# (e.g. Indonesia, no WS15 data) still gets its metadata and shared covariates
# from the open defecation training data — only its basic-sanitation-specific
# outcome columns should end up NA, not analysis_year / region / income group.
country_covariates <- dplyr::bind_rows(
    basic_sanitation_training,
    open_defecation_training
  ) %>%
  dplyr::group_by(country_outcome, analysis_year) %>%
  dplyr::summarise(
    gdp_per_capita_constant_2015_usd   = dplyr::first(na.omit(gdp_per_capita_constant_2015_usd)),
    secondary_education_duration_years = dplyr::first(na.omit(secondary_education_duration_years)),
    ww_collection_percent              = dplyr::first(na.omit(ww_collection_percent)),
    ww_treatment_percent               = dplyr::first(na.omit(ww_treatment_percent)),
    ww_reuse_percent                   = dplyr::first(na.omit(ww_reuse_percent)),
    governance_effectiveness           = dplyr::first(na.omit(governance_effectiveness)),
    control_of_corruption              = dplyr::first(na.omit(control_of_corruption)),
    political_stability                = dplyr::first(na.omit(political_stability)),
    regulatory_quality                 = dplyr::first(na.omit(regulatory_quality)),
    rule_of_law                        = dplyr::first(na.omit(rule_of_law)),
    voice_and_accountability           = dplyr::first(na.omit(voice_and_accountability)),
    wb_income_group                    = dplyr::first(na.omit(wb_income_group)),
    sdg_region                         = dplyr::first(na.omit(sdg_region)),
    iso3                               = dplyr::first(na.omit(iso3)),
    jmp_sanitation_basic               = dplyr::first(na.omit(sanitation_basic)),
    jmp_open_defecation                = dplyr::first(na.omit(open_defecation)),
    .groups = "drop"
  ) %>%
  dplyr::rename(country = country_outcome)

# ── Within-country range of raw subnational estimates ─────────────────────────
# Spread (max - min, in percentage points) of admin-1 outcome values within each
# country, after averaging each admin-1 unit across survey years. NA where a
# country has only one surveyed admin-1 unit (no range to speak of).
within_country_range <- function(df) {
  df %>%
    dplyr::filter(!is.na(country_outcome), !is.na(HH7_region_outcome),
                  !is.na(outcome_value)) %>%
    dplyr::group_by(country_outcome, HH7_region_outcome) %>%
    dplyr::summarise(value = mean(outcome_value, na.rm = TRUE), .groups = "drop") %>%
    dplyr::group_by(country_outcome) %>%
    dplyr::summarise(
      range = dplyr::if_else(dplyr::n() >= 2,
                             100 * (max(value) - min(value)), NA_real_),
      .groups = "drop"
    )
}

range_basic_sanitation <- within_country_range(basic_sanitation_training) %>%
  dplyr::rename(country = country_outcome, range_basic_sanitation_pp = range)
range_open_defecation <- within_country_range(open_defecation_training) %>%
  dplyr::rename(country = country_outcome, range_open_defecation_pp = range)

# ── Assemble final table ──────────────────────────────────────────────────────
descriptive_stats <- survey_dates %>%
  dplyr::left_join(country_covariates,   by = "country") %>%
  dplyr::left_join(national_sanitation,  by = "country") %>%
  dplyr::left_join(n_regions_basic_sanitation, by = c("country" = "country_outcome")) %>%
  dplyr::left_join(n_regions_open_defecation,  by = c("country" = "country_outcome")) %>%
  dplyr::left_join(range_basic_sanitation, by = "country") %>%
  dplyr::left_join(range_open_defecation,  by = "country") %>%
  dplyr::mutate(
    n_regions_basic_sanitation_training = tidyr::replace_na(n_regions_basic_sanitation_training, 0L),
    n_regions_open_defecation_training  = tidyr::replace_na(n_regions_open_defecation_training, 0L)
  ) %>%
  dplyr::select(
    country,
    start_data_collection,
    end_data_collection,
    analysis_year,
    wb_income_group,
    sdg_region,
    iso3,
    prop_basic_sanitation_national,
    n_hh_basic_sanitation,
    pct_pop_known_basic_sanitation,
    n_regions_basic_sanitation_training,
    range_basic_sanitation_pp,
    prop_open_defecation_national,
    n_hh_open_defecation,
    n_regions_open_defecation_training,
    range_open_defecation_pp,
    ww_collection_percent,
    ww_treatment_percent,
    ww_reuse_percent,
    gdp_per_capita_constant_2015_usd,
    jmp_sanitation_basic,
    jmp_open_defecation,
    secondary_education_duration_years,
    governance_effectiveness,
    control_of_corruption,
    political_stability,
    regulatory_quality,
    rule_of_law,
    voice_and_accountability
  ) %>%
  dplyr::arrange(sdg_region, country)

# ── Write output ──────────────────────────────────────────────────────────────
DESCRIPTIVE_STATS_DIR <- here::here("outputs/descriptive_statistics")
dir.create(DESCRIPTIVE_STATS_DIR, recursive = TRUE, showWarnings = FALSE)

readr::write_csv(
  descriptive_stats,
  file.path(DESCRIPTIVE_STATS_DIR, "sanitation_survey_descriptive_statistics_v2.csv")
)
message("Written: sanitation_survey_descriptive_statistics_v2.csv")

knitr::kable(descriptive_stats, caption = "Sanitation survey data and country-level covariates")
```

------------------------------------------------------------------------

```{r}

descriptive_stats %>%
  tidyr::pivot_longer(c(range_basic_sanitation_pp, range_open_defecation_pp),
                      names_to = "outcome", values_to = "range_pp") %>%
  group_by(outcome) %>%
  summarise(
    n_countries  = sum(!is.na(range_pp)),
    median_range = median(range_pp, na.rm = TRUE),
    min_range    = min(range_pp,    na.rm = TRUE),
    max_range    = max(range_pp,    na.rm = TRUE),
    .groups = "drop"
  )
```

# 10. Surveys diverging from JMP by more than 5 percentage points

Flags surveys where the national estimate computed here differs from the JMP country estimate by more than 5 percentage points, separately for basic sanitation and open defecation. My `prop_*` estimates are proportions (0–1); JMP values (`jmp_*`) are percentages (0–100), so mine are converted to percentages and the difference is expressed in percentage points (`my_estimate − jmp_estimate`). Countries where my estimate is NA (e.g. dropped by the WS15 coverage rule) cannot be compared and are excluded.

```{r}
# Survey type (MICS / DHS) per country, from the retained household data.
# Uses the same country-name harmonisation as descriptive_stats so the join
# lines up.
survey_type_lookup <- df_hh_sanitation_latest %>%
  harmonise_household_country_names_for_modelling() %>%
  dplyr::mutate(
    survey_type = dplyr::case_when(
      stringr::str_detect(household_data_source, "MICS") ~ "MICS",
      stringr::str_detect(household_data_source, "DHS")  ~ "DHS",
      TRUE                                               ~ household_data_source
    )
  ) %>%
  dplyr::distinct(country, survey_type)

jmp_comparison <- descriptive_stats %>%
  dplyr::left_join(survey_type_lookup, by = "country") %>%
  dplyr::transmute(
    country,
    survey_type,
    analysis_year,
    my_basic_sanitation_pct  = 100 * prop_basic_sanitation_national,
    jmp_basic_sanitation_pct = jmp_sanitation_basic,
    diff_basic_sanitation    = 100 * prop_basic_sanitation_national - jmp_sanitation_basic,
    my_open_defecation_pct   = 100 * prop_open_defecation_national,
    jmp_open_defecation_pct  = jmp_open_defecation,
    diff_open_defecation     = 100 * prop_open_defecation_national - jmp_open_defecation
  )

# ── Basic sanitation: |difference| > 5 percentage points ─────────────────────
basic_sanitation_jmp_gt5 <- jmp_comparison %>%
  dplyr::filter(!is.na(diff_basic_sanitation), abs(diff_basic_sanitation) > 5) %>%
  dplyr::select(country, survey_type, analysis_year,
                my_basic_sanitation_pct, jmp_basic_sanitation_pct,
                diff_basic_sanitation) %>%
  dplyr::arrange(dplyr::desc(abs(diff_basic_sanitation)))

# ── Open defecation: |difference| > 5 percentage points ──────────────────────
open_defecation_jmp_gt5 <- jmp_comparison %>%
  dplyr::filter(!is.na(diff_open_defecation), abs(diff_open_defecation) > 5) %>%
  dplyr::select(country, survey_type, analysis_year,
                my_open_defecation_pct, jmp_open_defecation_pct,
                diff_open_defecation) %>%
  dplyr::arrange(dplyr::desc(abs(diff_open_defecation)))

cat("\n=== Basic sanitation: |my estimate - JMP| > 5 percentage points (",
    nrow(basic_sanitation_jmp_gt5), "surveys ) ===\n")
print(as.data.frame(basic_sanitation_jmp_gt5), row.names = FALSE)

cat("\n=== Open defecation: |my estimate - JMP| > 5 percentage points (",
    nrow(open_defecation_jmp_gt5), "surveys ) ===\n")
print(as.data.frame(open_defecation_jmp_gt5), row.names = FALSE)
```

------------------------------------------------------------------------

# 11. Within- vs between-country variance decomposition

Quantifies how much of the total variance in each outcome is attributable to differences *between* countries versus variation *within* countries across subnational regions. A high ICC (intraclass correlation coefficient) means most variance is between countries — national-level predictors are likely to dominate. A lower ICC means subnational predictors carry relatively more weight. This directly contextualises the SHAP feature importance results in the modelling script.

Survey weights (`n_psu_weight_scaled`) are incorporated via a weighted mixed-effects model so that larger surveys contribute proportionally.

```{r}
library(lme4)

# ── Load the final training CSVs written in Section 7 ────────────────────────
bs_train <- readr::read_csv(
  here::here("data/processed/training_subcomponents/basic_sanitation_training_with_covariates_v2.csv"),
  show_col_types = FALSE
) |>
  dplyr::filter(!is.na(outcome_value), !is.na(country_cov), !is.na(n_psu_weight_scaled))

od_train <- readr::read_csv(
  here::here("data/processed/training_subcomponents/open_defecation_training_with_covariates_v2.csv"),
  show_col_types = FALSE
) |>
  dplyr::filter(!is.na(outcome_value), !is.na(country_cov), !is.na(n_psu_weight_scaled))

# ── ICC helper ────────────────────────────────────────────────────────────────
# Fits a null (intercept-only) mixed-effects model with country as the random
# grouping factor, then extracts the proportion of variance attributable to
# between-country differences.
compute_icc <- function(df, outcome_col = "outcome_value",
                        country_col = "country_cov",
                        weight_col  = "n_psu_weight_scaled") {

  # Normalise weights to mean 1 (lmer requires positive weights, not probabilities)
  df <- df |>
    dplyr::mutate(.w = .data[[weight_col]] / mean(.data[[weight_col]], na.rm = TRUE))

  formula <- as.formula(
    paste0(outcome_col, " ~ 1 + (1 | ", country_col, ")")
  )

  fit <- lme4::lmer(formula, data = df, weights = df$.w,
                    control = lme4::lmerControl(optimizer = "bobyqa"))

  vc            <- as.data.frame(lme4::VarCorr(fit))
  var_between   <- vc$vcov[vc$grp == country_col]
  var_within    <- vc$vcov[vc$grp == "Residual"]
  var_total     <- var_between + var_within
  icc           <- var_between / var_total

  tibble::tibble(
    var_between  = var_between,
    var_within   = var_within,
    var_total    = var_total,
    icc          = icc,
    pct_between  = 100 * icc,
    pct_within   = 100 * (1 - icc),
  )
}

# ── Run for both outcomes ─────────────────────────────────────────────────────
icc_bs <- compute_icc(bs_train) |> dplyr::mutate(outcome = "basic_sanitation", .before = 1)
icc_od <- compute_icc(od_train) |> dplyr::mutate(outcome = "open_defecation",  .before = 1)

icc_results <- dplyr::bind_rows(icc_bs, icc_od)

# ── Print ─────────────────────────────────────────────────────────────────────
cat("\nWithin- vs between-country variance decomposition\n")
cat("──────────────────────────────────────────────────\n")
print(
  icc_results |>
    dplyr::select(outcome, pct_between, pct_within, icc) |>
    dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 3))),
  n = Inf
)

# ── Save ──────────────────────────────────────────────────────────────────────
DESCRIPTIVE_STATS_DIR <- here::here("outputs/descriptive_statistics")
dir.create(DESCRIPTIVE_STATS_DIR, recursive = TRUE, showWarnings = FALSE)

readr::write_csv(
  icc_results,
  file.path(DESCRIPTIVE_STATS_DIR, "sanitation_icc_variance_decomposition.csv")
)
message("Written: sanitation_icc_variance_decomposition.csv")

knitr::kable(
  icc_results |> dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 3))),
  caption = "Within- vs between-country variance decomposition (weighted ICC)"
)
```
