---
title: "Prepare SMDW Model Training Data"
subtitle: "Safe drinking water and subcomponent outcomes with geospatial covariates"
---

# 1. Setup

## Libraries and sourced functions

```{r}
library(tidyverse)
library(here)
library(fuzzyjoin)
library(stringdist)
library(stringi)

source(here::here("functions/create_SMDW_indicators.R"))
source(here::here("functions/join_and_structure_dataframes.R"))
source(here::here("functions/diagnostics.R"))           # generalised diagnostic helpers
source(here::here("functions/country_join_helpers.R"))  # unified add_country_join_key_extended()
source(here::here("configuration/covariate_cols.R"))           # loads `covariate_cols` vector
source(here::here("configuration/manual_region_fixes.R"))      # loads `manual_fixes_smdw`, `manual_fixes_smdw_ecoli`
source(here::here("configuration/paths.R"))                         # loads switchdrive paths
```

## Paths

All raw data paths are defined here so they are easy to update when data
moves.

```{r}
# ── Geospatial covariate paths ──────────────────────────────────────────────

# ── Household survey data ────────────────────────────────────────────────────
PATH_MICS_SMDW_WQ_OLD <- here::here("data/processed/household_surveys/df.SMDW_wq_MICS_old.csv")
PATH_MICS_SMDW_WQ_NEW <- here::here("data/processed/household_surveys/df.SMDW_wq_MICS_new.csv")
PATH_MICS_SMDW_OTHER  <- here::here("data/processed/household_surveys/df.SMDW_other_MICS.csv")
PATH_DHS_SMDW_WQ      <- here::here("data/processed/household_surveys/df.SMDW_wq_DHS.csv")
PATH_DHS_SMDW_OTHER   <- here::here("data/processed/household_surveys/df.SMDW_other_DHS.csv")

# ── Lookup tables ────────────────────────────────────────────────────────────
PATH_COUNTRY_KEY  <- here::here("configuration/country_name_key_WB.csv")

PATH_TO_REGION_AREAS_WQ    <- here::here("data/processed/sampled_features/water_quality_region_areas_km2.csv")
PATH_TO_REGION_AREAS_OTHER <- here::here("data/processed/sampled_features/other_surveys_region_areas_km2.csv")

# ── Outputs ──────────────────────────────────────────────────────────────────
OUTPUT_DIR           <- here::here("data/processed/training_subcomponents")
CROSSWALK_REVIEW_DIR <- here::here("data/processed/crosswalk_review")
```

------------------------------------------------------------------------

# 2. Input data

## Household survey data

Water-quality surveys (MICS new, MICS old, DHS) provide the E. coli and
SMDW composite. Other household surveys (MICS other, DHS other) provide
improved source, availability, and accessibility.

```{r}
df_mics_wq_new <- readr::read_csv(PATH_MICS_SMDW_WQ_NEW, show_col_types = FALSE) %>%
  dplyr::mutate(household_data_source = "MICS_water_quality")

df_mics_wq_old <- readr::read_csv(PATH_MICS_SMDW_WQ_OLD, show_col_types = FALSE) %>%
  dplyr::mutate(household_data_source = "MICS_water_quality")

df_dhs_wq <- readr::read_csv(PATH_DHS_SMDW_WQ, show_col_types = FALSE) %>%
  dplyr::mutate(
    HH7_region            = as.character(HH7_region),
    household_data_source = "DHS_water_quality"
  )

df_mics_other <- readr::read_csv(PATH_MICS_SMDW_OTHER, show_col_types = FALSE) %>%
  dplyr::mutate(household_data_source = "MICS_other")

df_dhs_other <- readr::read_csv(PATH_DHS_SMDW_OTHER, show_col_types = FALSE) %>%
  dplyr::mutate(household_data_source = "DHS_other")

# Water-quality pool: used for E. coli and SMDW composite
df_hh_water_quality <- dplyr::bind_rows(
  df_mics_wq_old,
  df_mics_wq_new,
  df_dhs_wq
)

# Other-survey pool: used for improved source, availability, accessibility
# Includes water-quality MICS data because those surveys also carry the
# other subcomponent questions.
df_hh_other <- dplyr::bind_rows(
  df_mics_wq_old,
  df_mics_wq_new,
  df_mics_other,
  df_dhs_other
)

# Standardise country name for DR Congo across both pools.
# The compile script assigns "DR Congo"; covariate files use the full name.
fix_country_names <- function(df) {
  df %>%
    dplyr::mutate(
      country = dplyr::case_when(
        country == "DR Congo" ~ "Democratic Republic of the Congo",
        TRUE ~ country
      )
    )
}

df_hh_water_quality <- fix_country_names(df_hh_water_quality)
df_hh_other         <- fix_country_names(df_hh_other)
```

## Lookup tables and metadata

```{r}
df_wgi_lmics        <- readWGI_lmics()
country_name_key_WB <- readr::read_csv(PATH_COUNTRY_KEY, show_col_types = FALSE)

oghist_raw <- suppressMessages(readxl::read_excel(
  PATH_OGHIST,
  sheet     = "Country Analytical History",
  col_names = FALSE
))

# JMP 2025 global dataset — two sheets
# "Sanitation" provides iso3, sdg_region, fragile_context, and country-level
#   sanitation estimates (used as contextual covariates for all outcomes).
# "Water" provides drinking water estimates specific to the SMDW outcomes.
jmp_raw_sanitation <- suppressMessages(readxl::read_excel(
  PATH_JMP,
  sheet     = "Sanitation",
  col_names = FALSE
))

jmp_raw_water <- suppressMessages(readxl::read_excel(
  PATH_JMP,
  sheet     = "Water",
  col_names = FALSE
))
```

------------------------------------------------------------------------

# 3. Survey selection

## Filter to study period (2015–2024)

Surveys outside the modelling window are dropped. Surveys near the
boundary are flagged before filtering so the exclusions are visible. The
same standardisation and year-range logic used for sanitation is applied
here to both household data pools.

```{r}
filter_to_study_period <- function(df) {
  df <- df %>%
    dplyr::mutate(
      HH5Y_original = HH5Y,
      HH5Y          = standardise_survey_year(country, HH5Y)
    )

  surveys_outside_range <- df %>%
    dplyr::filter(!is.na(HH5Y), HH5Y < 2015 | HH5Y > 2025) %>%
    dplyr::distinct(country, HH5Y_original, HH5Y) %>%
    dplyr::arrange(country, HH5Y)
  print(surveys_outside_range)

  countries_with_2024 <- df %>%
    dplyr::filter(HH5Y == 2024) %>%
    dplyr::distinct(country)

  df %>%
    dplyr::filter(
      is.na(HH5Y) | (HH5Y >= 2015 & HH5Y <= 2024) |
        (HH5Y == 2025 & country %in% countries_with_2024$country)
    )
}

df_hh_water_quality <- filter_to_study_period(df_hh_water_quality)
df_hh_other         <- filter_to_study_period(df_hh_other)
```

## Select most recent data source per country

Where a country has both MICS and DHS data, we keep the source with the
most recent survey year. Ties are broken by row count, then source name.
This is applied separately to each household data pool.

```{r}
keep_most_recent_source <- function(df, pool_label = "") {

  country_source_summary <- df %>%
    dplyr::mutate(HH5Y = as.integer(HH5Y)) %>%
    dplyr::group_by(country, household_data_source) %>%
    dplyr::summarise(
      min_year     = ifelse(all(is.na(HH5Y)), NA_integer_, min(HH5Y, na.rm = TRUE)),
      max_year     = ifelse(all(is.na(HH5Y)), NA_integer_, max(HH5Y, na.rm = TRUE)),
      survey_years = paste(sort(unique(HH5Y[!is.na(HH5Y)])), collapse = ", "),
      n_rows       = dplyr::n(),
      .groups      = "drop"
    )

  knitr::kable(
    country_source_summary %>% dplyr::arrange(country, household_data_source),
    caption = paste("Available surveys by country and data source", pool_label)
  ) %>% print()

  source_to_keep <- country_source_summary %>%
    dplyr::group_by(country) %>%
    dplyr::arrange(desc(!is.na(max_year)), desc(max_year), desc(n_rows),
                   household_data_source) %>%
    dplyr::slice(1) %>%
    dplyr::ungroup() %>%
    dplyr::select(country, household_data_source)

  source_to_drop <- country_source_summary %>%
    dplyr::anti_join(source_to_keep, by = c("country", "household_data_source")) %>%
    dplyr::arrange(country, household_data_source)

  knitr::kable(
    source_to_drop,
    caption = paste("Sources dropped (superseded by more recent data)", pool_label)
  ) %>% print()

  df %>% dplyr::semi_join(source_to_keep, by = c("country", "household_data_source"))
}

df_hh_water_quality <- keep_most_recent_source(df_hh_water_quality, "(water-quality pool)")
df_hh_other         <- keep_most_recent_source(df_hh_other,         "(other-surveys pool)")
```

## Drop Pakistan sub-national MICS surveys (other pool)

Pakistan sub-national MICS surveys (Balochistan, KP, Punjab, Sindh)
report at district level, which does not match the province-level
covariate boundaries. They are dropped from the other-surveys pool; the
national Pakistan DHS is used instead. The water-quality pool is
unaffected (Pakistan Punjab WQ uses province-level regions).

```{r}
pakistan_mics_other_dropped <- df_hh_other %>%
  dplyr::filter(
    stringr::str_detect(country, "^Pakistan ") &
    household_data_source %in% c("MICS_water_quality", "MICS_other")
  ) %>%
  dplyr::distinct(country, HH5Y, household_data_source)

knitr::kable(
  pakistan_mics_other_dropped,
  caption = "Pakistan sub-national MICS surveys dropped from other-surveys pool"
)

df_hh_other <- df_hh_other %>%
  dplyr::filter(
    !(stringr::str_detect(country, "^Pakistan ") &
      household_data_source %in% c("MICS_water_quality", "MICS_other"))
  )
```

## Region-level drops and merges

Known region labels that are uninformative or that need to be combined
before crosswalk matching are handled here.

```{r}
# ── Water-quality pool ───────────────────────────────────────────────────────
df_hh_water_quality <- df_hh_water_quality %>%
  dplyr::mutate(
    HH7_region = dplyr::case_when(
      # São Tomé and Príncipe: merge mainland districts
      country == "Sao Tome and Principe" & HH7_region %in% c(
        "DISTRITO DE ÁGUA GRANDE",
        "DISTRITO DE MÉ-ZÓCHI",
        "REGIÃO NORTE OESTE",
        "REGIÃO SUL ESTE"
      ) ~ "São Tomé",
      country == "Sao Tome and Principe" & HH7_region == "REGIÃO AUTÓNOMA DO PRÍNCIPE" ~ "Príncipe",
      # Samoa: merge Upolu sub-regions into one
      country == "Samoa" & HH7_region %in% c(
        "Apia Urban Area", "North West Upolu", "Rest of Upolu"
      ) ~ "Upolu",
      country == "Samoa" & HH7_region == "Savaii" ~ "Savaii",
      # Togo: merge Golfe Urbain and Lomé Commune
      country == "Togo" & HH7_region %in% c("Golfe Urbain", "Lomé Commune") ~ "Lomé Commune/Golfe Urbain",
      TRUE ~ HH7_region
    )
  )

# ── Other-surveys pool ───────────────────────────────────────────────────────
df_hh_other <- df_hh_other %>%
  dplyr::filter(
    !(country == "Nauru"  & HH7_region == "LOCATION")
  ) %>%
  dplyr::mutate(
    HH7_region = dplyr::case_when(
      # São Tomé and Príncipe: merge mainland districts
      country == "Sao Tome and Principe" & HH7_region %in% c(
        "DISTRITO DE ÁGUA GRANDE",
        "DISTRITO DE MÉ-ZÓCHI",
        "REGIÃO NORTE OESTE",
        "REGIÃO SUL ESTE"
      ) ~ "São Tomé",
      country == "Sao Tome and Principe" & HH7_region == "REGIÃO AUTÓNOMA DO PRÍNCIPE" ~ "Príncipe",
      # Samoa: merge Upolu sub-regions into one
      country == "Samoa" & HH7_region %in% c(
        "Apia Urban Area", "North West Upolu", "Rest of Upolu"
      ) ~ "Upolu",
      country == "Samoa" & HH7_region == "Savaii" ~ "Savaii",
      # Togo: merge Golfe Urbain and Lomé Commune
      country == "Togo" & HH7_region %in% c("Golfe Urbain", "Lomé Commune") ~ "Lomé Commune/Golfe Urbain",
      TRUE ~ HH7_region
    )
  )
```

## Write full household survey data

Full household-level observations (retained surveys only, most recent round) for both pools are merged and written to the switchdrive for sharing with collaborators. Raw survey data are protected and not stored in the GitHub repository.

Writing to switchdrive is commented out deliberately — this is a collaborator-facing deliverable and should only be regenerated intentionally, not as a side effect of rerunning this script. Uncomment when you actually want to publish an updated version.

```{r}
# SWITCHDRIVE_HH_DIR <- "/Users/esthergreenwood/switchdrive/Eawag/WorldBankProject/additional_data_0002020937/HH_surveys"
# dir.create(SWITCHDRIVE_HH_DIR, recursive = TRUE, showWarnings = FALSE)

df_hh_smdw_analysis <- dplyr::bind_rows(
  df_hh_water_quality %>% dplyr::mutate(pool = "water_quality"),
  df_hh_other         %>% dplyr::mutate(pool = "other")
)

# readr::write_csv(
#   df_hh_smdw_analysis,
#   file.path(SWITCHDRIVE_HH_DIR, "df_hh_smdw_analysis.csv")
# )
# message("Written: df_hh_smdw_analysis.csv")
```

------------------------------------------------------------------------

# 4. Geospatial covariates

## Build covariate tables

SMDW uses two covariate tables: one based on water-quality survey
boundaries (for E. coli and SMDW), one based on other survey boundaries
(for improved source, availability, accessibility). The `covariate_cols`
vector is defined in `configuration/covariate_cols.R`.

```{r}
pop_files_wq <- list.files(
  path       = path.expand(PATH_TO_POP_SUMS_WQ),
  pattern    = "\\.csv$",
  full.names = TRUE
)

pop_files_other <- list.files(
  path       = path.expand(PATH_TO_POP_SUMS_OTHER),
  pattern    = "\\.csv$",
  full.names = TRUE
)

df_training_covariates_wq <- make_training_covariates(
  pop_files      = pop_files_wq,
  nonpop_files   = FILE_NONPOP_WQ,
  covariate_cols = covariate_cols
)

df_training_covariates_other <- make_training_covariates(
  pop_files      = pop_files_other,
  nonpop_files   = FILE_NONPOP_OTHER,
  covariate_cols = covariate_cols
)
```

## Add region areas

```{r}
load_region_areas <- function(path) {
  readr::read_csv(path, show_col_types = FALSE) %>%
    dplyr::mutate(HH7_region = stringr::str_trim(HH7_region)) %>%
    dplyr::select(country, analysis_year, HH7_region, area_km2) %>%
    dplyr::group_by(country, analysis_year, HH7_region) %>%
    dplyr::slice(1) %>%
    dplyr::ungroup()
}

df_region_areas_wq    <- load_region_areas(PATH_TO_REGION_AREAS_WQ)
df_region_areas_other <- load_region_areas(PATH_TO_REGION_AREAS_OTHER)

df_training_covariates_wq <- df_training_covariates_wq %>%
  dplyr::left_join(df_region_areas_wq, by = c("country", "analysis_year", "HH7_region"))

df_training_covariates_other <- df_training_covariates_other %>%
  dplyr::left_join(df_region_areas_other, by = c("country", "analysis_year", "HH7_region"))

cat("WQ regions missing area_km2:    ", sum(is.na(df_training_covariates_wq$area_km2)),    "\n")
cat("Other regions missing area_km2: ", sum(is.na(df_training_covariates_other$area_km2)), "\n")
```

## Check for duplicate covariate keys

```{r}
check_duplicate_covariate_keys(df_training_covariates_wq)
check_duplicate_covariate_keys(df_training_covariates_other)
```

------------------------------------------------------------------------

# 5. Build regional outcomes

## Restrict to most recent survey year per country

Before building outcomes, each household data pool is filtered to the
most recent survey round (within a 2-year window) to avoid mixing
multiple rounds.

```{r}
keep_latest_survey_year_hh <- function(df, max_survey_duration = 2) {
  df %>%
    dplyr::group_by(country) %>%
    dplyr::filter(HH5Y >= max(HH5Y, na.rm = TRUE) - max_survey_duration) %>%
    dplyr::ungroup()
}

df_hh_water_quality_latest <- keep_latest_survey_year_hh(df_hh_water_quality)
df_hh_other_latest         <- keep_latest_survey_year_hh(df_hh_other)
```

## Analysis-year fixes for E. coli

Some countries require a manual override of the analysis year assigned
from the raw survey year. Add new entries here when new surveys are
incorporated.

```{r}
manual_analysis_year_fixes_ecoli <- tibble::tribble(
  ~country,                              ~analysis_year_manual,
  "Lao People's Democratic Republic",    2023L,
  "Palestina",                           2019L,
  "São Tomé and Príncipe",               2019L,
  "Zimbabwe",                            2019L
)
```

## Create outcomes

```{r}
SMDW_outcome             <- make_smdw_regional_outcome(df_hh_water_quality_latest)
ecoli_outcome            <- make_ecoli_regional_outcome(df_hh_water_quality_latest)
improved_source_outcome  <- make_improved_source_regional_outcome(df_hh_other_latest)
availability_outcome     <- make_availability_regional_outcome(df_hh_other_latest)
accessibility_outcome    <- make_accessibility_regional_outcome(df_hh_other_latest)
```

## Add PSU weights

Survey-design weights based on the number of primary sampling units are
attached to each region–year row. The unfiltered household data pools
are used here to count all available PSUs.

```{r}
wq_year_lookup    <- make_year_lookup_from_raw_household_data(
  df_hh_water_quality_latest,
  manual_analysis_year_fixes = manual_analysis_year_fixes_ecoli
)
other_year_lookup <- make_year_lookup_from_raw_household_data(df_hh_other_latest)

SMDW_outcome <- add_n_psu_weight(
  hh_df       = df_hh_water_quality,
  outcome_df  = SMDW_outcome,
  year_lookup = wq_year_lookup
)
ecoli_outcome <- add_n_psu_weight(
  hh_df       = df_hh_water_quality,
  outcome_df  = ecoli_outcome,
  year_lookup = wq_year_lookup
)
improved_source_outcome <- add_n_psu_weight(
  hh_df       = df_hh_other,
  outcome_df  = improved_source_outcome,
  year_lookup = other_year_lookup
)
availability_outcome <- add_n_psu_weight(
  hh_df       = df_hh_other,
  outcome_df  = availability_outcome,
  year_lookup = other_year_lookup
)
accessibility_outcome <- add_n_psu_weight(
  hh_df       = df_hh_other,
  outcome_df  = accessibility_outcome,
  year_lookup = other_year_lookup
)

# Quick sanity check on PSU coverage
dplyr::bind_rows(
  SMDW_outcome            %>% dplyr::mutate(outcome = "smdw"),
  ecoli_outcome           %>% dplyr::mutate(outcome = "ecoli_free"),
  improved_source_outcome %>% dplyr::mutate(outcome = "improved_source"),
  availability_outcome    %>% dplyr::mutate(outcome = "availability"),
  accessibility_outcome   %>% dplyr::mutate(outcome = "accessibility")
) %>%
  dplyr::group_by(outcome) %>%
  dplyr::summarise(
    n_rows        = dplyr::n(),
    missing_n_psu = sum(is.na(n_psu)),
    min_n_psu     = min(n_psu, na.rm = TRUE),
    median_n_psu  = median(n_psu, na.rm = TRUE),
    max_n_psu     = max(n_psu, na.rm = TRUE),
    .groups = "drop"
  )
```

------------------------------------------------------------------------

# 6. Covariate–outcome matching

## Filter covariates to outcome country–years

The full covariate tables are subset to only the country–year
combinations present in each outcome, using a country name crosswalk to
harmonise naming. Water-quality outcomes use the WQ covariate table;
other subcomponents use the other covariate table.

```{r}
df_covariates_SMDW <- filter_covariates_to_outcome_years(
  outcome_df             = SMDW_outcome,
  df_training_covariates = df_training_covariates_wq,
  country_name_key_WB    = country_name_key_WB
)
df_covariates_ecoli <- filter_covariates_to_outcome_years(
  outcome_df             = ecoli_outcome,
  df_training_covariates = df_training_covariates_wq,
  country_name_key_WB    = country_name_key_WB
)
df_covariates_improved_source <- filter_covariates_to_outcome_years(
  outcome_df             = improved_source_outcome,
  df_training_covariates = df_training_covariates_other,
  country_name_key_WB    = country_name_key_WB
)
df_covariates_availability <- filter_covariates_to_outcome_years(
  outcome_df             = availability_outcome,
  df_training_covariates = df_training_covariates_other,
  country_name_key_WB    = country_name_key_WB
)
df_covariates_accessibility <- filter_covariates_to_outcome_years(
  outcome_df             = accessibility_outcome,
  df_training_covariates = df_training_covariates_other,
  country_name_key_WB    = country_name_key_WB
)

check_duplicate_covariate_join_keys(df_covariates_SMDW)
check_duplicate_covariate_join_keys(df_covariates_ecoli)
check_duplicate_covariate_join_keys(df_covariates_improved_source)
check_duplicate_covariate_join_keys(df_covariates_availability)
check_duplicate_covariate_join_keys(df_covariates_accessibility)
```

## Diagnose unmatched country–years

Countries present in the outcome but absent from filtered covariates are
flagged here. Common causes: country name mismatch or missing covariate
data.

```{r}
country_year_filtering_issues <- dplyr::bind_rows(
  diagnose_country_year_filtering(
    outcome_name        = "smdw",
    outcome_df          = SMDW_outcome,
    raw_covariates      = df_training_covariates_wq,
    filtered_covariates = df_covariates_SMDW,
    country_name_key_WB = country_name_key_WB
  ),
  diagnose_country_year_filtering(
    outcome_name        = "ecoli_free",
    outcome_df          = ecoli_outcome,
    raw_covariates      = df_training_covariates_wq,
    filtered_covariates = df_covariates_ecoli,
    country_name_key_WB = country_name_key_WB
  ),
  diagnose_country_year_filtering(
    outcome_name        = "improved_source",
    outcome_df          = improved_source_outcome,
    raw_covariates      = df_training_covariates_other,
    filtered_covariates = df_covariates_improved_source,
    country_name_key_WB = country_name_key_WB
  ),
  diagnose_country_year_filtering(
    outcome_name        = "availability",
    outcome_df          = availability_outcome,
    raw_covariates      = df_training_covariates_other,
    filtered_covariates = df_covariates_availability,
    country_name_key_WB = country_name_key_WB
  ),
  diagnose_country_year_filtering(
    outcome_name        = "accessibility",
    outcome_df          = accessibility_outcome,
    raw_covariates      = df_training_covariates_other,
    filtered_covariates = df_covariates_accessibility,
    country_name_key_WB = country_name_key_WB
  )
)
print(country_year_filtering_issues)
```

## Build region crosswalks

Survey region names are matched to covariate region names using fuzzy
string matching. Rows that cannot be matched automatically are flagged
for manual review.

```{r}
crosswalk_SMDW <- build_crosswalk_for_one_outcome(
  outcome_name           = "smdw",
  outcome_df             = SMDW_outcome,
  df_training_covariates = df_covariates_SMDW,
  country_name_key_WB    = country_name_key_WB,
  output_dir             = CROSSWALK_REVIEW_DIR
)
crosswalk_ecoli <- build_crosswalk_for_one_outcome(
  outcome_name           = "ecoli_free",
  outcome_df             = ecoli_outcome,
  df_training_covariates = df_covariates_ecoli,
  country_name_key_WB    = country_name_key_WB,
  output_dir             = CROSSWALK_REVIEW_DIR
)
crosswalk_improved_source <- build_crosswalk_for_one_outcome(
  outcome_name           = "improved_source",
  outcome_df             = improved_source_outcome,
  df_training_covariates = df_covariates_improved_source,
  country_name_key_WB    = country_name_key_WB,
  output_dir             = CROSSWALK_REVIEW_DIR
)
crosswalk_availability <- build_crosswalk_for_one_outcome(
  outcome_name           = "availability",
  outcome_df             = availability_outcome,
  df_training_covariates = df_covariates_availability,
  country_name_key_WB    = country_name_key_WB,
  output_dir             = CROSSWALK_REVIEW_DIR
)
crosswalk_accessibility <- build_crosswalk_for_one_outcome(
  outcome_name           = "accessibility",
  outcome_df             = accessibility_outcome,
  df_training_covariates = df_covariates_accessibility,
  country_name_key_WB    = country_name_key_WB,
  output_dir             = CROSSWALK_REVIEW_DIR
)

# Inspect rows that need manual attention
crosswalk_SMDW$regions_to_review
crosswalk_ecoli$regions_to_review
crosswalk_improved_source$regions_to_review
crosswalk_availability$regions_to_review
crosswalk_accessibility$regions_to_review
```

## Write covariate region names for crosswalk review

```{r}
dir.create(CROSSWALK_REVIEW_DIR, recursive = TRUE, showWarnings = FALSE)

readr::write_csv(
  df_training_covariates_other %>%
    dplyr::distinct(country, analysis_year, source, HH7_region) %>%
    dplyr::arrange(country, analysis_year, source, HH7_region),
  file.path(CROSSWALK_REVIEW_DIR, "covariate_region_names_other.csv")
)
readr::write_csv(
  df_training_covariates_wq %>%
    dplyr::distinct(country, analysis_year, source, HH7_region) %>%
    dplyr::arrange(country, analysis_year, source, HH7_region),
  file.path(CROSSWALK_REVIEW_DIR, "covariate_region_names_wq.csv")
)
```

## Apply manual region name fixes

Known mismatches are resolved via the lookup tables in
`configuration/manual_region_fixes.R`. Add new entries there (not here)
when new surveys are incorporated.

-   `manual_fixes_smdw_ecoli` — used for SMDW composite and E. coli
    (water-quality survey boundaries; smaller set)
-   `manual_fixes_sanitation` — used for improved source, availability,
    accessibility (same survey boundaries as the sanitation pipeline)

```{r}
SMDW_crosswalk_final <- apply_manual_fixes_for_one_outcome(
  outcome_name        = "smdw",
  crosswalk_results   = crosswalk_SMDW,
  manual_region_fixes = manual_fixes_smdw_ecoli,
  output_dir          = CROSSWALK_REVIEW_DIR
)
ecoli_crosswalk_final <- apply_manual_fixes_for_one_outcome(
  outcome_name        = "ecoli_free",
  crosswalk_results   = crosswalk_ecoli,
  manual_region_fixes = manual_fixes_smdw_ecoli,
  output_dir          = CROSSWALK_REVIEW_DIR
)
improved_source_crosswalk_final <- apply_manual_fixes_for_one_outcome(
  outcome_name        = "improved_source",
  crosswalk_results   = crosswalk_improved_source,
  manual_region_fixes = manual_fixes_sanitation,
  output_dir          = CROSSWALK_REVIEW_DIR
)
availability_crosswalk_final <- apply_manual_fixes_for_one_outcome(
  outcome_name        = "availability",
  crosswalk_results   = crosswalk_availability,
  manual_region_fixes = manual_fixes_sanitation,
  output_dir          = CROSSWALK_REVIEW_DIR
)
accessibility_crosswalk_final <- apply_manual_fixes_for_one_outcome(
  outcome_name        = "accessibility",
  crosswalk_results   = crosswalk_accessibility,
  manual_region_fixes = manual_fixes_sanitation,
  output_dir          = CROSSWALK_REVIEW_DIR
)
```

## Join outcomes to covariates

```{r}
SMDW_training_output <- join_and_save_one_training_dataset(
  outcome_name           = "smdw",
  outcome_df             = SMDW_outcome,
  df_training_covariates = df_covariates_SMDW,
  region_crosswalk_final = SMDW_crosswalk_final,
  output_dir             = OUTPUT_DIR
)
ecoli_training_output <- join_and_save_one_training_dataset(
  outcome_name           = "ecoli_free",
  outcome_df             = ecoli_outcome,
  df_training_covariates = df_covariates_ecoli,
  region_crosswalk_final = ecoli_crosswalk_final,
  output_dir             = OUTPUT_DIR
)
improved_source_training_output <- join_and_save_one_training_dataset(
  outcome_name           = "improved_source",
  outcome_df             = improved_source_outcome,
  df_training_covariates = df_covariates_improved_source,
  region_crosswalk_final = improved_source_crosswalk_final,
  output_dir             = OUTPUT_DIR
)
availability_training_output <- join_and_save_one_training_dataset(
  outcome_name           = "availability",
  outcome_df             = availability_outcome,
  df_training_covariates = df_covariates_availability,
  region_crosswalk_final = availability_crosswalk_final,
  output_dir             = OUTPUT_DIR
)
accessibility_training_output <- join_and_save_one_training_dataset(
  outcome_name           = "accessibility",
  outcome_df             = accessibility_outcome,
  df_training_covariates = df_covariates_accessibility,
  region_crosswalk_final = accessibility_crosswalk_final,
  output_dir             = OUTPUT_DIR
)

# Model-ready objects
SMDW_training            <- SMDW_training_output$training_with_covariates
ecoli_training           <- ecoli_training_output$training_with_covariates
improved_source_training <- improved_source_training_output$training_with_covariates
availability_training    <- availability_training_output$training_with_covariates
accessibility_training   <- accessibility_training_output$training_with_covariates

# Diagnostic: regions that failed to join to covariates
SMDW_training_output$missing_covariates
ecoli_training_output$missing_covariates
improved_source_training_output$missing_covariates
availability_training_output$missing_covariates
accessibility_training_output$missing_covariates
```

------------------------------------------------------------------------

# 7. Country-level covariates

## GDP per capita and secondary education

Country-year level economic and education indicators from the World Bank
are joined by standardised country name. (Downloaded on 27 May 2026 from
<https://databank.worldbank.org/source/world-development-indicators/Type/TABLE/preview/on>)

```{r}
gdp_clean <- readr::read_csv(PATH_GDP, show_col_types = FALSE) %>%
  dplyr::rename(
    country_name                     = Country.Name,
    analysis_year                    = Year,
    gdp_per_capita_constant_2015_usd = `GDP per capita (constant 2015 US$)`
  ) %>%
  dplyr::mutate(
    analysis_year                    = as.integer(analysis_year),
    gdp_per_capita_constant_2015_usd = as.numeric(gdp_per_capita_constant_2015_usd)
  ) %>%
  add_country_join_key_extended(
    country_col = country_name, country_name_key_WB = country_name_key_WB
  ) %>%
  dplyr::select(country_join, analysis_year, gdp_per_capita_constant_2015_usd) %>%
  dplyr::distinct()

education_clean <- readr::read_csv(PATH_EDUCATION, show_col_types = FALSE) %>%
  dplyr::rename(
    country_name                       = Country.Name,
    analysis_year                      = Year,
    secondary_education_duration_years = `Secondary education, duration (years)`
  ) %>%
  dplyr::mutate(
    analysis_year                      = as.integer(analysis_year),
    secondary_education_duration_years = as.numeric(secondary_education_duration_years)
  ) %>%
  add_country_join_key_extended(
    country_col = country_name, country_name_key_WB = country_name_key_WB
  ) %>%
  dplyr::select(country_join, analysis_year, secondary_education_duration_years) %>%
  dplyr::distinct()

country_year_covariates <- dplyr::full_join(
  gdp_clean, education_clean, by = c("country_join", "analysis_year")
)

add_country_year_covariates <- function(training_df) {
  training_df %>%
    dplyr::mutate(analysis_year = as.integer(analysis_year)) %>%
    add_country_join_key_extended(
      country_col = country_outcome, country_name_key_WB = country_name_key_WB
    ) %>%
    dplyr::left_join(country_year_covariates, by = c("country_join", "analysis_year")) %>%
    dplyr::select(-country_join)
}

SMDW_training            <- add_country_year_covariates(SMDW_training)
ecoli_training           <- add_country_year_covariates(ecoli_training)
improved_source_training <- add_country_year_covariates(improved_source_training)
availability_training    <- add_country_year_covariates(availability_training)
accessibility_training   <- add_country_year_covariates(accessibility_training)

gdp_edu_vars <- c("gdp_per_capita_constant_2015_usd", "secondary_education_duration_years")
check_missing_covariates(SMDW_training,            "smdw",             gdp_edu_vars)
check_missing_covariates(ecoli_training,           "ecoli_free",       gdp_edu_vars)
check_missing_covariates(improved_source_training, "improved_source",  gdp_edu_vars)
check_missing_covariates(availability_training,    "availability",     gdp_edu_vars)
check_missing_covariates(accessibility_training,   "accessibility",    gdp_edu_vars)
```

## Wastewater treatment

Country-level wastewater collection, treatment, and reuse percentages.
These are static (no year dimension) and joined by country name only.
(Downloaded on 21 May 2026 from
<https://doi.pangaea.de/10.1594/PANGAEA.918731>)

```{r}
ww_country_clean <- readr::read_tsv(PATH_WASTEWATER, show_col_types = FALSE) %>%
  dplyr::rename(
    ww_country_name            = Country,
    ww_region                  = Region,
    ww_economic_classification = Economic_Classification,
    ww_collection_percent      = WWc_Percent,
    ww_treatment_percent       = WWt_Percent,
    ww_reuse_percent           = WWr_Percent
  ) %>%
  dplyr::mutate(across(ends_with("_percent"), as.numeric)) %>%
  add_country_join_key_extended(
    country_col = ww_country_name, country_name_key_WB = country_name_key_WB
  ) %>%
  dplyr::distinct(country_join, .keep_all = TRUE)

add_ww_covariates <- function(training_df) {
  training_df %>%
    dplyr::select(-dplyr::any_of(c(
      "country_join", "ww_country_name", "ww_region",
      "ww_economic_classification",
      "ww_collection_percent", "ww_treatment_percent", "ww_reuse_percent"
    ))) %>%
    add_country_join_key_extended(
      country_col = country_outcome, country_name_key_WB = country_name_key_WB
    ) %>%
    dplyr::left_join(
      ww_country_clean %>% dplyr::select(
        country_join, ww_region, ww_economic_classification,
        ww_collection_percent, ww_treatment_percent, ww_reuse_percent
      ),
      by = "country_join"
    ) %>%
    dplyr::select(-country_join)
}

SMDW_training            <- add_ww_covariates(SMDW_training)
ecoli_training           <- add_ww_covariates(ecoli_training)
improved_source_training <- add_ww_covariates(improved_source_training)
availability_training    <- add_ww_covariates(availability_training)
accessibility_training   <- add_ww_covariates(accessibility_training)

# Kosovo has no wastewater record; assign region manually
fix_ww_region_manual <- function(training_df) {
  training_df %>%
    dplyr::mutate(ww_region = dplyr::if_else(
      country_outcome == "Kosovo" & is.na(ww_region),
      "Eastern Europe & Central Asia", ww_region
    ))
}

SMDW_training            <- fix_ww_region_manual(SMDW_training)
ecoli_training           <- fix_ww_region_manual(ecoli_training)
improved_source_training <- fix_ww_region_manual(improved_source_training)
availability_training    <- fix_ww_region_manual(availability_training)
accessibility_training   <- fix_ww_region_manual(accessibility_training)

ww_vars <- c("ww_collection_percent", "ww_treatment_percent", "ww_reuse_percent")
check_missing_covariates(SMDW_training,            "smdw",            ww_vars)
check_missing_covariates(ecoli_training,           "ecoli_free",      ww_vars)
check_missing_covariates(improved_source_training, "improved_source", ww_vars)
check_missing_covariates(availability_training,    "availability",    ww_vars)
check_missing_covariates(accessibility_training,   "accessibility",   ww_vars)
```

## WGI governance indicators

Six World Governance Indicators are joined by country and year.
(Downloaded on 27 May 2026 from
<https://databank.worldbank.org/source/worldwide-governance-indicators/Type/TABLE/preview/on>)

```{r}
indicator_map <- c(
  "Government Effectiveness - Governance score (0-100)" = "governance_effectiveness",
  "Control of Corruption - Governance score (0-100)"    = "control_of_corruption",
  "Political Stability - Governance score (0-100)"      = "political_stability",
  "Regulatory Quality - Governance score (0-100)"       = "regulatory_quality",
  "Rule of Law - Governance score (0-100)"              = "rule_of_law",
  "Voice and Accountability - Governance score (0-100)" = "voice_and_accountability"
)

wgi_clean <- df_wgi_lmics %>%
  dplyr::filter(`Series.Name` %in% names(indicator_map)) %>%
  dplyr::mutate(indicator = unname(indicator_map[`Series.Name`])) %>%
  tidyr::pivot_longer(
    cols      = dplyr::matches("^X20[0-9]{2}$"),
    names_to  = "analysis_year",
    values_to = "value"
  ) %>%
  dplyr::mutate(
    analysis_year = as.integer(stringr::str_remove(analysis_year, "^X")),
    value         = as.numeric(value)
  ) %>%
  add_country_join_key_extended(
    country_col = `Country.Name`, country_name_key_WB = country_name_key_WB
  ) %>%
  dplyr::select(country_join, analysis_year, indicator, value) %>%
  tidyr::pivot_wider(names_from = indicator, values_from = value) %>%
  dplyr::distinct()

add_wgi_covariates <- function(training_df) {
  training_df %>%
    dplyr::select(-dplyr::any_of(c("country_join", unname(indicator_map)))) %>%
    dplyr::mutate(analysis_year = as.integer(analysis_year)) %>%
    add_country_join_key_extended(
      country_col = country_outcome, country_name_key_WB = country_name_key_WB
    ) %>%
    dplyr::left_join(wgi_clean, by = c("country_join", "analysis_year")) %>%
    dplyr::select(-country_join)
}

SMDW_training            <- add_wgi_covariates(SMDW_training)
ecoli_training           <- add_wgi_covariates(ecoli_training)
improved_source_training <- add_wgi_covariates(improved_source_training)
availability_training    <- add_wgi_covariates(availability_training)
accessibility_training   <- add_wgi_covariates(accessibility_training)

check_missing_covariates(SMDW_training,            "smdw",            unname(indicator_map))
check_missing_covariates(ecoli_training,           "ecoli_free",      unname(indicator_map))
check_missing_covariates(improved_source_training, "improved_source", unname(indicator_map))
check_missing_covariates(availability_training,    "availability",    unname(indicator_map))
check_missing_covariates(accessibility_training,   "accessibility",   unname(indicator_map))
```

------------------------------------------------------------------------

## World Bank Income Groups

Downloaded from the World Bank website on 13 June 2026
(<https://datahelpdesk.worldbank.org/knowledgebase/articles/906519-world-bank-country-and-lending-groups>)

```{r}
col_names_vec <- c(
  "country_code", "country_name",
  as.character(suppressWarnings(as.integer(as.numeric(unlist(oghist_raw[6, 3:40])))))
)

oghist_countries <- oghist_raw[12:nrow(oghist_raw), 1:40]
names(oghist_countries) <- col_names_vec

target_years <- as.character(2015:2024)

income_group_clean <- oghist_countries %>%
  dplyr::select(country_code, country_name, dplyr::all_of(target_years)) %>%
  dplyr::filter(!is.na(country_code), !is.na(country_name)) %>%
  tidyr::pivot_longer(
    cols      = dplyr::all_of(target_years),
    names_to  = "analysis_year",
    values_to = "wb_income_group"
  ) %>%
  dplyr::mutate(
    analysis_year   = as.integer(analysis_year),
    wb_income_group = dplyr::if_else(wb_income_group == "..", NA_character_, wb_income_group)
  )

add_income_group_covariates <- function(training_df) {
  training_df %>%
    dplyr::select(-dplyr::any_of(c("country_join", "wb_income_group"))) %>%
    dplyr::mutate(analysis_year = as.integer(analysis_year)) %>%
    add_country_join_key_extended(
      country_col         = country_outcome,
      country_name_key_WB = country_name_key_WB
    ) %>%
    dplyr::left_join(
      income_group_clean %>%
        add_country_join_key_extended(
          country_col         = country_name,
          country_name_key_WB = country_name_key_WB
        ) %>%
        dplyr::select(country_join, analysis_year, wb_income_group),
      by = c("country_join", "analysis_year")
    ) %>%
    dplyr::select(-country_join)
}

SMDW_training            <- add_income_group_covariates(SMDW_training)
ecoli_training           <- add_income_group_covariates(ecoli_training)
improved_source_training <- add_income_group_covariates(improved_source_training)
availability_training    <- add_income_group_covariates(availability_training)
accessibility_training   <- add_income_group_covariates(accessibility_training)

# Diagnostics: rows missing income group
check_missing_income_group <- function(training_df, outcome_name) {
  training_df %>%
    dplyr::filter(is.na(wb_income_group)) %>%
    dplyr::distinct(country_outcome, analysis_year) %>%
    dplyr::mutate(outcome_type = outcome_name) %>%
    dplyr::arrange(country_outcome, analysis_year)
}

missing_income_group <- dplyr::bind_rows(
  check_missing_income_group(SMDW_training,            "smdw"),
  check_missing_income_group(ecoli_training,           "ecoli_free"),
  check_missing_income_group(improved_source_training, "improved_source"),
  check_missing_income_group(availability_training,    "availability"),
  check_missing_income_group(accessibility_training,   "accessibility")
)
print(missing_income_group)
```

## JMP Indicators

Extracts country-level estimates and regional classifications from the
JMP 2025 global dataset (downloaded on 13 June 2026 from
<https://washdata.org/data/downloads>) and joins them to the training
data by country and year.

Two sheets are used:

-   **Sanitation sheet** — provides iso3, sdg_region, fragile_context,
    and country-level sanitation estimates used as contextual covariates
    across all SMDW outcomes.
-   **Water sheet** — provides drinking water estimates specific to the
    SMDW outcomes: at least basic, accessible on premises, available
    when needed, free from contamination, piped, and non-piped.

Time-invariant fields (iso3, sdg_region, fragile_context) are carried
forward from any available survey year to fill gaps where a specific
year is missing.

```{r}
# ── Helper: parse a numeric-or-bounded JMP cell ─────────────────────────────
parse_jmp_pct <- function(x) {
  dplyr::case_when(
    stringr::str_starts(x, "<") ~ 0,
    stringr::str_starts(x, ">") ~ 100,
    TRUE                        ~ suppressWarnings(as.numeric(x))
  )
}

# ── Sanitation sheet ─────────────────────────────────────────────────────────
# Columns (1-indexed): 1 country, 2 year, 17 sanitation_basic, 20 open_defecation,
#                      52 iso3, 53 sdg_region, 59 fragile_context
jmp_san_cols <- c(
  country_name     = 1,
  analysis_year    = 2,
  sanitation_basic = 17,
  open_defecation  = 20,
  iso3             = 52,
  sdg_region       = 53,
  fragile_context  = 59
)

jmp_san_clean <- jmp_raw_sanitation[4:nrow(jmp_raw_sanitation), unname(jmp_san_cols)] %>%
  purrr::set_names(names(jmp_san_cols)) %>%
  dplyr::filter(!is.na(country_name), !is.na(analysis_year)) %>%
  dplyr::mutate(
    analysis_year    = as.integer(analysis_year),
    sanitation_basic = parse_jmp_pct(sanitation_basic),
    open_defecation  = parse_jmp_pct(open_defecation)
  )

# Country-level lookup for time-invariant fields (iso3, sdg_region, fragile_context)
jmp_country_meta <- jmp_san_clean %>%
  dplyr::group_by(country_name) %>%
  dplyr::summarise(
    iso3            = dplyr::first(iso3[!is.na(iso3)]),
    sdg_region      = dplyr::first(sdg_region[!is.na(sdg_region)]),
    fragile_context = dplyr::first(fragile_context[!is.na(fragile_context)]),
    .groups = "drop"
  ) %>%
  add_country_join_key_extended(
    country_col         = country_name,
    country_name_key_WB = country_name_key_WB
  ) %>%
  dplyr::select(country_join, iso3, sdg_region, fragile_context)

# Year-varying sanitation estimates
jmp_san_estimates <- jmp_san_clean %>%
  dplyr::select(country_name, analysis_year, sanitation_basic, open_defecation) %>%
  add_country_join_key_extended(
    country_col         = country_name,
    country_name_key_WB = country_name_key_WB
  ) %>%
  dplyr::select(country_join, analysis_year, sanitation_basic, open_defecation)

# ── Water sheet ──────────────────────────────────────────────────────────────
# Columns (1-indexed): 1 country, 2 year,
#   15 (O)  at least basic,
#   38 (AL) accessible on premises,
#   39 (AM) available when needed,
#   40 (AN) free from contamination,
#   42 (AP) piped,
#   43 (AQ) non-piped
jmp_water_cols <- c(
  country_name                          = 1,
  analysis_year                         = 2,
  drinking_water_basic                  = 15,
  drinking_water_accessible_on_premises = 38,
  drinking_water_available_when_needed  = 39,
  drinking_water_free_from_contamination = 40,
  drinking_water_piped                  = 42,
  drinking_water_non_piped              = 43
)

jmp_water_vars <- c(
  "drinking_water_basic", "drinking_water_accessible_on_premises",
  "drinking_water_available_when_needed", "drinking_water_free_from_contamination",
  "drinking_water_piped", "drinking_water_non_piped"
)

jmp_water_estimates <- jmp_raw_water[4:nrow(jmp_raw_water), unname(jmp_water_cols)] %>%
  purrr::set_names(names(jmp_water_cols)) %>%
  dplyr::filter(!is.na(country_name), !is.na(analysis_year)) %>%
  dplyr::mutate(
    analysis_year = as.integer(analysis_year),
    dplyr::across(dplyr::all_of(jmp_water_vars), parse_jmp_pct)
  ) %>%
  add_country_join_key_extended(
    country_col         = country_name,
    country_name_key_WB = country_name_key_WB
  ) %>%
  dplyr::select(country_join, analysis_year, dplyr::all_of(jmp_water_vars))

# ── Join all JMP covariates into training datasets ───────────────────────────
add_jmp_covariates <- function(training_df) {
  training_df %>%
    dplyr::select(-dplyr::any_of(c(
      "country_join",
      "sanitation_basic", "open_defecation",
      "iso3", "sdg_region", "fragile_context",
      jmp_water_vars
    ))) %>%
    dplyr::mutate(analysis_year = as.integer(analysis_year)) %>%
    add_country_join_key_extended(
      country_col         = country_outcome,
      country_name_key_WB = country_name_key_WB
    ) %>%
    dplyr::left_join(jmp_san_estimates,   by = c("country_join", "analysis_year")) %>%
    dplyr::left_join(jmp_water_estimates, by = c("country_join", "analysis_year")) %>%
    dplyr::left_join(jmp_country_meta,    by = "country_join") %>%
    dplyr::select(-country_join)
}

SMDW_training            <- add_jmp_covariates(SMDW_training)
ecoli_training           <- add_jmp_covariates(ecoli_training)
improved_source_training <- add_jmp_covariates(improved_source_training)
availability_training    <- add_jmp_covariates(availability_training)
accessibility_training   <- add_jmp_covariates(accessibility_training)

# Manual patch for countries whose names don't match JMP
patch_jmp_meta <- function(training_df) {
  training_df %>%
    dplyr::mutate(
      iso3 = dplyr::case_when(
        country_outcome == "Kosovo"    & is.na(iso3) ~ "XKX",
        country_outcome == "Palestina" & is.na(iso3) ~ "PSE",
        country_outcome == "Tanzania"  & is.na(iso3) ~ "TZA",
        TRUE                                         ~ iso3
      ),
      sdg_region = dplyr::case_when(
        country_outcome == "Kosovo"    & is.na(sdg_region) ~ "Europe and Northern America",
        country_outcome == "Palestina" & is.na(sdg_region) ~ "Northern Africa and Western Asia",
        country_outcome == "Tanzania"  & is.na(sdg_region) ~ "Sub-Saharan Africa",
        TRUE                                               ~ sdg_region
      )
    )
}

SMDW_training            <- patch_jmp_meta(SMDW_training)
ecoli_training           <- patch_jmp_meta(ecoli_training)
improved_source_training <- patch_jmp_meta(improved_source_training)
availability_training    <- patch_jmp_meta(availability_training)
accessibility_training   <- patch_jmp_meta(accessibility_training)

# Diagnostics
jmp_check_vars <- c("drinking_water_basic", "sanitation_basic", "iso3", "sdg_region")

check_missing_jmp <- function(training_df, outcome_name) {
  training_df %>%
    dplyr::filter(dplyr::if_any(dplyr::all_of(jmp_check_vars), is.na)) %>%
    dplyr::distinct(country_outcome, analysis_year,
                    dplyr::across(dplyr::all_of(jmp_check_vars))) %>%
    dplyr::mutate(outcome_type = outcome_name) %>%
    dplyr::arrange(country_outcome, analysis_year)
}

missing_jmp <- dplyr::bind_rows(
  check_missing_jmp(SMDW_training,            "smdw"),
  check_missing_jmp(ecoli_training,           "ecoli_free"),
  check_missing_jmp(improved_source_training, "improved_source"),
  check_missing_jmp(availability_training,    "availability"),
  check_missing_jmp(accessibility_training,   "accessibility")
)
print(missing_jmp)
```

------------------------------------------------------------------------

# 8. Output

## Write training CSVs

Final training datasets are written to `data/training_subcomponents/`.

```{r}
dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)

readr::write_csv(SMDW_training,
  file.path(OUTPUT_DIR, "smdw_training_with_covariates.csv"))
readr::write_csv(ecoli_training,
  file.path(OUTPUT_DIR, "ecoli_free_training_with_covariates.csv"))
readr::write_csv(improved_source_training,
  file.path(OUTPUT_DIR, "improved_source_training_with_covariates.csv"))
readr::write_csv(availability_training,
  file.path(OUTPUT_DIR, "availability_training_with_covariates.csv"))
readr::write_csv(accessibility_training,
  file.path(OUTPUT_DIR, "accessibility_training_with_covariates.csv"))

message("Written: smdw_training_with_covariates.csv")
message("Written: ecoli_free_training_with_covariates.csv")
message("Written: improved_source_training_with_covariates.csv")
message("Written: availability_training_with_covariates.csv")
message("Written: accessibility_training_with_covariates.csv")
```

## Summary: wastewater region distribution

A quick check on regional representation in the final training data.

```{r}
dplyr::bind_rows(
  SMDW_training            %>% dplyr::mutate(outcome_type = "smdw"),
  ecoli_training           %>% dplyr::mutate(outcome_type = "ecoli_free"),
  improved_source_training %>% dplyr::mutate(outcome_type = "improved_source"),
  availability_training    %>% dplyr::mutate(outcome_type = "availability"),
  accessibility_training   %>% dplyr::mutate(outcome_type = "accessibility")
) %>%
  dplyr::mutate(ww_region = dplyr::if_else(
    is.na(ww_region), "Missing ww_region", ww_region
  )) %>%
  dplyr::count(outcome_type, ww_region, name = "n_observations") %>%
  dplyr::group_by(outcome_type) %>%
  dplyr::mutate(percent = round(100 * n_observations / sum(n_observations), 2)) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(outcome_type, desc(n_observations))
```
