---
title: "Preparing the Final Sanitation Estimates Dataframe"
author: "Esther Greenwood"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: flatly
    self-contained: true
execute:
  warning: false
  message: false
---

## Overview

This script prepares the final sanitation estimates dataset submitted as a project deliverable. It combines two admin1-level prediction files from the TabPFN model into a single joined dataframe:

- **Basic sanitation** — proportion of the population using at least basic sanitation services
- **Open defecation** — proportion of the population practising open defecation

Both outcomes are predicted at the first-level administrative (admin1) district level across up to 130 low- and middle-income countries (LMICs). For each outcome the model outputs a median prediction and a 90% prediction interval.

The joined file is saved directly to the project deliverables folder.

```{r}
#| label: setup
library(tidyverse)
source(here::here("configuration/paths.R"))   # loads switchdrive paths
```

---

## Step 1: Load the Sanitation Prediction Files

We load the two sanitation prediction CSVs and select the same set of columns used across all deliverable datasets: the district identifiers, contextual grouping variables, and the three prediction columns (median, lower and upper bounds of the 90% prediction interval). The prediction columns are renamed with a suffix identifying the outcome.

```{r}
#| label: load-predictions

pred_dir <- here::here("outputs/model_performance/predictions")

id_cols   <- c("GID_0", "GID_1", "NAME_0", "NAME_1",
               "sdg_region", "wb_income_group", "fragile_context")
pred_cols <- c("median", "pi90_lower", "pi90_upper")

read_predictions <- function(filename, suffix) {
  read_csv(file.path(pred_dir, filename), show_col_types = FALSE) |>
    select(all_of(c(id_cols, pred_cols))) |>
    rename_with(~ paste0(., "_", suffix), all_of(pred_cols))
}

basic_sanitation <- read_predictions("basic_sanitation_tabpfn_lmic_predictions.csv", "basic_sanitation")
open_defecation  <- read_predictions("open_defecation_tabpfn_lmic_predictions.csv",  "open_defecation")

cat("Basic sanitation rows:", nrow(basic_sanitation),
    "| Countries:", n_distinct(basic_sanitation$GID_0), "\n")
cat("Open defecation rows:", nrow(open_defecation),
    "| Countries:", n_distinct(open_defecation$GID_0), "\n")
```

---

## Step 2: Join into a Single Dataframe

The two files share the same set of admin1 districts. We join on `GID_0` and `GID_1`, retaining the district identifier and contextual columns from the basic sanitation file as the reference.

```{r}
#| label: join

sanitation <- basic_sanitation |>
  left_join(
    open_defecation |> select(GID_0, GID_1,
                               median_open_defecation,
                               pi90_lower_open_defecation,
                               pi90_upper_open_defecation),
    by = c("GID_0", "GID_1")
  )

cat("Joined dataset rows:", nrow(sanitation),
    "| Countries:", n_distinct(sanitation$GID_0), "\n")
cat("Districts missing open defecation predictions:",
    sum(is.na(sanitation$median_open_defecation)), "\n")

glimpse(sanitation)
```

---

## Step 3: Diagnostics

```{r}
#| label: diagnostics

sanitation |>
  select(median_basic_sanitation, median_open_defecation) |>
  summary()
```

```{r}
#| label: plot

sanitation |>
  pivot_longer(
    cols = c(median_basic_sanitation, median_open_defecation),
    names_to  = "indicator",
    values_to = "median"
  ) |>
  mutate(indicator = recode(indicator,
    median_basic_sanitation = "Basic sanitation",
    median_open_defecation  = "Open defecation"
  )) |>
  ggplot(aes(x = median, fill = wb_income_group)) +
  geom_histogram(bins = 50, colour = "white", linewidth = 0.2) +
  facet_wrap(~ indicator, scales = "free_x") +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(
    title = "Distribution of Sanitation Estimates by Admin1 District",
    subtitle = "Across up to 130 LMICs",
    x    = "Predicted proportion",
    y    = "Number of districts",
    fill = "World Bank income group"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

## Step 4: Export

```{r}
#| label: export

out_file <- PATH_OUT_SANITATION

sanitation |>
  write_csv(out_file)

cat("Results saved to:\n", out_file, "\n")
cat("Rows exported:", nrow(sanitation), "\n")
```
