---
title: "Sanitation RF Models — Feature Set Comparison & Final Model"
subtitle: "Basic sanitation and open defecation outcomes"
---

# 1. Setup

## Parameters

Tune here before running. Use fewer trees and repeats for fast
exploratory runs; increase for final model selection.

```{r}
# ── Feature set comparison ────────────────────────────────────────────────────
K_VALUES            <- c(5, 10, 15, 20, 30, 45, 90, 100)   # cluster k to compare (99 = full sanitation feature set)
N_TREES_COMPARISON  <- 100    # trees per model during feature set comparison
N_REPEATS           <- 5      # stratified CV repeats per feature set
FOLD_FRACTION       <- 0.15   # target fraction of training obs held out per fold

# ── Final model ───────────────────────────────────────────────────────────────
# Set manually after inspecting comparison results (Section 6)
FINAL_K             <- 45   # e.g. 20; NULL until chosen
N_TREES_FINAL       <- 500

# ── Reproducibility ───────────────────────────────────────────────────────────
SEED <- 123
```

## Libraries

```{r}
library(tidyverse)
library(here)
library(h2o)
library(svglite)
library(viridis)
library(forcats)
```

## Paths

```{r}
TRAINING_DIR          <- here::here("data/training_subcomponents")
CLUSTER_OUTPUT_DIR    <- here::here("outputs/03b_cluster_analysis/sanitation")
PRED_COVARIATES_PATH  <- here::here("data/prediction/prediction_covariates_2024.csv")
OUTPUT_DIR            <- here::here("outputs/04_model_performance")
FIGURE_DIR            <- file.path(OUTPUT_DIR, "figures")
dir.create(FIGURE_DIR, recursive = TRUE, showWarnings = FALSE)
```

------------------------------------------------------------------------

# 2. Data

## Load training data

```{r}
basic_sanitation_raw <- readr::read_csv(
  file.path(TRAINING_DIR, "basic_sanitation_training_with_covariates.csv"),
  show_col_types = FALSE
)

open_defecation_raw <- readr::read_csv(
  file.path(TRAINING_DIR, "open_defecation_training_with_covariates.csv"),
  show_col_types = FALSE
)
```

## Remove excluded countries

```{r}
remove_excluded_countries <- function(df) {
  df %>%
    dplyr::filter(!stringr::str_detect(
      stringr::str_to_lower(country_outcome),
      "^(indonesia)$"
    ))
}

basic_sanitation_df <- remove_excluded_countries(basic_sanitation_raw)
open_defecation_df <- open_defecation_raw
```

## Column name constants

```{r}
OUTCOME_COL  <- "outcome_value"
WEIGHT_COL   <- "n_psu_weight_scaled"
COUNTRY_COL  <- "country_cov"
SDG_COL      <- "sdg_region"
INCOME_COL   <- "wb_income_group"
```

## Full feature list (k = 103 baseline)

```{r}
all_EO_features <- c(
  "CGIAR_Aridity_Index", "CGIAR_PET",
  "CHELSA_BIO_Annual_Mean_Temperature",
  "CHELSA_BIO_Annual_Precipitation",
  "CHELSA_BIO_Precipitation_Seasonality",
  "CHELSA_BIO_Precipitation_of_Coldest_Quarter",
  "CHELSA_BIO_Precipitation_of_Driest_Month",
  "CHELSA_BIO_Precipitation_of_Driest_Quarter",
  "CHELSA_BIO_Precipitation_of_Warmest_Quarter",
  "CHELSA_BIO_Precipitation_of_Wettest_Month",
  "CHELSA_BIO_Precipitation_of_Wettest_Quarter",
  "CHELSA_BIO_Temperature_Annual_Range",
  "CHELSA_BIO_Temperature_Seasonality",
  "CIFOR_TropicalPeatlandExtent",
  "CSP_Global_Human_Modification",
  "ConsensusLandCoverClass_Barren",
  "ConsensusLandCoverClass_Cultivated_and_Managed_Vegetation",
  "ConsensusLandCoverClass_Deciduous_Broadleaf_Trees",
  "ConsensusLandCoverClass_Evergreen_Broadleaf_Trees",
  "ConsensusLandCoverClass_Evergreen_Deciduous_Needleleaf_Trees",
  "ConsensusLandCoverClass_Herbaceous_Vegetation",
  "ConsensusLandCoverClass_Mixed_Other_Trees",
  "ConsensusLandCoverClass_Open_Water",
  "ConsensusLandCoverClass_Regularly_Flooded_Vegetation",
  "ConsensusLandCoverClass_Shrubs",
  "ConsensusLandCoverClass_Snow_Ice",
  "ConsensusLandCoverClass_Urban_Builtup",
  "ConsensusLandCover_Human_Development_Percentage",
  "CrowtherLab_Tree_Density",
  "EarthEnvCloudCover_CloudForestPrediction",
  "EarthEnvTexture_CoOfVar_EVI",
  "EarthEnvTexture_Contrast_EVI",
  "EarthEnvTexture_Correlation_EVI",
  "EarthEnvTexture_Dissimilarity_EVI",
  "EarthEnvTexture_Entropy_EVI",
  "EarthEnvTexture_Evenness_EVI",
  "EarthEnvTexture_Homogeneity_EVI",
  "EarthEnvTexture_Maximum_EVI",
  "EarthEnvTexture_Range_EVI",
  "EarthEnvTexture_Shannon_Index",
  "EarthEnvTexture_Simpson_Index",
  "EarthEnvTexture_Std_EVI",
  "EarthEnvTexture_Uniformity_EVI",
  "EarthEnvTexture_Variance_EVI",
  "EarthEnvTopoMed_1stOrderPartialDerivEW",
  "EarthEnvTopoMed_1stOrderPartialDerivNS",
  "EarthEnvTopoMed_Eastness",
  "EarthEnvTopoMed_Elevation",
  "EarthEnvTopoMed_Roughness",
  "EarthEnvTopoMed_Slope",
  "EarthEnvTopoMed_TerrainRuggednessIndex",
  "EarthEnvTopoMed_TopoPositionIndex",
  "EsaCci_BurntAreasProbability",
  "FanEtAl_Depth_to_Water_Table_AnnualMean",
  "FanEtAl_Depth_to_Water_Table_AnnualSD",
  "GHS_Population_Density",
  "GLW3_RuminantsDistribution_downsampled10km",
  "GPWv4_Population_Density",
  "GiriEtAl_MangrovesExtent",
  "MODIS_EVI", "MODIS_NDVI", "MODIS_NPP",
  "PelletierEtAl_SoilAndSedimentaryDepositThicknesses",
  "SG_Absolute_depth_to_bedrock",
  "SG_Bulk_density_015cm",
  "SG_Depth_to_bedrock",
  "SG_H2O_Capacity_015cm",
  "SG_Saturated_H2O_Content_015cm",
  "TootchiEtAl_WetlandsRegularlyFlooded",
  "WCS_Human_Footprint_2009",
  "chirps_annual_precipitation",
  "era5_temperature_2m",
  "ghsl_built_surface",
  "ghsl_population",
  "ghsl_urban_frac",
  "jrc_building_height",
  "map_friction",
  "modis_evi",
  "modis_ndvi",
  "runoff_max_annualmax",
  "runoff_min_annualmin",
  "temperature_2m_max_annualmax",
  "viirs_average",
  "worldpop",
  "worldpop_sum",
  "ghsl_population_sum",
  "area_km2"
)

country_level_features <- c(
  "gdp_per_capita_constant_2015_usd",
  "secondary_education_duration_years",
  "ww_collection_percent",
  "ww_treatment_percent",
  "ww_reuse_percent",
  "control_of_corruption",
  "governance_effectiveness",
  "political_stability",
  "regulatory_quality",
  "rule_of_law",
  "voice_and_accountability",
  "sanitation_basic",
  "open_defecation"
)

all_features_k103 <- c(all_EO_features, country_level_features)

cat("Full sanitation feature set (k = 99):", length(all_features_k103), "features\n")
```

## Load cluster feature sets

Representative variables per cluster, selected manually in
`01_cluster_analysis.qmd`. Each CSV has columns `cluster` and
`representative_variable`.

k = 103 uses the full feature set above (no cluster file needed).

```{r}
# Load the representative variable list for a given shortlist size k.
# - k = 99 is a special code meaning "use the full feature set" (no shortlisting).
# - For any other k, reads the CSV saved by the cluster analysis script and
#   returns only variables that also exist in the master feature list.
load_feature_set <- function(k, full_features = all_features_k103) {
  if (k == 99) return(full_features)  # 99 = full sanitation feature set

  # Build the path to the CSV saved during cluster analysis for this k
  path <- file.path(CLUSTER_OUTPUT_DIR,
                    paste0("representative_variables_k", k, ".csv"))

  # Stop with a clear message if the cluster file hasn't been generated yet
  if (!file.exists(path)) {
    stop("Cluster feature file not found: ", path,
         "\nRun 01_cluster_analysis.qmd first.")
  }

  # Read the CSV, extract the variable names, and drop any that are not
  # in the master feature list (guards against stale or mismatched files)
  readr::read_csv(path, show_col_types = FALSE) %>%
    dplyr::pull(representative_variable) %>%
    intersect(full_features)
}

# Run load_feature_set() for every k in K_VALUES and store results in a named list.
# Example: feature_sets[["k20"]] contains the 20 representative variable names.
feature_sets <- purrr::set_names(
  purrr::map(K_VALUES, load_feature_set),
  paste0("k", K_VALUES)
)

# Print a summary table showing how many variables were loaded for each k —
# useful to confirm all cluster files were found and none were unexpectedly empty
purrr::imap_dfr(feature_sets, ~ tibble::tibble(k = .y, n_features = length(.x))) %>%
  knitr::kable(caption = "Features per cluster k")
```

------------------------------------------------------------------------

# 3. Stratified fold sampler

Folds are defined at the **country level**: all regions of a held-out
country are excluded from training. This prevents data leakage — no
partial-country exposure.

Target proportions come from the prediction dataset (admin-1 regions by
SDG region × income group), so the held-out fold mirrors the global
distribution the model will predict into.

## Country region counts

Countries regions are sampled with probability proportional to their number of
training regions (probability-proportional-to-size sampling). Larger
countries are therefore more likely to be drawn, which naturally fills
the fold target in fewer draws.

```{r}
country_strata <- basic_sanitation_df %>%
  dplyr::group_by(country = .data[[COUNTRY_COL]]) %>%
  dplyr::summarise(
    sdg_region      = dplyr::first(.data[[SDG_COL]]),
    wb_income_group = dplyr::first(.data[[INCOME_COL]]),
    n_regions       = dplyr::n(),
    .groups         = "drop"
  )

cat("Training countries:", nrow(country_strata), "\n")
cat("Total training regions:", sum(country_strata$n_regions), "\n")
print(country_strata %>% dplyr::arrange(desc(n_regions)))
```
## Define fractions of SDG regions and income groups represented

```{r}
pred_covariates <- readr::read_csv(PRED_COVARIATES_PATH, show_col_types = FALSE)

total_train_regions <- nrow(basic_sanitation_df)

# Fractions by SDG region
pred_covariates %>%
  dplyr::filter(!is.na(sdg_region)) %>%
  dplyr::count(sdg_region, name = "n_pred_regions") %>%
  dplyr::mutate(
    fraction  = round(n_pred_regions / sum(n_pred_regions), 3),
    n_in_fold = round(fraction * FOLD_FRACTION * total_train_regions)
  ) %>%
  dplyr::arrange(desc(n_pred_regions))

# Fractions by income group
pred_covariates %>%
  dplyr::filter(!is.na(wb_income_group)) %>%
  dplyr::count(wb_income_group, name = "n_pred_regions") %>%
  dplyr::mutate(
    fraction  = round(n_pred_regions / sum(n_pred_regions), 3),
    n_in_fold = round(fraction * FOLD_FRACTION * total_train_regions)
  ) %>%
  dplyr::arrange(desc(n_pred_regions))
```



```{r}
# ── Stratum targets from prediction space ─────────────────────────────────────
# Count how many prediction-space regions fall in each SDG geographic zone,
# then convert to proportions. Use those proportions to set a target number
# of training regions to hold out per zone per fold — this ensures each test
# fold mirrors the geographic mix of the places we ultimately want to predict.
stratum_targets <- pred_covariates %>%
  dplyr::filter(!is.na(sdg_region)) %>%
  dplyr::count(sdg_region, name = "n_pred_regions") %>%
  dplyr::mutate(
    fraction  = n_pred_regions / sum(n_pred_regions),
    n_in_fold = round(fraction * FOLD_FRACTION * total_train_regions)
  )

cat("Target regions per stratum per fold:\n")
print(stratum_targets %>% dplyr::arrange(desc(n_pred_regions)))

# ── Country stratum lookup ────────────────────────────────────────────────────
# Build a one-row-per-country reference table recording each country's
# SDG region, World Bank income group, and how many sub-national regions
# it contributes to the training data. Used by the fold sampler below.
country_strata <- basic_sanitation_df %>%
  dplyr::group_by(country = .data[[COUNTRY_COL]]) %>%
  dplyr::summarise(
    sdg_region      = dplyr::first(.data[[SDG_COL]]),
    wb_income_group = dplyr::first(.data[[INCOME_COL]]),
    n_regions       = dplyr::n(),
    .groups         = "drop"
  )

# ── Fold sampler ──────────────────────────────────────────────────────────────
# For each stratum, sample whole countries without replacement until the
# cumulative region count reaches n_in_fold for that stratum.
# Countries are drawn with equal probability within each stratum.
# High-income countries (wb_income_group == "H") are never held out —
# they are always kept in training because the model is intended for
# lower- and middle-income settings.
# If a stratum has fewer training regions than targeted, all its countries
# are held out and a warning is issued.

sample_held_out_countries <- function(
    country_strata,
    stratum_targets,
    seed_i = 1
) {
  set.seed(seed_i)   # different seed per fold repeat produces independent folds

  held_out <- character(0)

  # Remove high-income countries from the pool before sampling
  eligible_countries <- country_strata %>%
    dplyr::filter(wb_income_group != "H")

  for (stratum in stratum_targets$sdg_region) {

    # How many regions should this stratum contribute to the fold?
    target_n <- stratum_targets$n_in_fold[stratum_targets$sdg_region == stratum]

    # All eligible countries in this SDG zone
    stratum_countries <- eligible_countries %>%
      dplyr::filter(sdg_region == stratum)

    if (nrow(stratum_countries) == 0 || target_n == 0) next

    sampled   <- character(0)
    sampled_n <- 0
    remaining <- stratum_countries

    # Keep drawing countries one at a time (without replacement) until
    # enough sub-national regions have been accumulated for this stratum
    while (sampled_n < target_n && nrow(remaining) > 0) {
      idx     <- sample(nrow(remaining), size = 1)
      chosen  <- remaining$country[idx]
      sampled <- c(sampled, chosen)
      sampled_n <- sampled_n + remaining$n_regions[idx]
      remaining <- remaining[-idx, ]
    }

    # Warn if the stratum ran out of countries before reaching the target
    if (sampled_n < target_n) {
      warning("Stratum '", stratum, "': only ", sampled_n,
              " regions available, target was ", target_n,
              ". All countries in this stratum held out.")
    }

    held_out <- c(held_out, sampled)
  }

  held_out
}

# ── Generate folds ────────────────────────────────────────────────────────────
# Create N_REPEATS independent held-out country lists, each using a different
# random seed so the folds are not correlated with one another.
set.seed(SEED)
fold_country_lists <- purrr::map(
  seq_len(N_REPEATS),
  ~ sample_held_out_countries(country_strata, stratum_targets, seed_i = .x)
)

# ── Diagnostics ───────────────────────────────────────────────────────────────
# For each fold, report how many countries and regions were held out,
# what percentage of the total training data that represents, and how many
# SDG zones are covered. Ideally all folds should be similar in size (~15%)
# and cover multiple strata.
fold_diagnostics <- purrr::imap_dfr(fold_country_lists, function(countries, i) {
  fold_regions <- basic_sanitation_df %>%
    dplyr::filter(.data[[COUNTRY_COL]] %in% countries)

  tibble::tibble(
    repeat_i         = i,
    n_countries      = length(countries),
    n_regions        = nrow(fold_regions),
    pct_of_total     = round(100 * nrow(fold_regions) / total_train_regions, 1),
    n_strata_covered = dplyr::n_distinct(fold_regions[[SDG_COL]])
  )
})

cat("Fold diagnostics across", N_REPEATS, "repeats:\n")
print(fold_diagnostics)
cat("Mean % of training data held out per fold:",
    round(mean(fold_diagnostics$pct_of_total), 1), "%\n")
```

```{r}
# ── Prediction space distributions ───────────────────────────────────────────
pred_dist_income <- pred_covariates %>%
  dplyr::filter(!is.na(.data[[INCOME_COL]])) %>%
  dplyr::count(.data[[INCOME_COL]], name = "n") %>%
  dplyr::rename(wb_income_group = 1) %>%
  dplyr::mutate(pct = 100 * n / sum(n), source = "Prediction space")

pred_dist_sdg <- pred_covariates %>%
  dplyr::filter(!is.na(.data[[SDG_COL]])) %>%
  dplyr::count(.data[[SDG_COL]], name = "n") %>%
  dplyr::rename(sdg_region = 1) %>%
  dplyr::mutate(pct = 100 * n / sum(n), source = "Prediction space")

# ── Fold distributions ────────────────────────────────────────────────────────
fold_dist_income <- purrr::imap_dfr(fold_country_lists, function(countries, i) {
  basic_sanitation_df %>%
    dplyr::filter(.data[[COUNTRY_COL]] %in% countries,
                  !is.na(.data[[INCOME_COL]])) %>%
    dplyr::count(.data[[INCOME_COL]], name = "n") %>%
    dplyr::rename(wb_income_group = 1) %>%
    dplyr::mutate(pct = 100 * n / sum(n), source = paste0("Fold ", i))
})

fold_dist_sdg <- purrr::imap_dfr(fold_country_lists, function(countries, i) {
  basic_sanitation_df %>%
    dplyr::filter(.data[[COUNTRY_COL]] %in% countries,
                  !is.na(.data[[SDG_COL]])) %>%
    dplyr::count(.data[[SDG_COL]], name = "n") %>%
    dplyr::rename(sdg_region = 1) %>%
    dplyr::mutate(pct = 100 * n / sum(n), source = paste0("Fold ", i))
})

# ── Plot: income group ────────────────────────────────────────────────────────
# Note: folds were stratified by SDG region only, so income group
# distribution across folds may vary more than the SDG region plot.
p_income <- dplyr::bind_rows(pred_dist_income, fold_dist_income) %>%
  dplyr::mutate(is_pred = source == "Prediction space") %>%
  ggplot2::ggplot(ggplot2::aes(x = wb_income_group, y = pct)) +
  ggplot2::geom_col(
    data     = ~ dplyr::filter(.x, !is_pred),
    mapping  = ggplot2::aes(group = source),
    position = ggplot2::position_dodge(width = 0.7),
    width = 0.6, fill = "black", alpha = 0.5
  ) +
  ggplot2::geom_col(
    data  = ~ dplyr::filter(.x, is_pred),
    width = 0.6, fill = "steelblue", alpha = 0.85
  ) +
  ggplot2::labs(
    title    = "Region distribution by income group",
    subtitle = "Black = prediction space, blue = individual folds",
    x = "World Bank income group", y = "% of regions"
  ) +
  ggplot2::theme_bw(base_size = 10) +
  ggplot2::theme(legend.position = "none")

# ── Plot: SDG region ──────────────────────────────────────────────────────────
p_sdg <- dplyr::bind_rows(pred_dist_sdg, fold_dist_sdg) %>%
  dplyr::mutate(
    is_pred    = source == "Prediction space",
    sdg_region = stringr::str_wrap(sdg_region, width = 20)
  ) %>%
  ggplot2::ggplot(ggplot2::aes(x = sdg_region, y = pct)) +
  ggplot2::geom_col(
    data     = ~ dplyr::filter(.x, !is_pred),
    mapping  = ggplot2::aes(group = source),
    position = ggplot2::position_dodge(width = 0.7),
    width = 0.6, fill = "black", alpha = 0.5
  ) +
  ggplot2::geom_col(
    data  = ~ dplyr::filter(.x, is_pred),
    width = 0.6, fill = "steelblue", alpha = 0.85
  ) +
  ggplot2::labs(
    title    = "Region distribution by SDG region",
    subtitle = "Black = prediction space, blue = individual folds",
    x = NULL, y = "% of regions"
  ) +
  ggplot2::theme_bw(base_size = 10) +
  ggplot2::theme(
    legend.position = "none",
    axis.text.x     = ggplot2::element_text(angle = 35, hjust = 1)
  )

p_income
p_sdg
```



------------------------------------------------------------------------

# 4. Feature set comparison (stratified CV)

For each feature set k and each repeat, the RF is trained on the
non-held-out countries and evaluated on the held-out countries. Weighted
MAE, RMSE, and R² are averaged across repeats.

## H2O initialisation

```{r}
h2o::h2o.init(nthreads = -1, max_mem_size = "12G")
h2o::h2o.no_progress()
```

## CV helper functions

```{r}
# Compute unweighted metrics on a data frame with columns
# `observed` and `predicted`.
# PSU weights are used only for model fitting, not for evaluation metrics.
compute_metrics <- function(df) {
  r      <- df$observed - df$predicted
  ss_res <- sum(r^2)
  ss_tot <- sum((df$observed - mean(df$observed))^2)

  tibble::tibble(
    mae  = mean(abs(r)),
    rmse = sqrt(mean(r^2)),
    r2   = 1 - ss_res / ss_tot,
    n    = nrow(df)
  )
}

# Single train/predict cycle for one fold
run_one_fold <- function(
    training_df,
    held_out_countries,
    features,
    outcome_col   = OUTCOME_COL,
    weight_col    = WEIGHT_COL,
    country_col   = COUNTRY_COL,
    ntrees        = N_TREES_COMPARISON,
    seed          = SEED
) {
  train_df <- training_df %>%
    dplyr::filter(!.data[[country_col]] %in% held_out_countries) %>%
    dplyr::select(dplyr::all_of(c(features, outcome_col, weight_col))) %>%
    tidyr::drop_na(dplyr::all_of(c(outcome_col, weight_col)))

  test_df <- training_df %>%
    dplyr::filter(.data[[country_col]] %in% held_out_countries) %>%
    dplyr::select(dplyr::all_of(c(features, outcome_col))) %>%
    tidyr::drop_na(dplyr::all_of(outcome_col))

  if (nrow(test_df) == 0) return(NULL)

  valid_features <- intersect(features, names(train_df))

  train_h2o <- h2o::as.h2o(train_df)
  test_h2o  <- h2o::as.h2o(test_df)

  rf <- h2o::h2o.randomForest(
    x              = valid_features,
    y              = outcome_col,
    training_frame = train_h2o,
    ntrees         = ntrees,
    weights_column = weight_col,
    seed           = seed
  )

  preds <- h2o::h2o.predict(rf, test_h2o) %>%
    as.data.frame() %>%
    dplyr::pull(predict)

  h2o::h2o.rm(rf@model_id)
  h2o::h2o.rm(train_h2o)
  h2o::h2o.rm(test_h2o)

  compute_metrics(tibble::tibble(
    observed  = test_df[[outcome_col]],
    predicted = preds
  ))
}
```

## Run comparison across all feature sets and outcomes

This loop runs `length(K_VALUES) × N_REPEATS` models per outcome.
Expect this to take some time.

```{r}
# Run the feature set comparison for one outcome (e.g. basic sanitation).
# For every shortlist size (k5, k10, ...) and every fold repeat:
#   - Train on all countries except the held-out ones
#   - Predict on the held-out countries
#   - Record accuracy metrics
# Then average the metrics across the N_REPEATS folds to get a stable estimate
# of how well each shortlist size performs for this outcome.
run_feature_set_comparison <- function(training_df, outcome_name) {

  message("== ", outcome_name, " ==")

  # Loop over every feature shortlist size
  purrr::imap_dfr(feature_sets, function(features, k_label) {
    message("  Feature set ", k_label, " (", length(features), " features)")

    # Loop over the N_REPEATS independent folds and collect per-fold metrics.
    # A different random seed per repeat ensures any stochasticity in the model
    # does not systematically favour one fold over another.
    repeat_metrics <- purrr::imap_dfr(fold_country_lists, function(countries, rep_i) {
      result <- run_one_fold(
        training_df        = training_df,
        held_out_countries = countries,
        features           = features,
        seed               = SEED + rep_i
      )
      if (is.null(result)) return(NULL)   # skip fold if test set was empty
      dplyr::mutate(result, repeat_i = rep_i)
    })

    # Summarise across folds: mean and SD for each accuracy metric.
    # SD tells us how stable the performance is across different held-out sets.
    repeat_metrics %>%
      dplyr::summarise(
        mean_mae  = mean(mae,  na.rm = TRUE),
        sd_mae    = sd(mae,    na.rm = TRUE),
        mean_rmse = mean(rmse, na.rm = TRUE),
        sd_rmse   = sd(rmse,   na.rm = TRUE),
        mean_r2   = mean(r2,   na.rm = TRUE),
        sd_r2     = sd(r2,     na.rm = TRUE),
        n_repeats = dplyr::n()
      ) %>%
      # Tag each row with the outcome name, shortlist label, and feature count
      # so results from different outcomes and k values can be compared in one table
      dplyr::mutate(
        outcome      = outcome_name,
        feature_set  = k_label,
        n_features   = length(features)
      )
  })
}

# Run the comparison separately for each outcome
comparison_basic_sanitation <- run_feature_set_comparison(
  basic_sanitation_df, "basic_sanitation"
)
comparison_open_defecation <- run_feature_set_comparison(
  open_defecation_df, "open_defecation"
)

# Stack both outcome results into a single table for easy side-by-side comparison
cv_comparison_results <- dplyr::bind_rows(
  comparison_basic_sanitation,
  comparison_open_defecation
)

# Save to CSV so the results can be reviewed before selecting the final k values
readr::write_csv(
  cv_comparison_results,
  file.path(OUTPUT_DIR, "feature_set_cv_comparison.csv")
)
```

## Results summary

```{r}
cv_comparison_results %>%
  dplyr::select(outcome, feature_set, n_features,
                mean_r2, sd_r2, mean_mae, sd_mae, mean_rmse, sd_rmse) %>%
  dplyr::arrange(outcome, desc(mean_r2)) %>%
  knitr::kable(digits = 3, caption = "Stratified CV results by feature set")
```

```{r}
# Plot: R² across feature sets by outcome
cv_comparison_results %>%
  dplyr::mutate(feature_set = forcats::fct_reorder(feature_set, n_features)) %>%
  ggplot2::ggplot(ggplot2::aes(
    x    = feature_set,
    y    = mean_r2,
    ymin = mean_r2 - sd_r2,
    ymax = mean_r2 + sd_r2,
    colour = outcome,
    group  = outcome
  )) +
  ggplot2::geom_line() +
  ggplot2::geom_point(size = 2) +
  ggplot2::geom_errorbar(width = 0.2, alpha = 0.5) +
  ggplot2::labs(
    title  = "Stratified CV performance by feature set",
    x      = "Feature set (k clusters)",
    y      = "Mean R² (±1 SD across repeats)",
    colour = "Outcome"
  ) +
  ggplot2::theme_bw(base_size = 10)
```

------------------------------------------------------------------------

# 5. Select final model

**MANUAL STEP** — inspect the CV table and plot above, then set
`FINAL_K_BASIC_SANITATION` and `FINAL_K_OPEN_DEFECATION` below. The
two outcomes can differ. Re-run from here once both are set.

```{r}
FINAL_K_BASIC_SANITATION <- 45   # e.g. 20
FINAL_K_OPEN_DEFECATION  <- 100   # e.g. 15

if (is.null(FINAL_K_BASIC_SANITATION) || is.null(FINAL_K_OPEN_DEFECATION)) {
  stop("Set both FINAL_K_BASIC_SANITATION and FINAL_K_OPEN_DEFECATION before continuing.")
}

final_features_basic_sanitation <- feature_sets[[paste0("k", FINAL_K_BASIC_SANITATION)]]
final_features_open_defecation  <- feature_sets[[paste0("k", FINAL_K_OPEN_DEFECATION)]]

cat("Basic sanitation — k =", FINAL_K_BASIC_SANITATION,
    "| features:", length(final_features_basic_sanitation), "\n")
cat("Open defecation  — k =", FINAL_K_OPEN_DEFECATION,
    "| features:", length(final_features_open_defecation), "\n")
```

------------------------------------------------------------------------

# 6. Final model: LOCO-CV

The final model for each outcome is trained with leave-one-country-out
cross-validation using its selected feature set. LOCO-CV is kept
strictly separate from the feature selection step above.

```{r}
# Clear all H2O models from feature comparison to free memory
h2o::h2o.removeAll()

prepare_loco_data <- function(training_df, features, outcome_name) {
  model_features <- intersect(features, names(training_df))

  training_df %>%
    dplyr::mutate(
      country_fold = as.factor(as.integer(factor(.data[[COUNTRY_COL]])))
    ) %>%
    dplyr::select(
      dplyr::all_of(model_features),
      dplyr::all_of(OUTCOME_COL),
      dplyr::all_of(WEIGHT_COL),
      country_fold,
      country_outcome,
      HH7_region_outcome,
      analysis_year,
      sdg_region,
      wb_income_group,
      ww_region
    ) %>%
    tidyr::drop_na(
      dplyr::all_of(OUTCOME_COL),
      dplyr::all_of(WEIGHT_COL)
    ) %>%
    list(data = ., features = model_features, outcome_name = outcome_name)
}

run_rf_loco <- function(loco_input, final_k, ntrees = N_TREES_FINAL, seed = SEED) {
  train_data     <- loco_input$data
  model_features <- loco_input$features
  outcome_name   <- loco_input$outcome_name

  train_h2o <- h2o::as.h2o(
    train_data,
    destination_frame = paste0("loco_train_", outcome_name)
  )

  rf_model <- h2o::h2o.randomForest(
    x                                     = model_features,
    y                                     = OUTCOME_COL,
    training_frame                        = train_h2o,
    ntrees                                = ntrees,
    mtries                                = -1,
    sample_rate                           = 0.632,
    fold_column                           = "country_fold",
    weights_column                        = WEIGHT_COL,
    seed                                  = seed,
    keep_cross_validation_predictions     = TRUE,
    keep_cross_validation_fold_assignment = TRUE
  )

  cv_preds <- h2o::h2o.getFrame(
    rf_model@model$cross_validation_holdout_predictions_frame_id$name
  ) %>%
    as.data.frame()

  cv_eval <- train_data %>%
    dplyr::mutate(
      predicted = cv_preds$predict,
      observed  = .data[[OUTCOME_COL]]
    )

  metrics <- tibble::tibble(
    outcome_name = outcome_name,
    feature_set  = paste0("k", final_k),
    n_features   = length(model_features),
    n_rows       = nrow(train_data),
    n_countries  = dplyr::n_distinct(train_data[[COUNTRY_COL]]),
    loco_r2      = as.numeric(h2o::h2o.r2(rf_model,   xval = TRUE)),
    loco_mae     = as.numeric(h2o::h2o.mae(rf_model,  xval = TRUE)),
    loco_rmse    = as.numeric(h2o::h2o.rmse(rf_model, xval = TRUE))
  )

  var_imp <- h2o::h2o.varimp(rf_model) %>%
    as.data.frame() %>%
    dplyr::mutate(outcome_name = outcome_name)

  list(
    model       = rf_model,
    cv_eval     = cv_eval,
    metrics     = metrics,
    varimp      = var_imp,
    features    = model_features,
    final_k     = final_k
  )
}

print_and_save_svg <- function(plot_obj, filename, width = 8, height = 6) {
  print(plot_obj)
  ggplot2::ggsave(
    filename = file.path(FIGURE_DIR, filename),
    plot     = plot_obj,
    device   = svglite::svglite,
    width    = width, height = height, units = "in"
  )
}

plot_pred_obs <- function(cv_eval, training_df, plot_title) {
  region_meta <- training_df %>%
  dplyr::distinct(country_outcome, HH7_region_outcome, analysis_year,
                  sdg_region, ww_region)

  cv_eval %>%
    dplyr::left_join(region_meta,
                     by = c("country_outcome", "HH7_region_outcome",
                            "analysis_year")) %>%
    dplyr::mutate(
      sdg_region = dplyr::coalesce(sdg_region.x, sdg_region.y),
      ww_region  = dplyr::coalesce(ww_region.x,  ww_region.y)
    ) %>%
    ggplot2::ggplot(ggplot2::aes(
      x      = observed,
      y      = predicted,
      colour = sdg_region,
      size   = .data[[WEIGHT_COL]]
    )) +
    ggplot2::geom_point(shape = 1, alpha = 0.8) +
    ggplot2::coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
    ggplot2::geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    ggplot2::labs(
      title  = plot_title,
      x      = "Observed proportion",
      y      = "Predicted proportion",
      colour = "SDG region",
      size   = "PSU weight"
    ) +
    ggplot2::theme_minimal(base_size = 9)
}
```

## Basic sanitation

```{r}
loco_basic_sanitation <- prepare_loco_data(
  basic_sanitation_df, final_features_basic_sanitation, "basic_sanitation"
) %>% run_rf_loco(final_k = FINAL_K_BASIC_SANITATION)

print(loco_basic_sanitation$metrics)

readr::write_csv(
  loco_basic_sanitation$metrics,
  file.path(OUTPUT_DIR,
            paste0("loco_metrics_basic_sanitation_k",
                   FINAL_K_BASIC_SANITATION, ".csv"))
)

plot_loco_basic_sanitation <- plot_pred_obs(
  loco_basic_sanitation$cv_eval,
  basic_sanitation_df,
  paste0("Basic sanitation — LOCO-CV (k = ", FINAL_K_BASIC_SANITATION, ")")
)
plot_loco_basic_sanitation
print_and_save_svg(
  plot_loco_basic_sanitation,
  paste0("loco_pred_obs_basic_sanitation_k", FINAL_K_BASIC_SANITATION, ".svg"),
  width = 7, height = 5
)
```

## Open defecation

```{r}
loco_open_defecation <- prepare_loco_data(
  open_defecation_df, final_features_open_defecation, "open_defecation"
) %>% run_rf_loco(final_k = FINAL_K_OPEN_DEFECATION)

print(loco_open_defecation$metrics)

readr::write_csv(
  loco_open_defecation$metrics,
  file.path(OUTPUT_DIR,
            paste0("loco_metrics_open_defecation_k",
                   FINAL_K_OPEN_DEFECATION, ".csv"))
)

plot_loco_open_defecation <- plot_pred_obs(
  loco_open_defecation$cv_eval,
  open_defecation_df,
  paste0("Open defecation — LOCO-CV (k = ", FINAL_K_OPEN_DEFECATION, ")")
)
plot_loco_open_defecation
print_and_save_svg(
  plot_loco_open_defecation,
  paste0("loco_pred_obs_open_defecation_k", FINAL_K_OPEN_DEFECATION, ".svg"),
  width = 7, height = 5
)
```

------------------------------------------------------------------------

# 7. SHAP values (final models only)

```{r}
feature_label_map <- c(
  "worldpop"                               = "WorldPop population density",
  "worldpop_sum"                           = "WorldPop population sum",
  "ghsl_population"                        = "GHSL population density",
  "ghsl_population_sum"                    = "GHSL population sum",
  "GHS_Population_Density"                 = "GHS population density",
  "GPWv4_Population_Density"               = "GPWv4 population density",
  "ghsl_built_surface"                     = "Built surface fraction",
  "ghsl_urban_frac"                        = "Urban fraction",
  "jrc_building_height"                    = "Building height",
  "viirs_average"                          = "Night-time lights",
  "gdp_per_capita_constant_2015_usd"       = "GDP per capita",
  "secondary_education_duration_years"     = "Secondary education duration",
  "ww_collection_percent"                  = "Wastewater collection",
  "ww_treatment_percent"                   = "Wastewater treatment",
  "ww_reuse_percent"                       = "Wastewater reuse",
  "control_of_corruption"                  = "Control of corruption",
  "governance_effectiveness"               = "Government effectiveness",
  "political_stability"                    = "Political stability",
  "regulatory_quality"                     = "Regulatory quality",
  "rule_of_law"                            = "Rule of law",
  "voice_and_accountability"               = "Voice and accountability",
  "sanitation_basic"                       = "JMP basic sanitation (national)",
  "open_defecation"                        = "JMP open defecation (national)",
  "CGIAR_Aridity_Index"                    = "Aridity index",
  "CGIAR_PET"                              = "Potential evapotranspiration",
  "CHELSA_BIO_Annual_Mean_Temperature"     = "Annual mean temperature",
  "CHELSA_BIO_Annual_Precipitation"        = "Annual precipitation",
  "EarthEnvTopoMed_Elevation"              = "Elevation",
  "EarthEnvTopoMed_Slope"                  = "Slope",
  "FanEtAl_Depth_to_Water_Table_AnnualMean" = "Depth to water table",
  "map_friction"                           = "Travel friction",
  "WCS_Human_Footprint_2009"               = "Human footprint",
  "CSP_Global_Human_Modification"          = "Human modification"
)

label_feature <- function(x) {
  x <- as.character(x)
  mapped <- dplyr::recode(x, !!!feature_label_map, .default = NA_character_)
  dplyr::if_else(
    is.na(mapped),
    stringr::str_replace_all(x, "_", " ") %>% stringr::str_squish(),
    mapped
  )
}

make_shap_plot <- function(loco_result, y_lab, sample_n = 2000) {
  outcome_name <- loco_result$outcome_name
  final_k      <- loco_result$final_k

  shap_df <- loco_result$cv_eval %>%
    dplyr::select(dplyr::all_of(loco_result$features))

  if (nrow(shap_df) > sample_n) {
    set.seed(SEED)
    shap_df <- shap_df %>% dplyr::slice_sample(n = sample_n)
  }

  shap_h2o <- h2o::as.h2o(shap_df,
                            destination_frame = paste0("shap_", outcome_name))
  raw_plot  <- h2o::h2o.shap_summary_plot(loco_result$model, shap_h2o)

  plot_data <- raw_plot[["data"]] %>%
    dplyr::mutate(
      feature       = label_feature(as.character(feature)),
      mean_abs_shap = mean(abs(contribution), na.rm = TRUE),
      .by           = feature
    )

  ggplot2::ggplot(
    dplyr::mutate(plot_data,
                  feature = forcats::fct_reorder(feature, mean_abs_shap)),
    ggplot2::aes(x = feature, y = contribution, colour = normalized_value)
  ) +
    ggplot2::geom_hline(yintercept = 0, linetype = "dashed") +
    ggplot2::geom_point(
      position = h2o:::position_jitter_density(),
      alpha = 0.5, size = 1.2
    ) +
    ggplot2::scale_colour_gradientn(
      colours = viridis::viridis_pal()(9),
      breaks  = c(0, 1), labels = c("low", "high"), limits = c(0, 1)
    ) +
    ggplot2::coord_flip() +
    ggplot2::labs(
      title  = paste("SHAP summary:", outcome_name, "(k =", final_k, ")"),
      y      = y_lab, x = "",
      colour = "Normalised\nfeature value"
    ) +
    ggplot2::theme_bw(base_size = 9)
}
```

## Basic sanitation

```{r}
shap_basic_sanitation <- make_shap_plot(
  loco_basic_sanitation,
  "SHAP contribution to basic sanitation prediction"
)
shap_basic_sanitation
print_and_save_svg(
  shap_basic_sanitation,
  paste0("shap_basic_sanitation_k", FINAL_K_BASIC_SANITATION, ".svg"),
  width = 9, height = 6
)
```

## Open defecation

```{r}
shap_open_defecation <- make_shap_plot(
  loco_open_defecation,
  "SHAP contribution to open defecation prediction"
)
shap_open_defecation
print_and_save_svg(
  shap_open_defecation,
  paste0("shap_open_defecation_k", FINAL_K_OPEN_DEFECATION, ".svg"),
  width = 9, height = 6
)
```

------------------------------------------------------------------------

# 8. Save final models

```{r}
model_dir <- here::here("outputs/models")
dir.create(model_dir, recursive = TRUE, showWarnings = FALSE)

h2o::h2o.saveModel(loco_basic_sanitation$model, path = model_dir, force = TRUE)
h2o::h2o.saveModel(loco_open_defecation$model,  path = model_dir, force = TRUE)

message("Models saved to: ", model_dir)
```
