---
title: "1_rf_modelling_water_quality"
format: html
editor: visual
---

```{r}
# ============================================================
# 1. Read final training datasets
# ============================================================

training_dir <- here::here("./data/training_subcomponents")

ecoli_training_with_covariates <- readr::read_csv(
  file.path(training_dir, "ecoli_training_with_covariates.csv"),
  show_col_types = FALSE
)

improved_source_training_with_covariates <- readr::read_csv(
  file.path(training_dir, "improved_source_training_with_covariates.csv"),
  show_col_types = FALSE
)

availability_training_with_covariates <- readr::read_csv(
  file.path(training_dir, "availability_training_with_covariates.csv"),
  show_col_types = FALSE
)

accessibility_training_with_covariates <- readr::read_csv(
  file.path(training_dir, "accessibility_training_with_covariates.csv"),
  show_col_types = FALSE
)
```


```{r}
all_EO_features <- c(
  "CGIAR_Aridity_Index", "CGIAR_PET",
  "CHELSA_BIO_Annual_Mean_Temperature",
  "CHELSA_BIO_Annual_Precipitation",
  "CHELSA_BIO_Precipitation_Seasonality",
  "CHELSA_BIO_Precipitation_of_Coldest_Quarter",
  "CHELSA_BIO_Precipitation_of_Driest_Month",
  "CHELSA_BIO_Precipitation_of_Driest_Quarter",
  "CHELSA_BIO_Precipitation_of_Warmest_Quarter",
  "CHELSA_BIO_Precipitation_of_Wettest_Month",
  "CHELSA_BIO_Precipitation_of_Wettest_Quarter",
  "CHELSA_BIO_Temperature_Annual_Range",
  "CHELSA_BIO_Temperature_Seasonality",
  "CIFOR_TropicalPeatlandExtent",
  "CSP_Global_Human_Modification",
  "ConsensusLandCoverClass_Barren",
  "ConsensusLandCoverClass_Cultivated_and_Managed_Vegetation",
  "ConsensusLandCoverClass_Deciduous_Broadleaf_Trees",
  "ConsensusLandCoverClass_Evergreen_Broadleaf_Trees",
  "ConsensusLandCoverClass_Evergreen_Deciduous_Needleleaf_Trees",
  "ConsensusLandCoverClass_Herbaceous_Vegetation",
  "ConsensusLandCoverClass_Mixed_Other_Trees",
  "ConsensusLandCoverClass_Open_Water",
  "ConsensusLandCoverClass_Regularly_Flooded_Vegetation",
  "ConsensusLandCoverClass_Shrubs",
  "ConsensusLandCoverClass_Snow_Ice",
  "ConsensusLandCoverClass_Urban_Builtup",
  "ConsensusLandCover_Human_Development_Percentage",
  "CrowtherLab_Tree_Density",
  "EarthEnvCloudCover_CloudForestPrediction",
  "EarthEnvTexture_CoOfVar_EVI",
  "EarthEnvTexture_Contrast_EVI",
  "EarthEnvTexture_Correlation_EVI",
  "EarthEnvTexture_Dissimilarity_EVI",
  "EarthEnvTexture_Entropy_EVI",
  "EarthEnvTexture_Evenness_EVI",
  "EarthEnvTexture_Homogeneity_EVI",
  "EarthEnvTexture_Maximum_EVI",
  "EarthEnvTexture_Range_EVI",
  "EarthEnvTexture_Shannon_Index",
  "EarthEnvTexture_Simpson_Index",
  "EarthEnvTexture_Std_EVI",
  "EarthEnvTexture_Uniformity_EVI",
  "EarthEnvTexture_Variance_EVI",
  "EarthEnvTopoMed_1stOrderPartialDerivEW",
  "EarthEnvTopoMed_1stOrderPartialDerivNS",
  "EarthEnvTopoMed_Eastness",
  "EarthEnvTopoMed_Elevation",
  "EarthEnvTopoMed_Roughness",
  "EarthEnvTopoMed_Slope",
  "EarthEnvTopoMed_TerrainRuggednessIndex",
  "EarthEnvTopoMed_TopoPositionIndex",
  "EsaCci_BurntAreasProbability",
  "FanEtAl_Depth_to_Water_Table_AnnualMean",
  "FanEtAl_Depth_to_Water_Table_AnnualSD",
  "GHS_Population_Density",
  "GLW3_RuminantsDistribution_downsampled10km",
  "GPWv4_Population_Density",
  "GiriEtAl_MangrovesExtent",
  "MODIS_EVI", "MODIS_NDVI", "MODIS_NPP",
  "PelletierEtAl_SoilAndSedimentaryDepositThicknesses",
  "SG_Absolute_depth_to_bedrock",
  "SG_Bulk_density_015cm",
  "SG_Depth_to_bedrock",
  "SG_H2O_Capacity_015cm",
  "SG_Saturated_H2O_Content_015cm",
  "TootchiEtAl_WetlandsRegularlyFlooded",
  "WCS_Human_Footprint_2009",
  "chirps_annual_precipitation",
  "era5_temperature_2m",
  "ghsl_built_surface",
  "ghsl_population",
  "ghsl_urban_frac",
  "jrc_building_height",
  "map_friction",
  "modis_evi",
  "modis_ndvi",
  "runoff_max_annualmax",
  "runoff_min_annualmin",
  "temperature_2m_max_annualmax",
  "viirs_average",
  "worldpop",
  "worldpop_sum",
  "ghsl_population_sum"
)

```

```{r}
# -----------------------------
# 1. Setup
# -----------------------------

outcome_col <- "No_EcoliAtRegionalLevel"
weight_col  <- "HouseholdsInRegion.Freq.x"
country_col <- "country_cov"   # use country_cov for the joined training data

eo_features <- intersect(all_EO_features, names(EcoliFreeRegionalProportion_with_covariates_enriched))

```

```{r}
# -----------------------------
# 2. Create LOCO-CV training data
# -----------------------------

train_data_eo <- EcoliFreeRegionalProportion_with_covariates_enriched %>%
  dplyr::mutate(
    country_eval = as.character(country_cov),
    country_fold = as.factor(as.integer(factor(country_eval)))
  ) %>%
  dplyr::select(
    dplyr::all_of(eo_features),
    dplyr::all_of(outcome_col),
    dplyr::all_of(weight_col),
    country_eval,
    country_fold
  ) %>%
  tidyr::drop_na(
    dplyr::all_of(outcome_col),
    dplyr::all_of(weight_col),
    country_eval
  )

train_data_eo %>%
  dplyr::count(country_eval, country_fold) %>%
  dplyr::arrange(country_eval)
```

```{r}
# -----------------------------
# 3. Run default RF with leave-one-country-out CV
# -----------------------------

train_h2o_eo <- h2o::as.h2o(
  train_data_eo,
  destination_frame = "train_h2o_eo"
)

rf_loco_eo <- h2o::h2o.randomForest(
  x = eo_features,
  y = outcome_col,
  training_frame = train_h2o_eo,
  ntrees = 500,
  mtries = -1,
  sample_rate = 0.632,
  fold_column = "country_fold",
  weights_column = weight_col,
  seed = 123,
  keep_cross_validation_predictions = TRUE,
  keep_cross_validation_fold_assignment = TRUE
)
```
```{r}
# -----------------------------
# 4. Overall LOCO-CV metrics
# -----------------------------

loco_metrics_eo <- tibble::tibble(
  model = "Default RF - all EO features - LOCO CV",
  n_rows = nrow(train_data_eo),
  n_countries = dplyr::n_distinct(train_data_eo$country_eval),
  n_features = length(eo_features),
  r2 = as.numeric(h2o::h2o.r2(rf_loco_eo, xval = TRUE)),
  mae = as.numeric(h2o::h2o.mae(rf_loco_eo, xval = TRUE)),
  rmse = as.numeric(h2o::h2o.rmse(rf_loco_eo, xval = TRUE))
)

loco_metrics_eo
```


```{r}
# -----------------------------
# 5. Country influence on R2
# -----------------------------

cv_pred_id <- rf_loco_eo@model$cross_validation_holdout_predictions_frame_id$name

cv_predictions_eo <- h2o::h2o.getFrame(cv_pred_id) %>%
  as.data.frame()

cv_eval_eo <- train_data_eo
cv_eval_eo$pred <- cv_predictions_eo$predict
cv_eval_eo$actual <- cv_eval_eo[[outcome_col]]
cv_eval_eo$weight <- cv_eval_eo[[weight_col]]

weighted_r2 <- function(actual, pred, weight) {
  keep <- !is.na(actual) & !is.na(pred) & !is.na(weight)
  actual <- actual[keep]
  pred <- pred[keep]
  weight <- weight[keep]
  actual_mean <- weighted.mean(actual, weight)
  1 - sum(weight * (actual - pred)^2) / sum(weight * (actual - actual_mean)^2)
}

overall_r2_eo <- weighted_r2(cv_eval_eo$actual, cv_eval_eo$pred, cv_eval_eo$weight)

country_influence_eo <- purrr::map_dfr(
  sort(unique(cv_eval_eo$country_eval)),
  function(cty) {
    df_without <- cv_eval_eo %>%
      dplyr::filter(country_eval != cty)
    
    tibble::tibble(
      country_eval = cty,
      n_rows_country = sum(cv_eval_eo$country_eval == cty),
      r2_all = overall_r2_eo,
      r2_without_country = weighted_r2(df_without$actual, df_without$pred, df_without$weight),
      r2_change_when_removed = r2_without_country - r2_all
    )
  }
)

country_influence_eo %>%
  dplyr::arrange(dplyr::desc(r2_change_when_removed))
```

```{r}
# Extract CV predictions
cv_predictions_eo <- h2o::h2o.getFrame(
  rf_loco_eo@model$cross_validation_holdout_predictions_frame_id$name
) %>%
  as.data.frame()

# Create plot data
plot_data_cv_eo <- train_data_eo %>%
  dplyr::mutate(
    predict = cv_predictions_eo$predict,
    observed = .data[[outcome_col]],
    nr_HH_in_region = .data[[weight_col]]
  ) %>%
  dplyr::left_join(
    EcoliFreeRegionalProportion_with_covariates_enriched %>%
      dplyr::distinct(country_cov, HH7_region_cov, Income.group., UN.region),
    by = c("country_eval" = "country_cov")
  )

# Plot
colours <- c("#E64B35FF", "#91D1C2FF", "#3C5488FF", "#4DBBD5FF", "#EFC00F")

plotPredObs_ecoli_eo_cv <- ggplot(
  plot_data_cv_eo,
  aes(
    x = predict,
    y = observed,
    size = nr_HH_in_region,
    colour = UN.region
  )
) +
  geom_point(shape = 1, alpha = 0.8) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  stat_smooth(se = TRUE, colour = "black", size = 0.5, method = "lm") +
  scale_colour_manual(values = colours) +
  labs(
    x = "Predicted proportion of population with E. coli-free drinking water",
    y = "Observed proportion of population with E. coli-free drinking water",
    size = "Households in region (n)",
    colour = "UN region"
  ) +
  theme_minimal(base_size = 9)

plotPredObs_ecoli_eo_cv
```


### 2. Selecting test-data set

```{r}

df_training_final_wq <- EcoliFreeRegionalProportion_with_covariates %>%
  filter(!is.na(No_EcoliAtRegionalLevel))

country_key <- df.water_quality_overview %>%
  transmute(
    country_key = country_match_key(Country),
    income_group = Income.group.,
    UN_region = UN.region
  ) %>%
  distinct(country_key, .keep_all = TRUE)

df_training_final_wq <- df_training_final_wq %>%
  mutate(country_key = country_match_key(country_cov)) %>%
  left_join(country_key, by = "country_key") %>%
  select(-country_key)
```


```{r}
set.seed(12)

# Use only rows with outcome
df_model <- df_training_final_wq  %>%
  filter(!is.na(No_EcoliAtRegionalLevel))

# Change these names if your columns are named differently
region_col <- "UN_region"
income_col <- "income_group"

country_meta <- df_model %>%
  distinct(
    country.x,
    UN_region = .data[[region_col]],
    income_group = .data[[income_col]]
  )

# Target distributions from prediction set
target_region <- tibble(
  UN_region = c("Africa", "Americas", "Asia", "Europe", "Oceania"),
  target = c(0.4140625, 0.171875, 0.3359375, 0, 0.078125)
)

target_income <- tibble(
  income_group = c("Low income", "Lower middle income", "Upper middle income"),
  target = c(0.1953125, 0.390625, 0.4140625)
)

n_countries <- n_distinct(country_meta$country.x)
test_sizes <- floor(0.10 * n_countries):ceiling(0.17 * n_countries)

score_sample <- function(test_countries) {
  
  test_meta <- country_meta %>%
    filter(country.x %in% test_countries)
  
  region_score <- test_meta %>%
    count(UN_region) %>%
    mutate(prop = n / sum(n)) %>%
    right_join(target_region, by = "UN_region") %>%
    mutate(prop = replace_na(prop, 0)) %>%
    summarise(score = sum(abs(prop - target))) %>%
    pull(score)
  
  income_score <- test_meta %>%
    count(income_group) %>%
    mutate(prop = n / sum(n)) %>%
    right_join(target_income, by = "income_group") %>%
    mutate(prop = replace_na(prop, 0)) %>%
    summarise(score = sum(abs(prop - target))) %>%
    pull(score)
  
  region_score + income_score
}

# Try many random country samples and keep the best one
candidate_samples <- map_dfr(1:10000, function(i) {
  
  size <- sample(test_sizes, 1)
  countries <- sample(country_meta$country.x, size)
  
  tibble(
    iteration = i,
    n_test_countries = size,
    score = score_sample(countries),
    countries = list(countries)
  )
})

best_sample <- candidate_samples %>%
  arrange(score) %>%
  slice(1)

test_countries <- best_sample$countries[[1]]

test_set <- df_model %>%
  filter(country.x %in% test_countries)

training_set <- df_model %>%
  filter(!country.x %in% test_countries)
```

```{r}
test_countries

country_meta %>%
  mutate(set = if_else(country.x %in% test_countries, "Test", "Training")) %>%
  count(set, UN_region) %>%
  group_by(set) %>%
  mutate(prop = n / sum(n))

country_meta %>%
  mutate(set = if_else(country.x %in% test_countries, "Test", "Training")) %>%
  count(set, income_group) %>%
  group_by(set) %>%
  mutate(prop = n / sum(n))
```

```{r}
#test_countries <- c(
  #"Lao People's Democratic Republic",
  #"Tuvalu",
  #"Ghana",
  #"Paraguay",
  #"Guinea-Bissau"
#)

test_countries <- c(
"Guinea Bissau",
"Dominican Republic",
"Sao Tome and Principe",   
  "Georgia",
  "Central African Republic",
  "Zimbabwe",
  "Lao PDR"
)

test_country_keys <- country_match_key(test_countries)

# split observed WQ data into train/test countries
observed_split <- df_training_final_wq %>%
  mutate(
    country_key = country_match_key(country.x),
    dataset = if_else(country_key %in% test_country_keys, "Test", "Training")
  )

# -----------------------------
# Country-level distributions
# -----------------------------

country_region_compare_split <- bind_rows(
  observed_split %>%
    distinct(country.x, UN_region, dataset) %>%
    count(dataset, UN_region, name = "n"),
  
  features_pred_enriched %>%
    distinct(country, un_region) %>%
    count(un_region, name = "n") %>%
    rename(UN_region = un_region) %>%
    mutate(dataset = "Prediction")
)

country_income_compare_split <- bind_rows(
  observed_split %>%
    distinct(country.x, income_group, dataset) %>%
    count(dataset, income_group, name = "n"),
  
  features_pred_enriched %>%
    distinct(country, income_group) %>%
    count(income_group, name = "n") %>%
    mutate(dataset = "Prediction")
)

plot_prop <- function(df, xvar, title_text) {
  df %>%
    group_by(dataset) %>%
    mutate(prop = n / sum(n)) %>%
    ungroup() %>%
    ggplot(aes(x = {{ xvar }}, y = prop, fill = dataset)) +
    geom_col(position = "dodge") +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(x = "", y = "Proportion", fill = "Dataset", title = title_text) +
    theme_minimal()
}

plot_prop(country_region_compare_split, UN_region,
          "Country-level distribution across UN regions")

plot_prop(country_income_compare_split, income_group,
          "Country-level distribution across income groups")

```

```{r}
plot_count <- function(df, xvar, title_text) {
  df %>%
    ggplot(aes(x = {{ xvar }}, y = n, fill = dataset)) +
    geom_col(position = position_dodge(width = 0.9)) +
    geom_text(
      aes(label = n),
      position = position_dodge(width = 0.9),
      vjust = -0.3,
      size = 3
    ) +
    labs(x = "", y = "Number of countries", fill = "Dataset", title = title_text) +
    theme_minimal()
}

plot_count(country_region_compare_split, UN_region,
           "Country-level distribution across UN regions")

plot_count(country_income_compare_split, income_group,
           "Country-level distribution across income groups")
```

```{r}
test_country_keys <- country_match_key(test_countries)

test_set <- df_training_final_wq %>%
  mutate(country_key = country_match_key(country.x)) %>%
  filter(country_key %in% test_country_keys) %>%
  select(-country_key)

training_set <- df_training_final_wq %>%
  mutate(country_key = country_match_key(country.x)) %>%
  filter(!country_key %in% test_country_keys) %>%
  select(-country_key)


dir.create(here("outputs"), showWarnings = FALSE)

write.csv(
  training_set,
  here("outputs/03_sampled_features", "training_set.csv"),
  fileEncoding = "UTF-8",
  row.names = FALSE
)

write.csv(
  test_set,
  here("outputs/03_sampled_features", "test_set.csv"),
  fileEncoding = "UTF-8",
  row.names = FALSE
)
```

```{r}
#join training and test data sets (including environmental covariates) with WGI data

# ------------------------------------------------------------
# Join selected Worldwide Governance Indicators (WGI)
# to training/test using country and analysis year
# ------------------------------------------------------------

indicator_map <- c(
  "Government Effectiveness - Governance score (0-100)" = "governance_effectiveness",
  "Control of Corruption - Governance score (0-100)"    = "control_of_corruption",
  "Political Stability - Governance score (0-100)"      = "political_stability",
  "Regulatory Quality - Governance score (0-100)"       = "regulatory_quality",
  "Rule of Law - Governance score (0-100)"              = "rule_of_law",
  "Voice and Accountability - Governance score (0-100)" = "voice_and_accountability"
)

# Keep only the selected WGI indicators and reshape to one row per country-year
wgi_selected <- df.WGI_lmics %>%
  mutate(
    country_key = country_match_key(Country.Name),
    Series.Name = str_squish(as.character(Series.Name))
  ) %>%
  filter(Series.Name %in% names(indicator_map)) %>%
  mutate(
    indicator = recode(Series.Name, !!!indicator_map)
  ) %>%
  pivot_longer(
    cols = starts_with("X20"),
    names_to = "analysis_year",
    values_to = "wgi_value"
  ) %>%
  mutate(
    analysis_year = as.integer(str_remove(analysis_year, "^X")),
    wgi_value = na_if(wgi_value, ".."),
    wgi_value = as.numeric(wgi_value)
  ) %>%
  select(country_key, analysis_year, indicator, wgi_value) %>%
  distinct() %>%
  pivot_wider(
    names_from = indicator,
    values_from = wgi_value
  )

# ------------------------------------------------------------
# Join selected WGI indicators to training set
# ------------------------------------------------------------

training_set <- training_set %>%
  mutate(
    country_key = country_match_key(country.x)
  ) %>%
  left_join(
    wgi_selected,
    by = c("country_key", "analysis_year")
  ) %>%
  select(-country_key)

# ------------------------------------------------------------
# Join selected WGI indicators to test set
# ------------------------------------------------------------

test_set <- test_set %>%
  mutate(
    country_key = country_match_key(country.x)
  ) %>%
  left_join(
    wgi_selected,
    by = c("country_key", "analysis_year")
  ) %>%
  select(-country_key)

# ------------------------------------------------------------
# Optional check:
# country-years with no selected WGI data after join
# ------------------------------------------------------------

wgi_vars <- unname(indicator_map)

training_set %>%
  filter(if_all(all_of(wgi_vars), is.na)) %>%
  distinct(country.x, analysis_year) %>%
  arrange(country.x, analysis_year)

test_set %>%
  filter(if_all(all_of(wgi_vars), is.na)) %>%
  distinct(country.x, analysis_year) %>%
  arrange(country.x, analysis_year)
```

```{r}

combined_MICS_envir_human_e.coli  <- training_set %>%
  select(all_of("HH7_region"),all_of("country.x"), all_of(bandNames_39), all_of(new_features), all_of("No_EcoliAtRegionalLevel"),all_of("HouseholdsInRegion.Freq.x")) 

combined_MICS_envir_human_e.coli %>%
  distinct(country.x) %>%
  arrange(country.x)

combined_MICS_envir_human_e.coli <- createColumnForCVfoldBasedOnCountries(
  combined_MICS_envir_human_e.coli,
  country_var = "country.x",
  countries_per_fold = 3,
  seed = 123
)

combined_MICS_envir_human_e.coli %>%
  distinct(country.x, country_fold) %>%
  arrange(country_fold, country.x)
```

## Modelling E.coli contamination using random forest models

```{r}
k_values <- c(7, 12, 17, 25, 45, 60)

# Function to read feature list
read_feature_list <- function(path) {
  read_csv(path, show_col_types = FALSE) %>%
    pull(representative_variable)
}

# Read sampled feature lists
sampled_features <- setNames(
  lapply(k_values, function(k) {
    read_feature_list(
      here::here(
        "outputs",
        "03_sampled_features",
        paste0("representative_variables_k", k, ".csv")
      )
    )
  }),
  paste0("k", k_values)
)

# Variables that should always be included
ecoli_base_vars <- c(
  "HH7_region",
  "country.x",
  "No_EcoliAtRegionalLevel",
  "HouseholdsInRegion.Freq.x"
)

# Function to create one modelling dataset
make_ecoli_dataset <- function(data, feature_names) {
  data %>%
    select(
      all_of(ecoli_base_vars),
      all_of(feature_names)
    )
}

# Training datasets for each k
ecoli_training_sets <- map(
  sampled_features,
  ~ make_ecoli_dataset(training_set, .x)
)

# Test datasets for each k
ecoli_test_sets <- map(
  sampled_features,
  ~ make_ecoli_dataset(test_set, .x)
)
```

## 3. Random forest modelling:

```{r}

# Outcome variable for E. coli model
varToModel_Ecoli <- "No_EcoliAtRegionalLevel"

# Weight column
weight_col <- "HouseholdsInRegion.Freq.x"

# Country column used for CV folds
country_col <- "country.x"

country_fold_lookup <- ecoli_training_sets$k7 %>%
  select(all_of(country_col)) %>%
  distinct() %>%
  createColumnForCVfoldBasedOnCountries(
    country_var = country_col,
    countries_per_fold = 1,
    seed = 123
  ) %>%
  select(all_of(country_col), country_fold)

country_fold_lookup %>%
  dplyr::count(country_fold)


```

```{r}
run_ecoli_rf_model <- function(k_name,
                               train_df,
                               test_df,
                               feature_names,
                               fold_lookup,
                               varToModel = "No_EcoliAtRegionalLevel",
                               weight_col = "HouseholdsInRegion.Freq.x",
                               country_col = "country.x",
                               ntrees = 500,
                               seed = 123) {
  
  # Add same country folds to training data
  train_df <- train_df %>%
    dplyr::left_join(fold_lookup, by = country_col)
  
  # Check required training columns
  required_train_cols <- c(feature_names, varToModel, weight_col, "country_fold")
  missing_train_cols <- setdiff(required_train_cols, names(train_df))
  
  if (length(missing_train_cols) > 0) {
    stop(
      "Missing columns in training data for ", k_name, ": ",
      paste(missing_train_cols, collapse = ", ")
    )
  }
  
  # Keep only variables needed for training
  train_reg_matrix <- train_df %>%
    dplyr::select(
      dplyr::all_of(feature_names),
      dplyr::all_of(varToModel),
      dplyr::all_of(weight_col),
      country_fold
    ) %>%
    tidyr::drop_na(country_fold, dplyr::all_of(varToModel), dplyr::all_of(weight_col))
  
  # Check required test columns
  required_test_cols <- c(feature_names, varToModel, weight_col)
  missing_test_cols <- setdiff(required_test_cols, names(test_df))
  
  if (length(missing_test_cols) > 0) {
    stop(
      "Missing columns in test data for ", k_name, ": ",
      paste(missing_test_cols, collapse = ", ")
    )
  }
  
  # Keep only variables needed for test set evaluation
  test_reg_matrix <- test_df %>%
    dplyr::select(
      dplyr::all_of(feature_names),
      dplyr::all_of(varToModel),
      dplyr::all_of(weight_col)
    ) %>%
    tidyr::drop_na(dplyr::all_of(varToModel), dplyr::all_of(weight_col))
  
  # Convert to H2O frames
  train_h2o <- h2o::as.h2o(
    train_reg_matrix,
    destination_frame = paste0("regMatrixH2O_Ecoli_train_", k_name)
  )
  
  test_h2o <- h2o::as.h2o(
    test_reg_matrix,
    destination_frame = paste0("regMatrixH2O_Ecoli_test_", k_name)
  )
  
  # Train random forest
  rf_model <- h2o::h2o.randomForest(
    x = feature_names,
    y = varToModel,
    training_frame = train_h2o,
    ntrees = ntrees,
    mtries = -1,
    sample_rate = 0.632,
    fold_column = "country_fold",
    weights_column = weight_col,
    seed = seed,
    fold_assignment = "AUTO",
    keep_cross_validation_predictions = TRUE,
    keep_cross_validation_fold_assignment = TRUE
  )
  
  # Cross-validation performance
  xval_r2 <- h2o::h2o.r2(rf_model, xval = TRUE)
  xval_mae <- h2o::h2o.mae(rf_model, xval = TRUE)
  
  # Test set performance
  test_perf <- h2o::h2o.performance(rf_model, newdata = test_h2o)
  test_r2 <- h2o::h2o.r2(test_perf)
  test_mae <- h2o::h2o.mae(test_perf)
  
  list(
    model = rf_model,
    train_h2o = train_h2o,
    test_h2o = test_h2o,
    test_performance = test_perf,
    metrics = tibble::tibble(
      k = k_name,
      n_features = length(feature_names),
      n_train = nrow(train_reg_matrix),
      n_test = nrow(test_reg_matrix),
      xval_r2 = as.numeric(xval_r2),
      xval_mae = as.numeric(xval_mae),
      test_r2 = as.numeric(test_r2),
      test_mae = as.numeric(test_mae)
    )
  )
}
```

```{r}
h2o.init(
  nthreads = 7,
  max_mem_size = "8g",
  ignore_config = TRUE
)
```

```{r}
rf_results_ecoli <- purrr::imap(
  sampled_features,
  function(feature_names, k_name) {
    run_ecoli_rf_model(
      k_name = k_name,
      train_df = ecoli_training_sets[[k_name]],
      test_df = ecoli_test_sets[[k_name]],
      feature_names = feature_names,
      fold_lookup = country_fold_lookup
    )
  }
)
```

```{r}
rf_metrics_ecoli <- purrr::map_dfr(
  rf_results_ecoli,
  "metrics"
)

rf_metrics_ecoli
```

## Grid-search: Tuning hyperparameters of random forest model with 39

features Outcome: access to E.coli free water

We performed a grid-search to iteratively explore the results of a suite
of random forest models trained on the selected covariates. A list of
hyperparameters was selected, varying in (1) the number of trees, (2)
the number of variables sampled at each split and (3) the minimum
observations per leaf and (4)sampling rate

```{r}
best_k <- "k60"

features_k60 <- sampled_features[[best_k]]
train_k60 <- ecoli_training_sets[[best_k]]
test_k60  <- ecoli_test_sets[[best_k]]

varToModel_Ecoli <- "No_EcoliAtRegionalLevel"
weight_col <- "HouseholdsInRegion.Freq.x"
country_col <- "country.x"

train_k60_model <- train_k60 %>%
  dplyr::left_join(country_fold_lookup, by = country_col) %>%
  dplyr::select(
    dplyr::all_of(features_k60),
    dplyr::all_of(varToModel_Ecoli),
    dplyr::all_of(weight_col),
    country_fold
  ) %>%
  tidyr::drop_na(country_fold, dplyr::all_of(varToModel_Ecoli), dplyr::all_of(weight_col))

test_k60_model <- test_k60 %>%
  dplyr::select(
    dplyr::all_of(features_k60),
    dplyr::all_of(varToModel_Ecoli),
    dplyr::all_of(weight_col)
  ) %>%
  tidyr::drop_na(dplyr::all_of(varToModel_Ecoli), dplyr::all_of(weight_col))

h2o.init(
  nthreads = 7,
  max_mem_size = "8g",
  ignore_config = TRUE
)

train_h2o_k60 <- h2o::as.h2o(train_k60_model, destination_frame = "train_h2o_k60")
test_h2o_k60  <- h2o::as.h2o(test_k60_model, destination_frame = "test_h2o_k60")
```

```{r}
# 1. Clear H2O memory
h2o::h2o.removeAll()
gc()

# 2. Recreate H2O frames
train_h2o_k60 <- h2o::as.h2o(train_k60_model, destination_frame = "train_h2o_k60")
test_h2o_k60  <- h2o::as.h2o(test_k60_model, destination_frame = "test_h2o_k60")

# 3. Run RF tuning grid
rf_hyper_params <- list(
  mtries = c(-1, floor(sqrt(length(features_k60))), floor(length(features_k60) / 3)),
  min_rows = c(1, 5, 10, 20),
  sample_rate = c(0.632, 0.8, 1.0),
  max_depth = c(10, 20, 30, 0)
)

rf_grid <- h2o::h2o.grid(
  algorithm = "randomForest",
  grid_id = "rf_grid_ecoli_k60",
  x = features_k60,
  y = varToModel_Ecoli,
  training_frame = train_h2o_k60,
  weights_column = weight_col,
  fold_column = "country_fold",
  ntrees = 500,
  seed = 123,
  hyper_params = rf_hyper_params,
  search_criteria = list(
    strategy = "RandomDiscrete",
    max_models = 25,
    seed = 123
  ),
  keep_cross_validation_predictions = TRUE,
  keep_cross_validation_fold_assignment = TRUE
)

# 4. Sort grid by cross-validation MAE
rf_grid_perf <- h2o::h2o.getGrid(
  grid_id = "rf_grid_ecoli_k60",
  sort_by = "mae",
  decreasing = FALSE
)

rf_grid_perf
```

## 4. Model comparison with countries incl. in Greenwood et al 2024

```{r}

# -----------------------------
# 1. Setup
# -----------------------------

features_k60 <- sampled_features$k60

country_col <- "country.x"
outcome_col <- "No_EcoliAtRegionalLevel"
weight_col  <- "HouseholdsInRegion.Freq.x"

old_countries <- c(
  "Algeria", "Bangladesh", "Central African Republic", "Chad", "Gambia",
  "Georgia", "Ghana", "Guinea-Bissau", "Guyana", "Iraq", "Kiribati",
  "Kosovo (under UNSC res. 1244)", "Lesotho", "Madagascar", "Mongolia",
  "Nigeria", "Pakistan", "Paraguay", "São Tome and Principe",
  "Sierra Leone", "Suriname", "Togo", "Tonga", "West Bank and Gaza",
  "Zimbabwe"
)

standardise_country_names <- function(df) {
  df %>%
    dplyr::mutate(
      country_raw = as.character(country.x),
      country_eval = dplyr::case_when(
        country_raw %in% c("Sao Tome and Principe", "São Tome and Principe") ~ "São Tome and Principe",
        country_raw %in% c("Palestina", "Palestine", "West Bank and Gaza") ~ "West Bank and Gaza",
        country_raw %in% c("The Gambia", "Gambia") ~ "Gambia",
        country_raw %in% c("Kosovo", "Kosovo (under UNSC res. 1244)") ~ "Kosovo (under UNSC res. 1244)",
        TRUE ~ country_raw
      )
    ) %>%
    dplyr::select(-country_raw)
}
```

```{r}
# -----------------------------
# 2. Combine train + test, keep old countries
# -----------------------------

old_country_data_k60 <- bind_rows(
  ecoli_training_sets$k60,
  ecoli_test_sets$k60
) %>%
  standardise_country_names() %>%
  filter(country_eval %in% old_countries) %>%
  select(
    all_of(features_k60),
    all_of(outcome_col),
    all_of(weight_col),
    country_eval
  ) %>%
  drop_na(all_of(outcome_col), all_of(weight_col), country_eval) %>%
  mutate(
    country_fold = as.factor(as.integer(factor(country_eval)))
  )

# Check countries included
old_country_data_k60 %>%
  dplyr::count(country_eval, country_fold) %>%
  dplyr::arrange(country_eval)
```

```{r}
# Clean restart of H2O
try(h2o::h2o.shutdown(prompt = FALSE), silent = TRUE)

gc()

h2o::h2o.init(
  nthreads = 7,
  max_mem_size = "8G",
  ignore_config = TRUE
)

# Check H2O cluster
h2o::h2o.clusterInfo()

old_country_h2o <- h2o::as.h2o(
  old_country_data_k60,
  destination_frame = "old_country_h2o_k60"
)

rf_loco_old_k60 <- h2o::h2o.randomForest(
  x = features_k60,
  y = outcome_col,
  training_frame = old_country_h2o,
  ntrees = 500,
  mtries = -1,
  sample_rate = 0.632,
  fold_column = "country_fold",
  weights_column = weight_col,
  seed = 123,
  keep_cross_validation_predictions = TRUE,
  keep_cross_validation_fold_assignment = TRUE
)
```

```{r}
loco_metrics_old_k60 <- tibble::tibble(
  model = "Default RF k60 - old countries LOCO CV",
  n_rows = nrow(old_country_data_k60),
  n_countries = dplyr::n_distinct(old_country_data_k60$country_eval),
  r2 = as.numeric(h2o::h2o.r2(rf_loco_old_k60, xval = TRUE)),
  mae = as.numeric(h2o::h2o.mae(rf_loco_old_k60, xval = TRUE)),
  rmse = as.numeric(h2o::h2o.rmse(rf_loco_old_k60, xval = TRUE))
)

loco_metrics_old_k60
```

## 5. Evaluate country performance based on cross-validation metrics

```{r}
# ============================================================
# Leave-one-country-out CV on current/new training data, k60
# ============================================================

features_k60 <- sampled_features$k60

country_col <- "country.x"
outcome_col <- "No_EcoliAtRegionalLevel"
weight_col  <- "HouseholdsInRegion.Freq.x"

new_train_data_k60 <- ecoli_training_sets$k60 %>%
  dplyr::mutate(
    country_eval = as.character(country.x),
    country_fold = as.factor(as.integer(factor(country_eval)))
  ) %>%
  dplyr::select(
    dplyr::all_of(features_k60),
    dplyr::all_of(outcome_col),
    dplyr::all_of(weight_col),
    country_eval,
    country_fold
  ) %>%
  tidyr::drop_na(
    dplyr::all_of(outcome_col),
    dplyr::all_of(weight_col),
    country_eval
  )

new_train_data_k60 %>%
  dplyr::count(country_eval, country_fold) %>%
  dplyr::arrange(country_eval)
```

```{r}
new_train_h2o_k60 <- h2o::as.h2o(
  new_train_data_k60,
  destination_frame = "new_train_h2o_k60"
)

rf_loco_new_k60 <- h2o::h2o.randomForest(
  x = features_k60,
  y = outcome_col,
  training_frame = new_train_h2o_k60,
  ntrees = 500,
  mtries = -1,
  sample_rate = 0.632,
  fold_column = "country_fold",
  weights_column = weight_col,
  seed = 123,
  keep_cross_validation_predictions = TRUE,
  keep_cross_validation_fold_assignment = TRUE
)
```

```{r}
loco_metrics_new_k60 <- tibble::tibble(
  model = "Default RF k60 - new training countries LOCO CV",
  n_rows = nrow(new_train_data_k60),
  n_countries = dplyr::n_distinct(new_train_data_k60$country_eval),
  r2 = as.numeric(h2o::h2o.r2(rf_loco_new_k60, xval = TRUE)),
  mae = as.numeric(h2o::h2o.mae(rf_loco_new_k60, xval = TRUE)),
  rmse = as.numeric(h2o::h2o.rmse(rf_loco_new_k60, xval = TRUE))
)

loco_metrics_new_k60
```

```{r}
cv_pred_id <- rf_loco_new_k60@model$cross_validation_holdout_predictions_frame_id$name

cv_predictions_new <- h2o::h2o.getFrame(cv_pred_id) %>%
  as.data.frame()

new_cv_eval <- new_train_data_k60 %>%
  dplyr::mutate(
    pred = cv_predictions_new$predict,
    actual = .data[[outcome_col]],
    weight = .data[[weight_col]]
  )
```

```{r}
weighted_r2 <- function(actual, pred, weight) {
  keep <- !is.na(actual) & !is.na(pred) & !is.na(weight)
  actual <- actual[keep]
  pred <- pred[keep]
  weight <- weight[keep]
  
  actual_mean <- weighted.mean(actual, weight)
  sse <- sum(weight * (actual - pred)^2)
  sst <- sum(weight * (actual - actual_mean)^2)
  
  1 - sse / sst
}

calc_r2 <- function(df) {
  weighted_r2(df$actual, df$pred, df$weight)
}
```

```{r}
overall_r2_new <- calc_r2(new_cv_eval)

country_influence_new <- purrr::map_dfr(
  sort(unique(new_cv_eval$country_eval)),
  function(country_i) {
    
    r2_without_country <- new_cv_eval %>%
      dplyr::filter(country_eval != country_i) %>%
      calc_r2()
    
    tibble::tibble(
      country_eval = country_i,
      n_rows_country = sum(new_cv_eval$country_eval == country_i),
      r2_all = overall_r2_new,
      r2_without_country = r2_without_country,
      r2_change_when_removed = r2_without_country - overall_r2_new
    )
  }
)

country_influence_new %>%
  dplyr::arrange(dplyr::desc(r2_change_when_removed))
```

```{r}
# These countries hurt performance:
# removing them increases R2
country_influence_new %>%
  dplyr::filter(r2_change_when_removed > 0) %>%
  dplyr::arrange(dplyr::desc(r2_change_when_removed))

# These countries help performance:
# removing them decreases R2
country_influence_new %>%
  dplyr::filter(r2_change_when_removed < 0) %>%
  dplyr::arrange(r2_change_when_removed)
```

```{r}
# Remove Pakistan surveys from k60 training data
ecoli_training_sets_no_pakistan <- ecoli_training_sets

ecoli_training_sets_no_pakistan$k60 <- ecoli_training_sets$k60 %>%
  dplyr::filter(
    !grepl("^Pakistan", country.x)
  )

new_train_data_k60_no_pakistan <- ecoli_training_sets_no_pakistan$k60 %>%
  dplyr::mutate(
    country_eval = as.character(country.x),
    country_fold = as.factor(as.integer(factor(country_eval)))
  ) %>%
  dplyr::select(
    dplyr::all_of(features_k60),
    dplyr::all_of(outcome_col),
    dplyr::all_of(weight_col),
    country_eval,
    country_fold
  ) %>%
  tidyr::drop_na(
    dplyr::all_of(outcome_col),
    dplyr::all_of(weight_col),
    country_eval
  )
```

```{r}
h2o::h2o.removeAll()

new_train_h2o_k60_no_pakistan <- h2o::as.h2o(
  new_train_data_k60_no_pakistan,
  destination_frame = "new_train_h2o_k60_no_pakistan"
)

rf_loco_new_k60_no_pakistan <- h2o::h2o.randomForest(
  x = features_k60,
  y = outcome_col,
  training_frame = new_train_h2o_k60_no_pakistan,
  ntrees = 500,
  mtries = -1,
  sample_rate = 0.632,
  fold_column = "country_fold",
  weights_column = weight_col,
  seed = 123,
  keep_cross_validation_predictions = TRUE,
  keep_cross_validation_fold_assignment = TRUE
)

loco_metrics_new_k60_no_pakistan <- tibble::tibble(
  model = "Default RF k60 - no Pakistan surveys LOCO CV",
  n_rows = nrow(new_train_data_k60_no_pakistan),
  n_countries = dplyr::n_distinct(new_train_data_k60_no_pakistan$country_eval),
  r2 = as.numeric(h2o::h2o.r2(rf_loco_new_k60_no_pakistan, xval = TRUE)),
  mae = as.numeric(h2o::h2o.mae(rf_loco_new_k60_no_pakistan, xval = TRUE)),
  rmse = as.numeric(h2o::h2o.rmse(rf_loco_new_k60_no_pakistan, xval = TRUE))
)

loco_metrics_new_k60_no_pakistan
```

```{r}
# Extract CV predictions
cv_pred_id <- rf_loco_new_k60_no_pakistan@model$cross_validation_holdout_predictions_frame_id$name

cv_predictions_no_pakistan <- h2o::h2o.getFrame(cv_pred_id) %>%
  as.data.frame()

# Combine observed + predicted
plot_data_ecoli_no_pakistan <- new_train_data_k60_no_pakistan

plot_data_ecoli_no_pakistan$predict <- cv_predictions_no_pakistan$predict
plot_data_ecoli_no_pakistan$train <- plot_data_ecoli_no_pakistan[[outcome_col]]
plot_data_ecoli_no_pakistan$nr_HH_in_region <- plot_data_ecoli_no_pakistan[[weight_col]]
```

```{r}
plotPredObs_ecoli_no_pakistan <- ggplot(
  plot_data_ecoli_no_pakistan,
  aes(
    x = predict,
    y = train,
    size = nr_HH_in_region,
    colour = country_eval
  )
) +
  geom_point(shape = 1, alpha = 0.8) +
  stat_smooth(
    se = TRUE,
    colour = "black",
    size = 0.5,
    method = "lm"
  ) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed"
  ) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    x = "Predicted proportion of population with E. coli-free drinking water",
    y = "Observed proportion of population with E. coli-free drinking water",
    size = "Households in region (n)",
    colour = "Country"
  ) +
  theme_bw(base_size = 9)

plotPredObs_ecoli_no_pakistan
```

```{r}

# Prepare test data
test_data_k60_no_pakistan <- ecoli_test_sets$k60 %>%
  dplyr::filter(!grepl("^Pakistan", country.x)) %>%
  dplyr::mutate(
    country_eval = as.character(country.x)
  ) %>%
  dplyr::select(
    dplyr::all_of(features_k60),
    dplyr::all_of(outcome_col),
    dplyr::all_of(weight_col),
    country_eval
  ) %>%
  tidyr::drop_na(
    dplyr::all_of(outcome_col),
    dplyr::all_of(weight_col)
  )

# Convert test data to H2O
test_h2o_k60_no_pakistan <- h2o::as.h2o(
  test_data_k60_no_pakistan,
  destination_frame = "test_h2o_k60_no_pakistan"
)

# Predict on test set
test_predictions_no_pakistan <- h2o::h2o.predict(
  rf_loco_new_k60_no_pakistan,
  test_h2o_k60_no_pakistan
) %>%
  as.data.frame()
```

```{r}
# Combine observed + predicted
plot_data_test_ecoli_no_pakistan <- test_data_k60_no_pakistan

plot_data_test_ecoli_no_pakistan$predict <- test_predictions_no_pakistan$predict
plot_data_test_ecoli_no_pakistan$test <- plot_data_test_ecoli_no_pakistan[[outcome_col]]
plot_data_test_ecoli_no_pakistan$nr_HH_in_region <- plot_data_test_ecoli_no_pakistan[[weight_col]]
```

```{r}
# Plot observed vs predicted for test set
plotPredObs_ecoli_test_no_pakistan <- ggplot(
  plot_data_test_ecoli_no_pakistan,
  aes(
    x = predict,
    y = test,
    size = nr_HH_in_region,
    colour = country_eval
  )
) +
  geom_point(shape = 1, alpha = 0.8) +
  stat_smooth(
    se = TRUE,
    colour = "black",
    size = 0.5,
    method = "lm"
  ) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed"
  ) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    x = "Predicted proportion of population with E. coli-free drinking water",
    y = "Observed proportion of population with E. coli-free drinking water",
    size = "Households in region (n)",
    colour = "Country"
  ) +
  theme_bw(base_size = 9)

plotPredObs_ecoli_test_no_pakistan
```

```{r}
test_perf_no_pakistan <- h2o::h2o.performance(
  rf_loco_new_k60_no_pakistan,
  newdata = test_h2o_k60_no_pakistan
)

h2o::h2o.r2(test_perf_no_pakistan)
h2o::h2o.mae(test_perf_no_pakistan)
h2o::h2o.rmse(test_perf_no_pakistan)
```

```{r}
h2o.init(nthreads = 7, max_mem_size = '8g', ignore_config = TRUE)

## modelling E.coli
varToModel_e.coli <-'No_EcoliAtRegionalLevel'

combined_MICS_envir_human_e.coli  <-createColumnForCVfoldBasedOnCountries(combined_MICS_envir_human_e.coli) 

combined_MICS_envir_human_e.coli <-combined_MICS_envir_human_e.coli %>% 
  drop_na(country_fold)

df.MICS_envir_human_e.coli_39 <- combined_MICS_envir_human_e.coli  %>% 
  select(all_of(bandNames_39),varToModel_e.coli,country_fold, HouseholdsInRegion.Freq.x) 

#write.csv(df.MICS_envir_human_e.coli_39, "./5_Predictions/prediction_results/Regre_Matrix_Ecoli.csv", fileEncoding = "UTF-8", row.names = F)

regMatrixH2O_e.coli_reg_39 <- as.h2o(df.MICS_envir_human_e.coli_39, destination_frame = "regMatrixH2O_e.coli_reg_39")

##grid search
rf.params_39_e.coli<-list(ntrees = c(100,200),
                mtries = c(6,10,13,20), #(about n^(1/2), 1/4, 1/3, 1/2 of number of features)
                min_rows = c(1,3,5),
                sample_rate = c(0.6,0.8)
                )

search.criteria<-list(strategy = 'RandomDiscrete', max_models = 150, seed = 0, max_runtime_secs = 3600)

rf.grid_39_e.coli<-h2o.grid('randomForest',
                  y = varToModel_e.coli,
                  grid_id = 'rf.grid_e.coli',
                  training_frame = regMatrixH2O_e.coli_reg_39,
                  seed = 0,
                  hyper_params = rf.params_39_e.coli,
                  fold_column = "country_fold",
                  weights_column = "HouseholdsInRegion.Freq.x",
                  fold_assignment = 'AUTO',
                  keep_cross_validation_predictions = TRUE, # Important!
                  keep_cross_validation_fold_assignment = TRUE, # Important!
                  search_criteria = search.criteria)

summary(rf.grid_39_e.coli, show_stack_traces = TRUE)
#retrieve grid searched model performance sort by RMSE/R2

rf_grid_perf_ecoli<-h2o.getGrid(grid_id = 'rf.grid_e.coli',
                          sort_by = 'R2', # This is also coefficient of determinaion values

                                                    decreasing = TRUE)
summary(rf.grid_39_e.coli, show_stack_traces = TRUE)

print(rf_grid_perf_ecoli@summary_table)

#write.csv(rf_grid_perf_ecoli@summary_table,"./4_Training/results_grid_search/rf_grid_ecoli_reg_39.csv")
```

## best RF model with 39 features for E.coli contamination

```{r}
# random forest regression using metrics of best performing model
library(h2o)
rf_model_ecoli_final <- h2o.randomForest(
  y = varToModel_e.coli,
  training_frame = regMatrixH2O_e.coli_reg_39,
  ntrees = 100,
  mtries = 13,
  sample_rate = 0.6,
  min_rows = 3,
  fold_column = "country_fold",
  weights_column = "HouseholdsInRegion.Freq.x",
  seed = 0,
  fold_assignment = 'AUTO',
  keep_cross_validation_predictions = TRUE, # Important!
  keep_cross_validation_fold_assignment = TRUE)

h2o.r2(rf_model_ecoli_final, xval = TRUE)
h2o.mae(rf_model_ecoli_final, xval = TRUE)

h2o.varimp_plot(rf_model_ecoli_final)
h2o.varimp(rf_model_ecoli_final)

perf_40_ecoli <- h2o.performance(rf_model_ecoli_final)
perf_40_ecoli
```

Calculating and plotting SHAP values for E.coli predictions

```{r}
shap_values_ecoli <- predict_contributions.H2OModel(
  rf_model_ecoli_final,
  regMatrixH2O_e.coli_reg_39)

mean_values_ecoli <- as.data.frame(colMeans(abs(as.data.frame(shap_values_ecoli) %>% select(-BiasTerm)))); names(mean_values_ecoli) <- 'Mean'
mean_values_ecoli %>% arrange(desc(Mean))

mean_values_ecoli <- cbind(feature = rownames(mean_values_ecoli), mean_values_ecoli)

mean_values_ecoli  <- rename39Features(mean_values_ecoli)

mean_values_ecoli <- group39FeaturesInto5IndicatorGroups(mean_values_ecoli)

#write.csv(mean_values_ecoli, "./5_Prediction/prediction_results/Shap_values_ecoli.csv", fileEncoding = "UTF-8", row.names = T)

```

Plotting summarized average SHAP values by Indicator Group for E.coli
estimates

```{r}
SHAP_values_Ecoli_grouped <- mean_values_ecoli %>% 
  group_by(Indicator.Group) %>% 
  summarise(SumOfAverageSHAPValue = sum(Mean))

Plot_SHAP_Ecoli_grouped <- ggplot(SHAP_values_Ecoli_grouped, aes(x=reorder(Indicator.Group, SumOfAverageSHAPValue), y=SumOfAverageSHAPValue, fill= Indicator.Group)) +
  ggtitle("Absence of faecal contamination")+
  geom_bar(stat = "identity") + 
  labs(y = "sum of average SHAP values", x="") +
  scale_y_continuous(limits = c(0,0.150)) +
coord_flip() +
  scale_fill_manual(values=colours)+
  guides(fill = FALSE, color = FALSE, linetype = FALSE, shape = FALSE)+
  theme_bw()

Plot_SHAP_Ecoli_grouped
```

Create the SHAP summary plot

```{r}
shap_summary_plot_ecoli <- h2o.shap_summary_plot(rf_model_ecoli_final, regMatrixH2O_e.coli_reg_39)
print(shap_summary_plot_ecoli)

shap_summary_plot_data <- shap_summary_plot_ecoli[['data']]
shap_summary_plot_data$feature <- as.character(shap_summary_plot_data$feature)
shap_summary_plot_data <-rename39Features(shap_summary_plot_data)

shap_summary_plot_data <- shap_summary_plot_data %>%
  left_join(mean_values_ecoli, by = c("feature"="feature"))

library(viridis)
shap_summary_plot_data %>%
  mutate(feature = fct_reorder(feature, Mean)) %>%
ggplot(aes(feature, contribution, color = normalized_value, text = row)) +
        geom_hline(yintercept = 0, linetype = "dashed") +
        geom_point(position = h2o:::position_jitter_density(), alpha = 0.5, size = 1.2) +
        scale_fill_gradientn(colors = viridis_pal()(9), breaks=c(0,1),labels=c("low","high"), limits=c(0, 1)) +
        coord_flip() +
        labs(y = "SHAP contribution to E. coli absence in drinking water source prediction", x = "", col = "Normalized\nfeature value") +
        theme_bw(base_size = 9) 

ggsave(
  "Fig_S13.pdf",
  dpi=300,
  width= 6, height = 4
)
```

Plot results from best RF Model for E.coli contamination Plot as
predicted versus observed E.coli contamination with results from
leave-one-country-out holdout predictions.

```{r}
TrainAndPredicted_ecoli <- as.data.frame(h2o.cross_validation_holdout_predictions(rf_model_ecoli_final))

TrainAndPredicted_ecoli$train <- df.MICS_envir_human_e.coli_39$No_EcoliAtRegionalLevel

TrainAndPredicted_ecoli$nr_HH_in_region <- df.MICS_envir_human_e.coli_39$HouseholdsInRegion.Freq.x

TrainAndPredicted_ecoli$Country <- combined_MICS_envir_human_e.coli $NAME_0

source(here("./2_Data_cleaning/3_MICS_Preprocessing_Dataframes/1_Compiling_MICS_test_and_training_sets/functionsForJoiningAndStructuringDataframes.R"))

TrainAndPredicted_ecoli<- joinWithCountryIncomeGroups(TrainAndPredicted_ecoli,UN_Population_Devision_Names, country_Income_Groups)

GlobalRegions <- readAndRenameStandardGlobalRegionNames()

TrainAndPredicted_ecoli_withGlobalRegions <- TrainAndPredicted_ecoli %>% left_join(GlobalRegions, by = c("ISO3.Alpha.code"="ISO.alpha3.Code"))

TrainAndPredicted_ecoli_withGlobalRegions$UNRegionName[TrainAndPredicted_ecoli_withGlobalRegions$Country == "Kosovo (under UNSC res. 1244)"] <- "Europe"

# Plot predicted vs observed values

colours <- c("#E64B35FF","#91D1C2FF","#3C5488FF","#4DBBD5FF", "#EFC00F")
                                    
plotPredObs_ecoli <- ggplot(TrainAndPredicted_ecoli_withGlobalRegions, 
                      aes(x = predict, 
                          y = train, size = nr_HH_in_region, colour= UNRegionName)) +
  geom_point(shape=1) +
  labs(x = "Predicted proportion of population with E.coli free drinking water",
       y = "Observed proportion of population with E.coli free drinking water") +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1)) +
stat_smooth(se = T,
              colour = "black",
              size = 0.5,
              method = "lm") +
  geom_abline(slope = 1,
              intercept = 0,
              na.rm = FALSE, 
              show.legend = NA, 
              linetype = "dashed") +
  labs(size="Households in region (n)", 
              col="UN region") +
  scale_colour_manual(values = colours) +
  theme_bw(base_size = 9) 

plotPredObs_ecoli

ggsave(
  "Fig_S6.pdf",
  dpi=300,
  width= 6, height = 4
)
```

## E.coli prediction at global administrative (GADM) area level 1

```{r}
#level 1 prediction

EO_GADM_level1 <- read.csv(here("1_Data/EnvironmentalFeatures/FeaturesAtGADMlevel1.csv"), encoding = "UTF-8")

regressionMatrixForPrediction <-EO_GADM_level1 %>%  
 select(all_of(bandNames_39))
 
regMatrixH2O_4pred_39 <- as.h2o(regressionMatrixForPrediction, destination_frame = "regMatrixH2O_4pred_39")

FullPrediction_ecoli <- as.data.frame(h2o.predict(rf_model_ecoli_final, regMatrixH2O_4pred_39))

FullPrediction_ecoli$region <- EO_GADM_level1[["NAME_1"]]
FullPrediction_ecoli$GID_1 <- EO_GADM_level1[["GID_1"]]
FullPrediction_ecoli$country <- EO_GADM_level1[["NAME_0"]]
FullPrediction_ecoli$GID_0 <- EO_GADM_level1[["GID_0"]]

country_Income_Groups <- renameCountryCodeForKosovo(country_Income_Groups)

FullPrediction_ecoli <- FullPrediction_ecoli  %>% 
  left_join(country_Income_Groups, by=(c("GID_0"="Code")))

#Venezuela was not in the WB income group classification file so we add it as a lower-middle-income country according to the Inter-American Development Bank

FullPrediction_ecoli$Income.group[FullPrediction_ecoli$country == "Venezuela"] <- "Lower middle income"
FullPrediction_ecoli$Income.group[FullPrediction_ecoli$country == "Tokelau"] <- "Lower middle income"
FullPrediction_ecoli$Income.group[FullPrediction_ecoli$country == "Saint Helena"] <- "Upper middle income"

FullPrediction_ecoli <- FullPrediction_ecoli %>%
  na.omit(Income.group)

#write.csv(FullPrediction_ecoli, "./5_Predictions/prediction_results/Fullprediction_Ecoli_newest.csv", fileEncoding = "UTF-8", row.names = F)
```

## Descriptive analysis of E.coli results

```{r}
source(here("./2_Data_cleaning/3_MICS_Preprocessing_Dataframes/1_Compiling_MICS_test_and_training_sets/functionsForJoiningAndStructuringDataframes.R"))

Ecoli_originalprediction <- read.csv(here("./5_Predictions/prediction_results/Fullprediction_ecoli.csv"))

GlobalRegions <- readAndRenameStandardGlobalRegionNames()

#joining and preparing dataframe for analysis
Ecoli_originalprediction <- Ecoli_originalprediction %>% 
  left_join(EO_GADM_level1, by = c("GID_1"="GID_1", "GID_0"="GID_0")) %>% 
  left_join(GlobalRegions, by = c("GID_0"="ISO.alpha3.Code")) 

Ecoli_originalprediction$country[Ecoli_originalprediction$country == "Kosovo"] <- "Kosovo (under UNSC res. 1244)"
  
Ecoli_originalprediction$UNRegionName[Ecoli_originalprediction$country == "Kosovo (under UNSC res. 1244)"] <- "Europe"

Ecoli_originalprediction$UNSubRegionName[Ecoli_originalprediction$country == "Kosovo (under UNSC res. 1244)"] <- "Eastern Europe"


Ecoli_originalprediction <- Ecoli_originalprediction  %>% 
  filter(Income.group == "Low income" |
         Income.group == "Lower middle income" |
         Income.group == "Upper middle income")

```

## E.coli estimates of across all LMICs

```{r}
Ecoli_originalprediction$numberOfPeopleinRegion <- Ecoli_originalprediction$GPWv4_Population_2020
Ecoli_originalprediction$numberOfPeopleWithoutEcoliInRegion <- Ecoli_originalprediction$GPWv4_Population_2020*Ecoli_originalprediction$predict

Ecoli_originalprediction$totalNumberOfPeopleLMICS <- sum(Ecoli_originalprediction$numberOfPeopleinRegion)#6.523e+09
Ecoli_originalprediction$NumberOfPeopleLMICSwithoutEcoliSMDW <- sum(Ecoli_originalprediction$numberOfPeopleWithoutEcoliInRegion)#3420936496

Ecoli_originalprediction$numberOfPeopleEcoliInRegion <- Ecoli_originalprediction$numberOfPeopleinRegion-Ecoli_originalprediction$numberOfPeopleWithoutEcoliInRegion

Ecoli_originalprediction$numberOfPeopleLMICSEcoli <- sum(Ecoli_originalprediction$numberOfPeopleEcoliInRegion)# 3102063955

```

## E.coli estimates at UN region

```{r}
Ecoli_originalprediction_UNregion <- Ecoli_originalprediction %>% 
  group_by(UNRegionName) %>% 
  summarise(numberOfPeopleinUNRegion = sum(numberOfPeopleinRegion))
Ecoli_originalprediction_UNregion_ <- Ecoli_originalprediction %>% 
  group_by(UNRegionName) %>% 
  summarise(numberOfPeopleWithEcoliInUNRegion = sum(numberOfPeopleEcoliInRegion))

Ecoli_originalprediction_UNregion$proportionPopWithEcoli <- Ecoli_originalprediction_UNregion_$numberOfPeopleWithEcoliInUNRegion/Ecoli_originalprediction_UNregion$numberOfPeopleinUNRegion

Ecoli_originalprediction_UNSubregion <- Ecoli_originalprediction %>% 
  group_by(UNSubRegionName) %>% 
  summarise(numberOfPeopleWithEcoliInUNSubRegion = sum(numberOfPeopleEcoliInRegion))

library(scales)

UNSubregionoriginal_numberEcoli <- ggplot(Ecoli_originalprediction_UNSubregion, aes(x=reorder(UNSubRegionName, numberOfPeopleWithEcoliInUNSubRegion), y=numberOfPeopleWithEcoliInUNSubRegion)) + 
  geom_bar(stat = "identity",fill="light blue" ) + 
  scale_y_continuous(labels = label_number(suffix = " M", scale = 1e-6), limits = c(0, 1250000000))+
  labs(y = "Number of people exposed to E.coli in their primary drinking water source", x="") +
coord_flip() +
  theme_bw(base_size = 9)
UNSubregionoriginal_numberEcoli

```

## E.coli estimates at UNsubregion levels

```{r}
populationEcoli_original_UNSubregion <- Ecoli_originalprediction %>% 
  group_by(UNSubRegionName) %>% 
  summarise(numberOfPeopleinUNSubRegion = sum(numberOfPeopleinRegion))

Ecoli_originalprediction_UNSubregion$ProportionNoEcoli <- 1-Ecoli_originalprediction_UNSubregion$numberOfPeopleWithEcoliInUNSubRegion/populationEcoli_original_UNSubregion$numberOfPeopleinUNSubRegion

UNSubregion_original_ProportionNoEcoli <- ggplot(Ecoli_originalprediction_UNSubregion, aes(x=reorder(UNSubRegionName, 1-ProportionNoEcoli), y=1-ProportionNoEcoli, fill=UNSubRegionName)) + 
  geom_bar(stat = "identity",fill="light blue" ) + 
  labs(y = "Proportion of population exposed to E.coli in the primary drinking water source",x="") +
coord_flip() +
  theme_bw(base_size = 9)

UNSubregion_original_ProportionNoEcoli 

ggsave(
  "Fig_S9.pdf",
  dpi=300,
  width= 6, height = 4
)

nationalpopulation_originalPercentageNoEcoli <- Ecoli_originalprediction %>% 
  #filter(UNSubRegionName== "Sub-Saharan Africa") %>%
  group_by(NAME_0) %>% 
  summarise(numberOfPeopleEcoliIncountry = sum(numberOfPeopleEcoliInRegion))

nationalpopulation_originalEcoli <- Ecoli_originalprediction %>% 
  #filter(UNSubRegionName== "Sub-Saharan Africa") %>%
  group_by(NAME_0) %>% 
  summarise(numberOfPeopleincountry = sum(numberOfPeopleinRegion))

nationalpopulation_originalPercentageNoEcoli$percentageNoEcoli <- 1-nationalpopulation_originalPercentageNoEcoli$numberOfPeopleEcoliIncountry/nationalpopulation_originalEcoli$numberOfPeopleincountry

nationalpopulation_originalPercentageNoEcoli_ <- nationalpopulation_originalPercentageNoEcoli %>% 
left_join(Ecoli_originalprediction, by = c("NAME_0"="NAME_0"))%>%
  dplyr::select(UNRegionName, UNSubRegionName, NAME_0, numberOfPeopleEcoliIncountry, percentageNoEcoli) %>% unique()

nationalpopulation_originalPercentageNoEcoli_ $propotionEcoli <- 1-nationalpopulation_originalPercentageNoEcoli_ $percentageNoEcoli

```

# Modeling drinking water availability

Here we create a dataframe with indicators for drinking water
availability.

```{r}
df.MICS_HH_availability <-createDataFrameWithAvailabilty(df.MICS_HH)

df.MICS_HH_availability <- renameDuplicateHH7RegionNamesFromDifferentCountries(df.MICS_HH_availability)

Availability_training <-createIndicatorForRegionalAvailability(df.MICS_HH_availability)

```

Joining dataframe with earth observation data

```{r}
combined_Availability_envir_human_39 <-  Availability_training %>%
    left_join(EO_features, by = c("HH7_region"="HH7_region")) %>%
    select(all_of("HH7_region"),all_of("NAME_0"), all_of(bandNames_39), all_of("AvailableAtRegionalLevel"),all_of("HouseholdsInRegion.Freq.x")) 
```

## Grid-search: Tuning hyperparameters of random forest model with 39

features Outcome: access to drinking water availability

grid-search to iteratively explore the results of a suite of random
forest models trained on the selected covariates. A list of
hyperparameters was selected, varying in (1) the number of trees, (2)
the number of variables sampled at each split and (3) the minimum
observations per leaf and (4) sampling rate.

```{r}
h2o.shutdown()
h2o.init(nthreads = 7, max_mem_size = '8g', ignore_config = TRUE)

varToModel_Avail <-'AvailableAtRegionalLevel'

combined_Availability_envir_human_39 <- createColumnForCVfoldBasedOnCountries(combined_Availability_envir_human_39)

combined_Availability_envir_human_39 <-combined_Availability_envir_human_39%>% drop_na(country_fold)

# select only variables needed for random forest models  

combined_Avail_envir_regMatrix_39 <- combined_Availability_envir_human_39%>%
  select(all_of(bandNames_39),all_of(varToModel_Avail), HouseholdsInRegion.Freq.x, country_fold)%>% drop_na(country_fold)

#write.csv(combined_Avail_envir_regMatrix_39, here("5_Predictions/prediction_results/Regre_Matrix_Avail.csv"), fileEncoding = "UTF-8", row.names = F)

  
#save regression Matrix as H2O data frames  

regMatrixH2O_Avail_reg_39 <- as.h2o(combined_Avail_envir_regMatrix_39, destination_frame = "regMatrixH2O_Avail_reg_39")

rf.params_39_Avail<-list(ntrees = c(100,200),
                mtries = c(6,10,13,20), #(approx. n^(1/2), 1/4, 1/3, 1/2 of number of features)
                min_rows = c(1,3,5),
                sample_rate = c(0.6,0.8)
                )

search.criteria<-list(strategy = 'RandomDiscrete', max_models = 150, seed = 0, max_runtime_secs = 3600)

rf.grid_<-h2o.grid('randomForest',
                  y = varToModel_Avail,
                  grid_id = 'rf.grid_',
                  training_frame = regMatrixH2O_Avail_reg_39,
                  seed = 0,
                  hyper_params = rf.params_39_Avail,
                  fold_column = "country_fold",
                  weights_column = "HouseholdsInRegion.Freq.x",
                  fold_assignment = 'AUTO',
                  keep_cross_validation_predictions = TRUE, # Important!
                  keep_cross_validation_fold_assignment = TRUE, # Important!
                  search_criteria = search.criteria)

summary(rf.grid_, show_stack_traces = TRUE)
#retrieve grid searched model performance sort by RMSE/R2

rf.grid.perf<-h2o.getGrid(grid_id = 'rf.grid_',
                          sort_by = 'R2', # This is also coefficient of determinaion values

                                                    decreasing = TRUE)
summary(rf.grid_, show_stack_traces = TRUE)

print(rf.grid.perf@summary_table)

#write.csv(rf.grid.perf@summary_table,"./4_Training/results_grid_search/rf_grid_Avail_reg_39_newest.csv")

```

## best RF model with 39 features for water availability

```{r}
# random forest regression

rf_model_Avail_final <- h2o.randomForest(
  y = varToModel_Avail,
  training_frame = regMatrixH2O_Avail_reg_39,
  ntrees = 200,
  mtries = 10,
  sample_rate = 0.6,
  min_rows = 1,
  fold_column = "country_fold",
  weights_column = "HouseholdsInRegion.Freq.x",
  seed = 0,
  #fold_assignment = 'AUTO',
  keep_cross_validation_predictions = TRUE, # Important!
  keep_cross_validation_fold_assignment = TRUE)

h2o.r2(rf_model_Avail_final, xval = TRUE)
h2o.mae(rf_model_Avail_final,  xval = TRUE)

```

Plot Predicted versus Observed Availability with results from
leave-one-country-out holdout predictions.

```{r}
TrainAndPredicted_Avail <- as.data.frame(h2o.cross_validation_holdout_predictions(rf_model_Avail_final))

TrainAndPredicted_Avail$train <- combined_Availability_envir_human_39$AvailableAtRegionalLevel

TrainAndPredicted_Avail$nr_HH_in_region <- combined_Availability_envir_human_39$HouseholdsInRegion.Freq.x

TrainAndPredicted_Avail$Country <- combined_Availability_envir_human_39$NAME_0

country_Income_Groups <- readCountryIncomeGroup()

TrainAndPredicted_Avail<- joinWithCountryIncomeGroups(TrainAndPredicted_Avail,UN_Population_Devision_Names, country_Income_Groups)

TrainAndPredicted_Avail_withGlobalRegions <- TrainAndPredicted_Avail %>% left_join(GlobalRegions, by = c("ISO3.Alpha.code"="ISO.alpha3.Code"))

TrainAndPredicted_Avail_withGlobalRegions$UNRegionName[TrainAndPredicted_Avail_withGlobalRegions$Country == "Kosovo (under UNSC res. 1244)"] <- "Europe"

# Plot predicted vs observed values
                                    
plotPredObs_Avail <- ggplot(TrainAndPredicted_Avail_withGlobalRegions, 
                      aes(x = predict, 
                          y = train, size = nr_HH_in_region, colour= UNRegionName)) +
  geom_point(shape=1) +
  labs(x = "Predicted proportion of population with available drinking water",
       y = "Observed proportion of population with available drinking water") +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1)) +
stat_smooth(se = T,
              colour = "black",
              size = 0.5,
              method = "lm") +
  geom_abline(slope = 1,
              intercept = 0,
              na.rm = FALSE, 
              show.legend = NA, 
              linetype = "dashed") +
  labs(size="Households in region (n)", 
              col="UN region") +
  scale_colour_manual(values = colours) +
  theme_bw(base_size = 9) +
  guides(size = guide_legend(order = 1),
         colour = guide_legend(order = 2))

plotPredObs_Avail   

ggsave(
  "Fig_S8.pdf",
  dpi=300,
  width= 6, height = 4
)
```

Calculating and plotting SHAP values for Water Availability predictions

```{r}
# Create the SHAP summary plot
shap_summary_plot_avail <- h2o.shap_summary_plot(rf_model_Avail_final, regMatrixH2O_Avail_reg_39)
print(shap_summary_plot_avail)

shap_values_Avail <- predict_contributions.H2OModel(
  rf_model_Avail_final,
  regMatrixH2O_Avail_reg_39)

mean_values_Avail <- as.data.frame(colMeans(abs(as.data.frame(shap_values_Avail) %>% select(-BiasTerm)))); names(mean_values_Avail) <- 'Mean'
mean_values_Avail %>% arrange(desc(Mean))

#write.csv(mean_values_Avail, "./5_Predictions/prediction_results/Shap_values_Avail_newest.csv", fileEncoding = "UTF-8", row.names = T)

mean_values_Avail <- cbind(feature = rownames(mean_values_Avail), mean_values_Avail)

mean_values_Avail  <- rename39Features(mean_values_Avail)

mean_values_Avail <- group39FeaturesInto5IndicatorGroups(mean_values_Avail)


shap_summary_plot_data <- shap_summary_plot_avail[['data']]
shap_summary_plot_data$feature <- as.character(shap_summary_plot_data$feature)
shap_summary_plot_data <-rename39Features(shap_summary_plot_data)

shap_summary_plot_data <- shap_summary_plot_data %>%
  left_join(mean_values_Avail, by = c("feature"="feature"))


shap_summary_plot_data %>%
  mutate(feature = fct_reorder(feature, Mean)) %>%
ggplot(aes(feature, contribution, color = normalized_value, text = row)) +
        geom_hline(yintercept = 0, linetype = "dashed") +
        geom_point(position = h2o:::position_jitter_density(), alpha = 0.5, size = 1.2) +
        scale_fill_gradientn(colors = viridis_pal()(9), breaks=c(0,1),labels=c("low","high"), limits=c(0, 1)) +
        coord_flip() +
        labs(y = "SHAP contribution to availability when needed prediction", x = "", col = "Normalized\nfeature value") +
        theme_bw(base_size = 9) 

ggsave(
  "Fig_S14.pdf",
  dpi=300,
  width= 6, height = 4
)
```

Plotting summarized average SHAP values by Indicator Group for
Availability

```{r}
#colours <- c("#E64B35FF", "#91D1C2FF","#3C5488FF","#4DBBD5FF", "#00A087FF")

SHAP_values_Avail_grouped <- mean_values_Avail %>% 
  group_by(Indicator.Group) %>% 
  summarise(SumOfAverageSHAPValue = sum(Mean))

Plot_SHAP_Avail_grouped <- ggplot(SHAP_values_Avail_grouped, aes(x=reorder(Indicator.Group, SumOfAverageSHAPValue), y=SumOfAverageSHAPValue, fill= Indicator.Group)) +
  ggtitle("Availability when needed")+
  geom_bar(stat = "identity") + 
  labs(y = "sum of average SHAP values", x="") +
  scale_y_continuous(limits = c(0,0.150)) +
coord_flip() +
  scale_fill_manual(values=colours)+
  guides(fill = FALSE, color = FALSE, linetype = FALSE, shape = FALSE)+
  theme_bw()

Plot_SHAP_Avail_grouped
```

## Availability prediction at global administrative (GADM) area level 1

```{r}
#level 1 prediction

EO_GADM_level1 <- read.csv(here("1_Data/EnvironmentalFeatures/FeaturesAtGADMlevel1.csv"))

regressionMatrixForPrediction <-EO_GADM_level1 %>%  
 select(all_of(bandNames_39))
 
regMatrixH2O_4pred_39 <- as.h2o(regressionMatrixForPrediction, destination_frame = "regMatrixH2O_4pred_39")

FullPrediction_Avail <- as.data.frame(h2o.predict(rf_model_Avail_final, regMatrixH2O_4pred_39))

FullPrediction_Avail$region <- EO_GADM_level1[["NAME_1"]]
FullPrediction_Avail$GID_1 <- EO_GADM_level1[["GID_1"]]
FullPrediction_Avail$country <- EO_GADM_level1[["NAME_0"]]
FullPrediction_Avail$GID_0 <- EO_GADM_level1[["GID_0"]]

country_Income_Groups <- renameCountryCodeForKosovo(country_Income_Groups)

FullPrediction_Avail <- FullPrediction_Avail  %>% 
  left_join(country_Income_Groups, by=(c("GID_0"="Code")))

#Venezuela was not in the WB income group classification file so we add it as a lower-middle-income country according to the Inter-American Development Bank

FullPrediction_Avail$Income.group[FullPrediction_Avail$country == "Venezuela"] <- "Lower middle income"
FullPrediction_Avail$Income.group[FullPrediction_Avail$country == "Tokelau"] <- "Lower middle income"
FullPrediction_Avail$Income.group[FullPrediction_Avail$country == "Saint Helena"] <- "Upper middle income"

FullPrediction_Avail <- FullPrediction_Avail %>%
  na.omit(Income.group)

#write.csv(FullPrediction_Avail, "~/GitHub/mapping-safe-drinking-water-use-LMICs/Prediction/prediction_results/Fullprediction_Avail_newest.csv", fileEncoding = "UTF-8", row.names = F)
```

# Modeling drinking water accessibility

Here we create a dataframe with indicators for drinking water
accessibility.

```{r}
df.MICS_HH_accessibility <-createDataFrameWithAccessibilityIndicator(df.MICS_HH)

df.MICS_HH_accessibility <- renameDuplicateHH7RegionNamesFromDifferentCountries(df.MICS_HH_accessibility)

Accessibility_training <-createIndicatorForRegionalWaterAccessibility (df.MICS_HH_accessibility)

```

Joining dataframe with earth observation data

```{r}
combined_Accessibility_envir_human_39 <-  Accessibility_training %>%
    left_join(EO_features, by = c("HH7_region"="HH7_region")) %>%
    select(all_of("HH7_region"),all_of("NAME_0"), all_of(bandNames_39), all_of("AccessAtRegionalLevel"),all_of("HouseholdsInRegion.Freq.x")) 
```

## RF models for water accessibility- using grid search

```{r}
## modelling water accessibility
## grid-search to iteratively explore the results of a suite of random forest models trained on the selected covariates. A list of hyperparameters was selected, varying in (1) the number of trees, (2) the number of variables sampled at each split  and (3) the minimum observations per leaf and (4)sampling rate

h2o.shutdown()
h2o.init(nthreads = 7, max_mem_size = '8g', ignore_config = TRUE)

## modelling accessibility

varToModel_Access <-'AccessAtRegionalLevel'

combined_Accessibility_envir_human_39 <- createColumnForCVfoldBasedOnCountries(combined_Accessibility_envir_human_39)

combined_Accessibility_envir_human_39 <- combined_Accessibility_envir_human_39 %>% 
  drop_na(country_fold)

# select only variables needed for random forest models  
combined_Access_envir_regMatrix_39 <- combined_Accessibility_envir_human_39%>%
  select(all_of(bandNames_39),all_of(varToModel_Access),HouseholdsInRegion.Freq.x, country_fold)

#write.csv(combined_Access_envir_regMatrix_39 , "./5_Prediction/prediction_results/Regre_Matrix_Access.csv", fileEncoding = "UTF-8", row.names = F)

regMatrixH2O_Access_reg_39 <- as.h2o(combined_Access_envir_regMatrix_39, destination_frame = "regMatrixH2O_Access_reg_39")

rf.params_39_Access<-list(ntrees = c(100,200),
                mtries = c(6,10,13,20), #(about n^(1/2), 1/4, 1/3, 1/2 of number of features)
                min_rows = c(1,3,5),
                sample_rate = c(0.6,0.8)
                )

search.criteria<-list(strategy = 'RandomDiscrete', max_models = 150, seed = 0, max_runtime_secs = 3600)

rf.grid_39_Access<-h2o.grid('randomForest',
                  y = varToModel_Access,
                  grid_id = 'rf.grid_Access',
                  training_frame = regMatrixH2O_Access_reg_39,
                  seed = 0,
                  hyper_params = rf.params_39_Access,
                  fold_column = "country_fold",
                  weights_column = "HouseholdsInRegion.Freq.x",
                  fold_assignment = 'AUTO',
                  keep_cross_validation_predictions = TRUE, # Important!
                  keep_cross_validation_fold_assignment = TRUE, # Important!
                  search_criteria = search.criteria)

summary(rf.grid_39_Access, show_stack_traces = TRUE)
#retrieve grid searched model performance sort by RMSE/R2

rf_grid_perf_Access<-h2o.getGrid(grid_id = 'rf.grid_Access',
                          sort_by = 'R2', # This is also coefficient of determinaion values

                                                    decreasing = TRUE)
summary(rf.grid_39_Access, show_stack_traces = TRUE)

print(rf_grid_perf_Access@summary_table)

#write.csv(rf_grid_perf_Access@summary_table,"./4_Training/grid_search_results/rf_grid_Access_reg_39_newest.csv")

```

## best RF model with 39 features for water accessibility

```{r}
# random forest regression

rf_model_Access_final <- h2o.randomForest(
  y = varToModel_Access,
  training_frame = regMatrixH2O_Access_reg_39,
  ntrees = 200,
  mtries = 20,
  sample_rate = 0.8,
  min_rows = 3,
  #nfolds = 10,
  fold_column = "country_fold",
  weights_column = "HouseholdsInRegion.Freq.x",
  seed = 0,
  keep_cross_validation_predictions = TRUE, # Important!
  keep_cross_validation_fold_assignment = TRUE)

h2o.r2(rf_model_Access_final, xval = TRUE)
h2o.mae(rf_model_Access_final, xval = TRUE)
```

Plot Predicted versus Observed Accessibility with results from
leave-one-country-out holdout predictions.

```{r}
TrainAndPredicted_Access <- as.data.frame(h2o.cross_validation_holdout_predictions(rf_model_Access_final))

TrainAndPredicted_Access$train <- combined_Accessibility_envir_human_39$AccessAtRegionalLevel

TrainAndPredicted_Access$nr_HH_in_region <- combined_Accessibility_envir_human_39$HouseholdsInRegion.Freq.x

TrainAndPredicted_Access$Country <- combined_Accessibility_envir_human_39$NAME_0

country_Income_Groups <- readCountryIncomeGroup()

TrainAndPredicted_Access<- joinWithCountryIncomeGroups(TrainAndPredicted_Access,UN_Population_Devision_Names, country_Income_Groups)

TrainAndPredicted_Access_withGlobalRegions <- TrainAndPredicted_Access %>% left_join(GlobalRegions, by = c("ISO3.Alpha.code"="ISO.alpha3.Code"))

TrainAndPredicted_Access_withGlobalRegions$UNRegionName[TrainAndPredicted_Access_withGlobalRegions$Country == "Kosovo (under UNSC res. 1244)"] <- "Europe"


# Plot predicted vs observed values

plotPredObs_Access <- ggplot(TrainAndPredicted_Access_withGlobalRegions, 
                      aes(x = predict, 
                          y = train, size = nr_HH_in_region, colour= UNRegionName)) +
  geom_point(shape=1) +
  labs(x = "Predicted proportion of population with drinking water on premises",
       y = "Observed proportion of population with drinking water on premises") +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1)) +
stat_smooth(se = T,
              colour = "black",
              size = 0.5,
              method = "lm") +
  geom_abline(slope = 1,
              intercept = 0,
              na.rm = FALSE, 
              show.legend = NA, 
              linetype = "dashed") +
  theme_bw(base_size = 9)+
  labs(size="Households in region (n)", 
              col="UN region") +
  scale_colour_manual(values = colours) +
  guides(size = guide_legend(order = 1),
         colour = guide_legend(order = 2))

plotPredObs_Access  

ggsave(
  "Fig_S5.pdf",
  dpi=300,
  width= 6, height = 4
)
```

Calculating and plotting SHAP values for Water Accessibility predictions

```{r}
# Create the SHAP summary plot
shap_summary_plot_Access <- h2o.shap_summary_plot(rf_model_Access_final, regMatrixH2O_Access_reg_39)
print(shap_summary_plot_Access)

shap_values_Access <- predict_contributions.H2OModel(
  rf_model_Access_final,
  regMatrixH2O_Access_reg_39)

mean_values_Access <- as.data.frame(colMeans(abs(as.data.frame(shap_values_Access) %>% select(-BiasTerm)))); names(mean_values_Access) <- 'Mean'
mean_values_Access %>% arrange(desc(Mean))

mean_values_Access <- cbind(feature = rownames(mean_values_Access), mean_values_Access)

mean_values_Access  <- rename39Features(mean_values_Access)

mean_values_Access <- group39FeaturesInto5IndicatorGroups(mean_values_Access)

shap_summary_plot_data <- shap_summary_plot_Access[['data']]
shap_summary_plot_data$feature <- as.character(shap_summary_plot_data$feature)
shap_summary_plot_data <-rename39Features(shap_summary_plot_data)

shap_summary_plot_data <- shap_summary_plot_data %>%
  left_join(mean_values_Access, by = c("feature"="feature"))

shap_summary_plot_data %>%
  mutate(feature = fct_reorder(feature, Mean)) %>%
ggplot(aes(feature, contribution, color = normalized_value, text = row)) +
        geom_hline(yintercept = 0, linetype = "dashed") +
        geom_point(position = h2o:::position_jitter_density(), alpha = 0.5, size = 1.2) +
        scale_fill_gradientn(colors = viridis_pal()(9), breaks=c(0,1),labels=c("low","high"), limits=c(0, 1)) +
        coord_flip() +
        labs(y = "SHAP contribution to accessibility on premises prediction", x = "", col = "Normalized\nfeature value") +
        theme_bw(base_size = 9) 

ggsave(
  "Fig_S12.pdf",
  dpi=300,
  width= 6, height = 4
)

#write.csv(mean_values_Access, "./5_Prediction/prediction_results/Shap_values_Access_newest.csv", fileEncoding = "UTF-8", row.names = T)
```

Plotting summarized average SHAP values by Indicator Group for Access

```{r}
SHAP_values_Access_grouped <- mean_values_Access %>% 
  group_by(Indicator.Group) %>% 
  summarise(SumOfAverageSHAPValue = sum(Mean))

Plot_SHAP_Access_grouped <- ggplot(SHAP_values_Access_grouped, aes(x=reorder(Indicator.Group, SumOfAverageSHAPValue), y=SumOfAverageSHAPValue, fill= Indicator.Group)) +
  ggtitle("Accessibility when needed")+
  geom_bar(stat = "identity") + 
  labs(y = "sum of average SHAP values", x="") +
  scale_y_continuous(limits = c(0,0.150)) +
coord_flip() +
  scale_fill_manual(values=colours)+
  guides(fill = FALSE, color = FALSE, linetype = FALSE, shape = FALSE)+
  theme_bw()

Plot_SHAP_Access_grouped
```

## Accessibility prediction at global administrative (GADM) area level 1

```{r}
#level 1 prediction

EO_GADM_level1 <- read.csv(here("1_Data/EnvironmentalFeatures/FeaturesAtGADMlevel1.csv"))

regressionMatrixForPrediction <-EO_GADM_level1 %>%  
 select(all_of(bandNames_39))
 
regMatrixH2O_4pred_39 <- as.h2o(regressionMatrixForPrediction, destination_frame = "regMatrixH2O_4pred_39")

FullPrediction_Access <- as.data.frame(h2o.predict(rf_model_Access_final, regMatrixH2O_4pred_39))

FullPrediction_Access$region <- EO_GADM_level1[["NAME_1"]]
FullPrediction_Access$GID_1 <- EO_GADM_level1[["GID_1"]]
FullPrediction_Access$country <- EO_GADM_level1[["NAME_0"]]
FullPrediction_Access$GID_0 <- EO_GADM_level1[["GID_0"]]

country_Income_Groups <- renameCountryCodeForKosovo(country_Income_Groups)

FullPrediction_Access <- FullPrediction_Access  %>% 
  left_join(country_Income_Groups, by=(c("GID_0"="Code")))

#Venezuela was not in the WB income group classification file so we add it as a lower-middle-income country according to the Inter-American Development Bank

FullPrediction_Access$Income.group[FullPrediction_Access$country == "Venezuela"] <- "Lower middle income"
FullPrediction_Access$Income.group[FullPrediction_Access$country == "Tokelau"] <- "Lower middle income"
FullPrediction_Access$Income.group[FullPrediction_Access$country == "Saint Helena"] <- "Upper middle income"

FullPrediction_Access <- FullPrediction_Access %>%
  na.omit(Income.group)

#write.csv(FullPrediction_Access, "./5_Prediction/prediction_results/FullPrediction_Access_newest.csv", fileEncoding = "UTF-8", row.names = F)
```

## Descriptive analysis of accessibility results

```{r}
#source("here(./2_Data_cleaning/3_MICS_Preprocessing_Dataframes/1_Compiling_MICS_test_and_training_sets/functionsForJoiningAndStructuringDataframes.R"))

Access_prediction <- read.csv(here("./5_Predictions/prediction_results/Fullprediction_Access_.csv"))

GlobalRegions <- readAndRenameStandardGlobalRegionNames()

#joining and preparing dataframe for analysis
Access_prediction <- Access_prediction %>% 
  left_join(EO_GADM_level1, by = c("GID_1"="GID_1", "GID_0"="GID_0")) %>% 
  left_join(GlobalRegions, by = c("GID_0"="ISO.alpha3.Code")) 

Access_prediction$country[Access_prediction$country == "Kosovo"] <- "Kosovo (under UNSC res. 1244)"
  
Access_prediction$UNRegionName[Access_prediction$country == "Kosovo (under UNSC res. 1244)"] <- "Europe"

Access_prediction$UNSubRegionName[Access_prediction$country == "Kosovo (under UNSC res. 1244)"] <- "Eastern Europe"

Access_prediction <- Access_prediction  %>% 
  filter(Income.group == "Low income" |
         Income.group == "Lower middle income" |
         Income.group == "Upper middle income")
```

## Accessibility estimates across all LMICs

```{r}
Access_prediction$numberOfPeopleinRegion <- Access_prediction$GPWv4_Population_2020
Access_prediction$numberOfPeopleWithAccessInRegion <- Access_prediction$GPWv4_Population_2020*Access_prediction$predict

Access_prediction$totalNumberOfPeopleLMICS <- sum(Access_prediction$numberOfPeopleinRegion)#6.523e+09
Access_prediction$NumberOfPeopleLMICSwithAccess <- sum(Access_prediction$numberOfPeopleWithAccessInRegion)#4203853245

```

## Access estimates at regional level

```{r}
Access_prediction$numberOfPeopleWithoutAccessInRegion <- Access_prediction$numberOfPeopleinRegion-Access_prediction$numberOfPeopleWithAccessInRegion

Access_prediction$numberOfPeopleLMICSWithoutAccess <- sum(Access_prediction$numberOfPeopleWithoutAccessInRegion)# 2319147206

Access_prediction_UNregion <- Access_prediction %>% 
  group_by(UNRegionName) %>% 
  summarise(numberOfPeopleinUNRegion = sum(numberOfPeopleinRegion))
Access_prediction_UNregion_ <- Access_prediction %>% 
  group_by(UNRegionName) %>% 
  summarise(numberOfPeopleWithoutAccessInUNRegion = sum(numberOfPeopleWithoutAccessInRegion))

Access_prediction_UNregion$proportionPopWithoutAccess <- Access_prediction_UNregion_$numberOfPeopleWithoutAccessInUNRegion/Access_prediction_UNregion$numberOfPeopleinUNRegion

Access_prediction_UNSubregion <- Access_prediction %>% 
  group_by(UNSubRegionName) %>% 
  summarise(numberOfPeopleWithoutAccessInUNSubRegion = sum(numberOfPeopleWithoutAccessInRegion))

```

# Modeling water source type

Here we create a dataframe with indicators for drinking water source
type (proportion of people using an improved drinking water source).

```{r}
df.MICS_HH_improved <-createDataFrameWithmainWaterSourceType(df.MICS_HH)

df.MICS_HH_improved <- renameDuplicateHH7RegionNamesFromDifferentCountries(df.MICS_HH_improved)

Improved_training <-createIndicatorForRegionalWaterSourceType(df.MICS_HH_improved)

```

Joining dataframe with earth observation data

```{r}
combined_Improved_envir_human_39 <-  Improved_training %>%
    left_join(EO_features, by = c("HH7_region"="HH7_region")) %>%
    select(all_of("HH7_region"),all_of("NAME_0"), all_of(bandNames_39), all_of("ImprovedAtRegionalLevel"),all_of("HouseholdsInRegion.Freq.x")) 
```

## Grid Search for water source type model

grid-search to iteratively explore the results of a suite of random
forest models trained on the selected covariates. A list of
hyperparameters was selected, varying in (1) the number of trees, (2)
the number of variables sampled at each split and (3) the minimum
observations per leaf and (4)sampling rate.

```{r}
h2o.shutdown(prompt = FALSE)
h2o.init(nthreads = 7, max_mem_size = '8g', ignore_config = TRUE)

## modelling water source type
varToModel_improved <-'ImprovedAtRegionalLevel'

combined_Improved_envir_human_39 <- createColumnForCVfoldBasedOnCountries(combined_Improved_envir_human_39)

combined_Improved_envir_human_39 <- combined_Improved_envir_human_39 %>%
  drop_na(country_fold)
                             
# select only variables needed for random forest models  
combined_Improved_envir_regMatrix_39 <- combined_Improved_envir_human_39 %>%
  select(all_of(bandNames_39),all_of(varToModel_improved),HouseholdsInRegion.Freq.x, country_fold) 

regMatrixH2O_improved_reg_39 <- as.h2o(combined_Improved_envir_regMatrix_39, destination_frame = "regMatrixH2O_improved_reg_39")

#write.csv(combined_Improved_envir_regMatrix_39, here("5_Predictions/prediction_results/Regre_Matrix_Improved.csv"), fileEncoding = "UTF-8", row.names = F)

rf.params_39_improved<-list(ntrees = c(100,200),
                mtries = c(6,10,13,20), #(about n^(1/2), 1/4, 1/3, 1/2 of number of features)
                min_rows = c(1,3,5),
                sample_rate = c(0.6,0.8)
                )

search.criteria<-list(strategy = 'RandomDiscrete', max_models = 150, seed = 0, max_runtime_secs = 3600)


rf.grid_39_improved<-h2o.grid('randomForest',
                  y = varToModel_improved,
                  grid_id = 'rf.grid_improved',
                  training_frame = regMatrixH2O_improved_reg_39,
                  seed = 0,
                  hyper_params = rf.params_39_improved,
                  fold_column = "country_fold",
                  weights_column = "HouseholdsInRegion.Freq.x",
                  fold_assignment = 'AUTO',
                  keep_cross_validation_predictions = TRUE, # Important!
                  keep_cross_validation_fold_assignment = TRUE, # Important!
                  search_criteria = search.criteria)

summary(rf.grid_39_improved, show_stack_traces = TRUE)
#retrieve grid searched model performance sort by RMSE/R2

rf_grid_perf_improved<-h2o.getGrid(grid_id = 'rf.grid_improved',
                          sort_by = 'R2', # This is also coefficient of determination values

                                                    decreasing = TRUE)
summary(rf.grid_39_improved, show_stack_traces = TRUE)

print(rf_grid_perf_improved@summary_table)

#write.csv(rf_grid_perf_improved@summary_table,"./4_Training/results_grid_search/grid_search_results_rf_grid_improved_reg_39_newest.csv")

```

## Best RF model for predicting water source type

```{r}
rf_model_improved_final <- h2o.randomForest(
  y = varToModel_improved,
  training_frame = regMatrixH2O_improved_reg_39,
  ntrees = 200,
  mtries = 13,
  sample_rate = 0.6,
  min_rows = 1,
  fold_column = "country_fold",
  weights_column = "HouseholdsInRegion.Freq.x",
  seed = 0,
  keep_cross_validation_predictions = TRUE, # Important!
  keep_cross_validation_fold_assignment = TRUE)

h2o.r2(rf_model_improved_final, xval = TRUE)
h2o.mae(rf_model_improved_final, xval = TRUE)

```

Plot predicted versus Observed populations using improved drinking water
sources with results from leave-one-country-out holdout predictions.

```{r}
TrainAndPredicted_Improved <- as.data.frame(h2o.cross_validation_holdout_predictions(rf_model_improved_final))

TrainAndPredicted_Improved$train <- combined_Improved_envir_human_39$ImprovedAtRegionalLevel

TrainAndPredicted_Improved$nr_HH_in_region <- combined_Improved_envir_human_39$HouseholdsInRegion.Freq.x

TrainAndPredicted_Improved$Country <- combined_Improved_envir_human_39$NAME_0

country_Income_Groups <- readCountryIncomeGroup()

TrainAndPredicted_Improved<- joinWithCountryIncomeGroups(TrainAndPredicted_Improved,UN_Population_Devision_Names, country_Income_Groups)

TrainAndPredicted_Improved_withGlobalRegions <- TrainAndPredicted_Improved %>% left_join(GlobalRegions, by = c("ISO3.Alpha.code"="ISO.alpha3.Code"))

TrainAndPredicted_Improved_withGlobalRegions$UNRegionName[TrainAndPredicted_Improved_withGlobalRegions$Country == "Kosovo (under UNSC res. 1244)"] <- "Europe"


# Plot predicted vs observed values
                                    
plotPredObs_Improved <- ggplot(TrainAndPredicted_Improved_withGlobalRegions, 
                      aes(x = predict, 
                          y = train, size = nr_HH_in_region, colour= UNRegionName)) +
  geom_point(shape=1) +
  labs(x = "Predicted proportion of population using an improved source",
       y = "Observed proportion of population using an improved source") +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1)) +
stat_smooth(se = T,
              colour = "black",
              size = 0.5,
              method = "lm") +
  geom_abline(slope = 1,
              intercept = 0,
              na.rm = FALSE, 
              show.legend = NA, 
              linetype = "dashed") +
  labs(size="Households in region (n)", 
              col="UN region") +
  scale_colour_manual(values = colours) +
  theme_bw(base_size= 9) +
  guides(size = guide_legend(order = 1),
         colour = guide_legend(order = 2))

plotPredObs_Improved  

ggsave(
  "Fig_S7.pdf",
  dpi=300,
  width= 6, height = 4
)
```

Calculating and plotting SHAP values for improved source predictions

```{r}
#Calculating average SHAP values
shap_values_Improved <- predict_contributions.H2OModel(
  rf_model_improved_final,
  regMatrixH2O_improved_reg_39)

mean_values_Improved <- as.data.frame(colMeans(abs(as.data.frame(shap_values_Improved) %>% select(-BiasTerm)))); names(mean_values_Improved) <- 'Mean'
mean_values_Improved %>% arrange(desc(Mean))

mean_values_Improved <- cbind(feature = rownames(mean_values_Improved), mean_values_Improved)

mean_values_Improved  <- rename39Features(mean_values_Improved)

mean_values_Improved <- group39FeaturesInto5IndicatorGroups(mean_values_Improved)

#write.csv(mean_values_Improved, "./5_Prediction/predition_results/Shap_values_Improved_newest.csv", fileEncoding = "UTF-8", row.names = T)
```

Plotting summarized average SHAP values by Indicator Group for Water
Source Type

```{r}
SHAP_values_Improved_grouped <- mean_values_Improved %>% 
  group_by(Indicator.Group) %>% 
  summarise(SumOfAverageSHAPValue = sum(Mean))

Plot_SHAP_Improved_grouped <- ggplot(SHAP_values_Improved_grouped, aes(x=reorder(Indicator.Group, SumOfAverageSHAPValue), y=SumOfAverageSHAPValue, fill= Indicator.Group)) +
  ggtitle("Use of improved drinking water source")+
  geom_bar(stat = "identity") + 
  labs(y = "sum of average SHAP values", x="") +
  scale_y_continuous(limits = c(0,0.150)) +
coord_flip() +
  scale_fill_manual(values=colours)+
  guides(fill = FALSE, color = FALSE, linetype = FALSE, shape = FALSE)+
  theme_bw()

Plot_SHAP_Improved_grouped
```

Create the SHAP summary plot

```{r}
shap_summary_plot_Improved <- h2o.shap_summary_plot(rf_model_improved_final, regMatrixH2O_improved_reg_39)
print(shap_summary_plot_Improved)

shap_summary_plot_data <- shap_summary_plot_Improved[['data']]
shap_summary_plot_data$feature <- as.character(shap_summary_plot_data$feature)
shap_summary_plot_data <-rename39Features(shap_summary_plot_data)

shap_summary_plot_data <- shap_summary_plot_data %>%
  left_join(mean_values_Improved, by = c("feature"="feature"))

shap_summary_plot_data %>%
  mutate(feature = fct_reorder(feature, Mean)) %>%
ggplot(aes(feature, contribution, color = normalized_value, text = row)) +
        geom_hline(yintercept = 0, linetype = "dashed") +
        geom_point(position = h2o:::position_jitter_density(), alpha = 0.5, size = 1.2) +
        scale_fill_gradientn(colors = viridis_pal()(9), breaks=c(0,1),labels=c("low","high"), limits=c(0, 1)) +
        coord_flip() +
        labs(y = "SHAP contribution to improved source prediction", x = "", col = "Normalized\nfeature value") +
        theme_bw(base_size = 9) 

ggsave(
  "Fig_S11.pdf",
  dpi=300,
  width= 6, height = 4
)

```

## Improved drinking water source prediction at global administrative

(GADM) area level 1

```{r}
#level 1 prediction

EO_GADM_level1 <- read.csv(here("1_Data/EnvironmentalFeatures/FeaturesAtGADMlevel1.csv"))

regressionMatrixForPrediction <-EO_GADM_level1 %>%  
 select(all_of(bandNames_39))
 
regMatrixH2O_4pred_39 <- as.h2o(regressionMatrixForPrediction, destination_frame = "regMatrixH2O_4pred_39")

FullPrediction_Improved <- as.data.frame(h2o.predict(rf_model_improved_final, regMatrixH2O_4pred_39))

FullPrediction_Improved$region <- EO_GADM_level1[["NAME_1"]]
FullPrediction_Improved$GID_1 <- EO_GADM_level1[["GID_1"]]
FullPrediction_Improved$country <- EO_GADM_level1[["NAME_0"]]
FullPrediction_Improved$GID_0 <- EO_GADM_level1[["GID_0"]]

country_Income_Groups <- renameCountryCodeForKosovo(country_Income_Groups)

FullPrediction_Improved <- FullPrediction_Improved  %>% 
  left_join(country_Income_Groups, by=(c("GID_0"="Code")))

#Venezuela was not in the WB income group classification file so we add it as a lower-middle-income country according to the Inter-American Development Bank

FullPrediction_Improved$Income.group[FullPrediction_Improved$country == "Venezuela"] <- "Lower middle income"
FullPrediction_Improved$Income.group[FullPrediction_Improved$country == "Tokelau"] <- "Lower middle income"
FullPrediction_Improved$Income.group[FullPrediction_Improved$country == "Saint Helena"] <- "Upper middle income"

FullPrediction_Improved <- FullPrediction_Improved %>%
  na.omit(Income.group)

#write.csv(FullPrediction_Improved, "./5_Prediction/prediction_results/FullPrediction_Improved_newest.csv", fileEncoding = "UTF-8", row.names = F)
```

## Descriptive analysis of improved drinking water source results

```{r}
source(here("./2_Data_cleaning/3_MICS_Preprocessing_Dataframes/1_Compiling_MICS_test_and_training_sets/functionsForJoiningAndStructuringDataframes.R"))

Improved_prediction <- read_csv("~/Documents/GitHub/mapping-safe-drinking-water-use-LMICs/5_Predictions/prediction_results/Fullprediction_improved.csv")

GlobalRegions <- readAndRenameStandardGlobalRegionNames()

#joining and preparing dataframe for analysis
Improved_prediction <- Improved_prediction %>% 
  left_join(EO_GADM_level1, by = c("GID_1"="GID_1", "GID_0"="GID_0")) %>% 
  left_join(GlobalRegions, by = c("GID_0"="ISO.alpha3.Code")) 

Improved_prediction$country[Improved_prediction$country == "Kosovo"] <- "Kosovo (under UNSC res. 1244)"
  
Improved_prediction$UNRegionName[Improved_prediction$country == "Kosovo (under UNSC res. 1244)"] <- "Europe"

Improved_prediction$UNSubRegionName[Improved_prediction$country == "Kosovo (under UNSC res. 1244)"] <- "Eastern Europe"

Improved_prediction <- Improved_prediction  %>% 
  filter(Income.group == "Low income" |
         Income.group == "Lower middle income" |
         Income.group == "Upper middle income")
```

## Water source type estimates of across all LMICs

```{r}
Improved_prediction$numberOfPeopleinRegion <- Improved_prediction$GPWv4_Population_2020
Improved_prediction$numberOfPeopleWithImprovedInRegion <- Improved_prediction$GPWv4_Population_2020*Improved_prediction$predict

Improved_prediction$totalNumberOfPeopleLMICS <- sum(Improved_prediction$numberOfPeopleinRegion)#6.523e+09
Improved_prediction$NumberOfPeopleLMICSwithImproved <- sum(Improved_prediction$numberOfPeopleWithImprovedInRegion)#5743373538

Improved_prediction$numberOfPeopleWithOutImprovedInRegion <- Improved_prediction$numberOfPeopleinRegion-Improved_prediction$numberOfPeopleWithImprovedInRegion

Improved_prediction$numberOfPeopleLMICSWithImproved <- sum(Improved_prediction$numberOfPeopleWithOutImprovedInRegion)#779626913

```
